---
title: "Measuring Who Connects with Whom: Homophily & Dyadic Analysis"
description: "Use homophily, mixing matrices, and dyadic attributes to study who connects with whom."
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measuring Who Connects with Whom: Homophily & Dyadic Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  dev = 'png', dpi = 150,
  cache = FALSE,
  echo = TRUE,
  message = FALSE
)

```

## vignette summary

This vignette shows how to use **netify** to explore relationships between international alliance patterns and country characteristics using data from the Correlates of War (COW) project and Alliance Treaty Obligations and Provisions (ATOP) data. The examples focus on four common international relations questions:

1. **Democratic Peace**: Do democracies cooperate more with each other than with non-democracies?
2. **Economic Interdependence**: Do countries with similar economic development levels cooperate more?
3. **Geographic Proximity**: Does geographic distance affect cooperation patterns?
4. **Regional Clustering**: Do countries primarily cooperate within their own regions?

We'll focus on how to do some exploratory statistical analysis with **netify**:

1. `homophily()`: Do Birds of a Feather Flock Together?

   - Tests whether similar countries tend to form alliances with each other. For example, do democracies primarily ally with other democracies? Do rich countries mainly partner with other rich countries?
   - Calculates correlations between attribute similarity and network tie presence using multiple similarity metrics. The optional permutation test compares the observed dyad-level association to a permuted similarity vector.
   - Function notes: 
     - **Flexible similarity metrics**: Correlation, euclidean, categorical, cosine and other methods
     - **Permutation summaries**: P-values and confidence intervals for exploratory comparisons
     - **Missing data handling**: Uses dyads with the needed edge and attribute values
     - **Multi-network ready**: Works across time periods and layers

2. `mixing_matrix()`: Who Actually Partners With Whom?

   - Creates detailed "who allies with whom" tables. Shows not just whether democracies ally with democracies, but how much they interact with autocracies, hybrid regimes, and other categories.
   - Constructs mixing matrices showing tie distributions across attribute combinations. Calculates assortativity coefficients, modularity scores, and entropy measures to quantify mixing patterns with optional normalization schemes.
   - Function notes:
     - **Cross-dimensional analysis**: How regime types mix across regions
     - **Rich summary statistics**: Assortativity, modularity, entropy, diagonal proportions  
     - **Flexible normalization**: Raw counts, proportions, or row-normalized
     - **Weighted network support**: Incorporates alliance strength, not just presence

3. `dyad_correlation()`: Which Relationship Factors Are Associated with Alliances?

   - Tests how characteristics of country pairs (like geographic distance, trade volume, or cultural similarity) are associated with alliance ties. Answers questions like "Do nearby countries ally more often?"
   - Correlates dyadic (pairwise) variables with network ties using multiple correlation methods. Supports partial correlation analysis to control for confounding dyadic factors while handling missing data through pairwise deletion.
   - Function notes:
     - **Partial correlation support**: Isolate specific effects while controlling for others
     - **Multiple correlation methods**: Pearson, Spearman, Kendall with significance testing
     - **Binary network options**: Analyze tie presence vs. strength separately
     - **Attribute diagnostics**: Descriptive stats for all variables

4. `attribute_report()`: Attribute Report Summary

   - Runs the main attribute-oriented analyses from one interface and returns results on actor position, mixing patterns, and dyadic associations.
   - Runs homophily analysis, mixing matrices, dyadic correlations, and centrality-attribute relationships from one call. Tries to automatically determine appropriate methods based on variable types.

```{r}
library(netify)
library(ggplot2)
library(peacesciencer)
library(dplyr)
library(countrycode)

fmt_p <- function(p, digits = 3) {
  cutoff <- 10^-digits
  ifelse(
    is.na(p),
    NA_character_,
    ifelse(
      p < cutoff,
      paste0("< ", formatC(cutoff, format = "f", digits = digits)),
      formatC(p, format = "f", digits = digits)
    )
  )
}
```

## data preparation

We'll use the Correlates of War data via the `peacesciencer` package to build a network of international alliances. This data includes measures of democracy, economic development, military capabilities, geographic relationships between countries, and alliance commitments from the ATOP dataset.

### cow data

```{r}
# download peacesciencer external data if needed
peacesciencer::download_extdata()

# build dyadic dataset for a recent 5-year period
cow_dyads <- create_dyadyears(subset_years = c(2010:2014)) |>
  add_cow_mids() |>
  add_capital_distance() |>
  add_democracy() |>
  add_sim_gdp_pop(keep = "pwtrgdp") |>
  add_nmc() |>
  add_atop_alliance()

# build alliance cooperation measure from atop alliance types
cow_dyads <- cow_dyads |>
  mutate(
    alliance_score = atop_defense + atop_offense + atop_neutral + atop_nonagg + atop_consul,
    alliance_norm = alliance_score / 5,
    cooperation = alliance_norm,
    region1 = countrycode(ccode1, "cown", "region"),
    region2 = countrycode(ccode2, "cown", "region"),
    log_gdp1 = log(pwtrgdp1 + 1),
    log_gdp2 = log(pwtrgdp2 + 1),
    log_capdist = log(capdist + 1),
    alliance_intensity = alliance_norm,
    defense_alliance = atop_defense
  )

# filter to 2012 for cross-sectional analysis
cow_2012 <- cow_dyads |>
  filter(year == 2012)
```


```{r}
# create alliance network
alliance_net <- netify(
  cow_2012,
  actor1 = 'ccode1', actor2 = 'ccode2',
  symmetric = TRUE,
  weight = 'cooperation'
)

alliance_net
```


**Nodal data** is one row per actor (here, per country) describing actor-level attributes — distinct from dyadic (pair-level) variables.

```{r}
# prepare nodal data with country attributes
nodal_data <- cow_2012 |>
  select(
    ccode1, region1, v2x_polyarchy1,
    log_gdp1, cinc1
    ) |>
  distinct() |>
  rename(
    actor = ccode1,
    region = region1,
    democracy = v2x_polyarchy1,
    log_gdp = log_gdp1,
    mil_capability = cinc1
  ) |>
  mutate(
    regime_type = case_when(
      democracy >= 0.6 ~ "Democracy",
      democracy >= 0.4 ~ "Hybrid",
      democracy < 0.4 ~ "Autocracy",
      TRUE ~ "Unknown"
    ),
    development = case_when(
      log_gdp >= quantile(log_gdp, 0.75, na.rm = TRUE) ~ "High",
      log_gdp >= quantile(log_gdp, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    )
  )

nodal_data$country_name <- countrycode(nodal_data$actor, "cown", "country.name")

alliance_net <- add_node_vars(alliance_net, nodal_data, actor = "actor")
```

Add dyadic (relationship-level) variables. A **dyad** is a pair of actors and a **dyadic variable** describes the relationship itself (e.g., distance between two capitals) rather than either actor alone.

```{r}
# prepare dyadic data
dyad_data <- cow_2012 |>
  select(ccode1, ccode2, log_capdist, alliance_norm, atop_defense) |>
  rename(
    actor1 = ccode1,
    actor2 = ccode2,
    geographic_distance = log_capdist,
    alliance_intensity = alliance_norm,
    defense_alliance = atop_defense
  )

alliance_net <- add_dyad_vars(
  alliance_net,
  dyad_data = dyad_data,
  actor1 = "actor1",
  actor2 = "actor2",
  dyad_vars = c("geographic_distance", "alliance_intensity", "defense_alliance"),
  dyad_vars_symmetric = c(TRUE, TRUE, TRUE)
)
```

## 1. testing the democratic peace with `homophily()`

Democratic peace arguments often lead researchers to ask whether democracies are more likely to cooperate with one another. Here we use alliance data to examine a narrower descriptive question: are allied country pairs more similar in democracy scores than non-allied pairs?

### 🔍 using `homophily()` for continuous variables

**Homophily** is the tendency for connected actors to be similar to each other on some attribute -- "birds of a feather flock together." The `homophily()` function is a tool in **netify** that tests whether similar actors tend to connect more in a network. It can handle both continuous and categorical attributes.

The function's optional **permutation test** permutes the dyad-level similarity values relative to tie indicators. The p-value is useful as an exploratory comparison, not as evidence that the network came from independent dyads or that similarity caused ties.

```{r}
# test whether countries with similar democracy levels form more alliances
democracy_homophily <- homophily(
  alliance_net,
  attribute = "democracy",
  method = "correlation",
  significance_test = TRUE
)

knitr::kable(democracy_homophily, digits=3, align='c')

democracy_summary <- paste0(
  "**Democracy Homophily Results:**\n\n",
  "- Homophily correlation: ", round(democracy_homophily$homophily_correlation, 3), "\n",
  "- Avg similarity among allies: ", round(democracy_homophily$mean_similarity_connected, 3), "\n",
  "- Avg similarity among non-allies: ", round(democracy_homophily$mean_similarity_unconnected, 3), "\n",
  "- P-value: ", fmt_p(democracy_homophily$p_value), "\n",
  if(democracy_homophily$p_value < 0.05) {
    "→ Alliance ties are more common among similarly democratic country pairs\n"
  } else {
    "→ No detectable democracy-similarity association in these alliance ties\n"
  }
)
```

```{r, echo=FALSE, results='asis'}
cat(democracy_summary)
```


#### understanding the output:

- **homophily_correlation**: Measures tendency for similar values to connect (0 to 1)
- **mean_similarity_connected**: Average similarity among connected pairs
- **mean_similarity_unconnected**: Average similarity among unconnected pairs
- **p_value**: Statistical significance of the homophily pattern

The democracy homophily analysis returns a small but statistically detectable association: countries with more similar democracy scores are slightly more likely to be alliance partners. The homophily correlation of about 0.14 (p < 0.05) is a modest positive association, not a causal estimate of why alliances form. The negative similarity values (roughly -0.25 for allies vs -0.32 for non-allies) reflect the correlation method's transformed distance metric, where higher (less negative) values indicate greater similarity. The ~0.07 difference between allied and non-allied pairs means that allied countries are, on average, modestly closer in regime score than non-allied ones. This descriptive comparison does not adjust for geography, power, timing, or other network dependence.

#### visualizing homophily patterns

We can visualize the homophily pattern to better understand how democracy similarity relates to alliance formation:

```{r, fig.width=8, fig.height=5}
plot_homophily(democracy_homophily, alliance_net,
               type = "distribution",
               attribute = "democracy",
               method = "correlation",
               sample_size = 5000) +
  labs(subtitle = "Allied countries show greater similarity in democracy scores") +
  xlim(c(-1, 1))
```

#### understanding the distribution shape

The distribution plot reveals the empirical density of pairwise similarity scores computed using the correlation method from `calculate_similarity_matrix()`. The distinctly non-normal shape arises from the specific calculation procedure:

**Details of the Similarity Calculation:**

For continuous attributes like democracy scores, when `method = "correlation"` is specified, the homophily function computes pairwise similarities as a negative absolute difference:

```r
# for each dyad (i,j), similarity is calculated as:
similarity[i,j] = -abs(attr[i] - attr[j])
```

This produces similarity scores that:

- Equal `0` when two actors have the same attribute value
- Become more negative as the attribute difference grows
- Generate the observed multimodal distribution due to the discrete clustering of democracy scores

#### interpretation of the result

The observed homophily correlation of about 0.14 indicates that, despite these distributional complexities, allied countries have higher democracy similarity scores than non-allied pairs in this snapshot. The mean difference (roughly -0.25 vs -0.32) is detectable in the dyad-level permutation summary even though both distributions have similar non-normal shapes.

However, the extensive overlap between distributions shows that democracy similarity alone does not explain alliance formation. Many democratic countries ally with non-democracies (left side of the blue distribution), while many similar democracies remain unallied (right side of the gold distribution). Geographic, security, economic, and temporal factors would need to be modeled separately before making stronger claims.

### using `homophily()` for categorical variables

Now let's move onto the categorical regime type variable we made: 

```{r}
head(attr(alliance_net, 'nodal_data'))
```

```{r}
regime_homophily <- homophily(
  alliance_net,
  attribute = "regime_type",
  method = "categorical",
  significance_test = TRUE)

knitr::kable(regime_homophily, digits=3, align='c')

regime_summary <- paste0(
  "**Regime Type Homophily Results:**\n\n",
  "- Homophily score: ", round(regime_homophily$homophily_correlation, 3), "\n",
  "- Same-regime alliances: ", round(regime_homophily$mean_similarity_connected * 100, 1), "%\n",
  "- Different-regime alliances: ", round((1 - regime_homophily$mean_similarity_connected) * 100, 1), "%\n",
  "- Expected if random: ", round(regime_homophily$mean_similarity_unconnected * 100, 1), "%\n",
  "- P-value: ", fmt_p(regime_homophily$p_value), "\n",
  if(regime_homophily$p_value < 0.05 && regime_homophily$homophily_correlation > 0.15) {
    "→ Same-regime alliance ties are clearly more common in this network\n"
  } else if(regime_homophily$p_value < 0.05 && regime_homophily$homophily_correlation > 0) {
    "→ Same-regime alliance ties are modestly more common in this network\n"
  } else {
    "→ No detectable regime-type association in these alliance ties\n"
  }
)
```

```{r, echo=FALSE, results='asis'}
cat(regime_summary)
```

The regime type analysis shows a small but detectable pattern of political homophily in alliance formation. With a homophily score of about 0.12 (p < 0.05), alliance ties are modestly more common within the same regime type. The similarity scores show that roughly 39% of allied pairs share the same regime type, compared to only 26% of non-allied pairs -- a roughly 13 percentage point difference. Treat this as a descriptive association unless a design explicitly accounts for alternative explanations.

The categorical nature of this analysis provides a clearer interpretation than continuous measures: when countries form alliances, there is about a 39% chance their partner shares the same regime type, compared to 26% for non-allied pairs. The modest association leaves plenty of room for cross-regime alliances tied to strategic, geographic, and economic considerations. Note that we are not testing the democratic peace idea specifically here, since we are amalgamating autocracy-autocracy and democracy-democracy pairs into the same-regime bucket.

#### visualizing categorical homophily

For categorical variables like regime type, the visualization looks different. Instead of continuous similarity distributions, we see discrete categories:

```{r, fig.width=8, fig.height=5}
plot_homophily(regime_homophily, alliance_net,
               type = "distribution",
               attribute = "regime_type",
               method = "categorical",
               sample_size = 5000) +
  labs(title = "Regime Type Homophily in Alliance Networks",
       subtitle = "Do similar political systems form more alliances?")
```

Unlike the continuous democracy score, regime type similarity is binary: country pairs either share the same regime type (similarity = 1) or they don't (similarity = 0). The visualization shows two bars comparing the proportion of alliances within each category. A higher blue bar at similarity = 1 indicates that countries with the same regime type are more likely to form alliances than those with different regime types. This categorical approach provides a clearer test of the "democracies ally with democracies" hypothesis, though it loses the nuance of how similar countries are on the democracy spectrum.



## 2. economic interdependence and development

International relations theory often asks whether countries at similar levels of economic development are more likely to cooperate. We can summarize that relationship with `homophily()`:

```{r}
gdp_homophily <- homophily(
  alliance_net,
  attribute = "log_gdp",
  method = "correlation",
  significance_test = TRUE)

knitr::kable(gdp_homophily, digits=3, align='c')

# gate substantive claims on magnitude (|r| >= 0.15) so a near-zero
# correlation does not get described as substantively large just because n is large
economic_summary <- paste0(
  "**Economic Development Homophily Results:**\n\n",
  "- Homophily correlation: ", round(gdp_homophily$homophily_correlation, 3), "\n",
  "- Similarity among allies: ", round(gdp_homophily$mean_similarity_connected, 3), "\n",
  "- Similarity among non-allies: ", round(gdp_homophily$mean_similarity_unconnected, 3), "\n",
  "- P-value: ", fmt_p(gdp_homophily$p_value), "\n",
	dplyr::case_when(
	  gdp_homophily$p_value >= 0.05 ~
	    "→ No detectable association between economic development similarity and alliance patterns\n",
	  gdp_homophily$homophily_correlation >= 0.15 ~
	    "→ Alliance ties are more common among countries at similar development levels\n",
	  gdp_homophily$homophily_correlation > 0 ~
	    "→ Detectable but very small (|r| < 0.15); the dyad count is doing most of the work behind the p-value\n",
	  TRUE ~
	    "→ Slight association with alliances between countries at different development levels (small heterophily)\n"
	)
)
```

```{r, echo=FALSE, results='asis'}
cat(economic_summary)
```

## 3. regional clustering in international cooperation

Do countries primarily form alliances within their own regions, or are alliances more globally distributed? Regional patterns provide another example of categorical homophily:

```{r}
region_homophily <- homophily(
  alliance_net,
  attribute = "region",
  method = "categorical",
  significance_test = TRUE)

knitr::kable(region_homophily, digits=3, align='c')

regional_summary <- paste0(
  "**Regional Clustering Results:**\n\n",
  "- Homophily score: ", round(region_homophily$homophily_correlation, 3), "\n",
  "- Within-region alliances: ", round(region_homophily$mean_similarity_connected, 3), "\n",
  "- Cross-region alliances: ", round(region_homophily$mean_similarity_unconnected, 3), "\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(regional_summary)
```

## 4. who forms alliances with whom? using `mixing_matrix()`

The `mixing_matrix()` function reveals detailed interaction patterns between different types of actors in your network. This is crucial for understanding not just *if* certain types connect, but *how much* and *with whom*.

A few summary statistics will show up below: **assortativity** is a -1 to 1 coefficient that captures how strongly ties prefer same-attribute partners (positive = same-type, 0 = random, negative = opposites-attract). **Modularity** rewards within-group edges over what you'd expect by chance, so higher values mean ties cluster more strongly within attribute groups. **Entropy** quantifies how spread out (versus concentrated) the mixing pattern is across the cells. **Diagonal proportion** is the share of ties that fall on the diagonal of the mixing matrix (i.e., within the same category).

### 📊 democracy mixing matrix

```{r}
regime_mixing <- mixing_matrix(
  alliance_net,
  attribute = "regime_type",
  normalized = TRUE
)

knitr::kable(round(regime_mixing$mixing_matrices[[1]], 3),
             caption = "Regime Type Alliance Matrix (normalized)",
             align = "c")

regime_mixing_summary <- paste0(
  "**Key Insights from mixing_matrix():**\n\n",
  "- Assortativity: ", round(regime_mixing$summary_stats$assortativity, 3), "\n",
  "  (Positive = similar types connect more; Negative = different types connect more)\n",
  "- Proportion of within-type alliances: ", round(regime_mixing$summary_stats$diagonal_proportion, 3), "\n",
  "  (Higher values indicate more homophily)\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(regime_mixing_summary)
```

#### how to read the mixing matrix:

- **Rows**: Source regime type
- **Columns**: Target regime type  
- **Values**: Proportion of ties from row type to column type
- **Diagonal**: Within-type alliances (homophily)

### 🌍 regional alliance patterns with row normalization

```{r}
region_mixing <- mixing_matrix(
  alliance_net,
  attribute = "region",
  normalized = TRUE,
  by_row = TRUE)

regional_mixing_header <- "**Regional Alliance Matrix (row-normalized):**\n\n"
```

```{r, echo=FALSE, results='asis'}
cat(regional_mixing_header)
```

```{r}
knitr::kable(round(region_mixing$mixing_matrices[[1]], 3),
             caption = "Regional Alliance Matrix (row-normalized)",
             align = "c")
```

#### visualizing regional alliance patterns

```{r, fig.width=9, fig.height=7}
plot_mixing_matrix(
    region_mixing,
    show_values = TRUE,
    value_digits = 2,
    text_size = 3,
    text_color_threshold = .7,
    diagonal_emphasis = TRUE,
    reorder_categories = FALSE
) +
  labs(title = "Regional Alliance Patterns",
       subtitle = "Within-region vs cross-region alliance formation",
       x = "Allied with region",
       y = "From region") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10))
```

The regional mixing matrix shows strong regional clustering in alliance formation. The emphasized diagonal shows that most regions primarily form alliances within their own geographic area, with some notable cross-regional partnerships.

### 🔀 cross-dimensional analysis: region × regime type

`mixing_matrix()` can analyze interactions across different attributes:

```{r}
cross_mixing <- mixing_matrix(
  alliance_net,
  attribute = "regime_type",
  row_attribute = "region",
  normalized = TRUE)

cross_mixing_header <- "**How different regime types form alliances across regions:**\n\n"
```

```{r, echo=FALSE, results='asis'}
cat(cross_mixing_header)
```

```{r}
knitr::kable(round(cross_mixing$mixing_matrices[[1]], 3),
             caption = "Cross-dimensional Analysis: Regime Types Across Regions",
             align = "c")
```

## 5. analyzing relationship-level factors with `dyad_correlation()`

A **dyad** is just a pair of actors (here, a pair of countries), and a **dyadic variable** describes the relationship itself (e.g., distance between two capitals, shared language) rather than either actor alone. The `dyad_correlation()` function examines how relationship-level (dyadic) variables correlate with network ties. This is useful for identifying pair-level factors associated with connections.

### 🌍 geographic distance and alliance formation

```{r}
geo_correlation <- dyad_correlation(
  alliance_net,
  dyad_vars = "geographic_distance",
  method = "pearson",
  significance_test = TRUE
)

geo_summary <- paste0(
  "**Geographic Distance and Alliance Formation (dyad_correlation results):**\n\n",
  "- Correlation coefficient: ", round(geo_correlation$correlation, 3), "\n",
  "- P-value: ", fmt_p(geo_correlation$p_value), "\n",
  "- Number of dyads analyzed: ", geo_correlation$n_pairs[1], "\n\n",
  if(geo_correlation$correlation < -0.1 && geo_correlation$p_value < 0.05) {
    "→ Alliance ties are more common among geographically closer country pairs.\n  (Negative correlation = shorter distance, more alliances)\n"
  } else if(geo_correlation$correlation > 0.1 && geo_correlation$p_value < 0.05) {
    "→ Greater distance is associated with more alliance ties in this dyad-level summary.\n  (Positive correlation = greater distance, more alliances)\n"
  } else {
    "→ No clear geographic-distance association appears in these alliance ties.\n  (No detectable correlation)\n"
  }
)
```

The reported p-value comes from the ordinary correlation test on dyad vectors. Use it as a descriptive screen; network dependence and omitted dyadic structure require a more explicit modeling strategy for stronger inference.

```{r, echo=FALSE, results='asis'}
cat(geo_summary)
```

### 🤝 analyzing multiple dyadic variables

```{r}
multi_dyad_correlation <- dyad_correlation(
  alliance_net,
  dyad_vars = c("geographic_distance", "alliance_intensity", "defense_alliance"),
  method = "pearson",
  significance_test = TRUE
)

multi_dyad_summary <- paste0(
  "**Multiple Dyadic Variables Analysis:**\n\n",
  paste(sapply(1:nrow(multi_dyad_correlation), function(i) {
    paste0(
      "**", multi_dyad_correlation$dyad_var[i], ":**\n",
      "  - Correlation: ", round(multi_dyad_correlation$correlation[i], 3), "\n",
      "  - P-value: ", fmt_p(multi_dyad_correlation$p_value[i]), "\n"
    )
  }), collapse = "\n")
)
```

```{r, echo=FALSE, results='asis'}
cat(multi_dyad_summary)
```

## 6. attribute reports with `attribute_report()`

**Centrality** measures how important / well-positioned each actor is in the network -- e.g., `degree` counts a node's ties, `betweenness` counts how often a node lies on the shortest path between two others, and `closeness` measures how short the average distance is from a node to everyone else. **Heterophily** is the opposite of homophily: ties are more common between unlike partners.

The `attribute_report()` function combines the previous analyses into one report.

### running the report

```{r}
attribute_results <- attribute_report(
  alliance_net,
  node_vars = c("region", "regime_type", "democracy", "log_gdp", "mil_capability"),
  dyad_vars = c("geographic_distance", "alliance_intensity", "defense_alliance"),
  include_centrality = TRUE,
  include_homophily = TRUE,
  include_mixing = TRUE,
  include_dyadic_correlations = TRUE,
  centrality_measures = c("degree", "betweenness", "closeness"),
  significance_test = TRUE
)
```

`attribute_report` returns a list with multiple components:

* **`homophily_analysis`**: Tests for each node attribute
* **`mixing_matrices`**: Interaction patterns for categorical variables
* **`centrality_correlations`**: How attributes relate to network position
* **`dyadic_correlations`**: How dyad attributes are associated with ties

### extracting results

```{r}
homophily_header <- paste0(
  "**=== HOMOPHILY ANALYSIS ===**\n\n",
  "Do similar countries form more alliances?\n\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(homophily_header)
homophily_summary <- attribute_results$homophily_analysis

# cohen-style magnitude bins for correlation: ~.1 small, ~.3 medium, ~.5 large
homophily_display <- homophily_summary |>
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "n.s."
    ),
    interpretation = case_when(
      homophily_correlation > 0.5 & p_value < 0.05 ~ "Strong homophily",
      homophily_correlation > 0.3 & p_value < 0.05 ~ "Moderate homophily",
      homophily_correlation > 0.15 & p_value < 0.05 ~ "Weak homophily",
      homophily_correlation > 0 & p_value < 0.05 ~ "Very weak (sig. but small)",
      p_value >= 0.05 ~ "No significant pattern",
      TRUE ~ "Heterophily"
    )
  )

knitr::kable(
  homophily_display[, 
  c(
    "attribute", "method", 
    "homophily_correlation", 
    "p_value", "significance", 
    "interpretation")],
  caption = "Homophily Analysis Results",
  col.names = c("Attribute", "Method", "Homophily Correlation", "P-value", "Significance", "Interpretation"),
  digits = 3,
  align = "c"
)
```

```{r}
power_header <- paste0(
  "**=== POWER AND INFLUENCE ===**\n\n",
  "What makes countries central in the alliance network?\n\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(power_header)
centrality_summary <- attribute_results$centrality_correlations

centrality_display <- centrality_summary |>
  mutate(
    interpretation = case_when(
      correlation > 0.3 & p_value < 0.05 ~ "Strongly associated with centrality",
      correlation > 0.1 & p_value < 0.05 ~ "Moderately associated with centrality",
      correlation < -0.1 & p_value < 0.05 ~ "Negatively associated with centrality",
      p_value < 0.05 ~ "Weakly associated with centrality",
      TRUE ~ "Not significantly related to centrality"
    )
  ) |>
  arrange(p_value)

knitr::kable(
  centrality_display[1:10, 
  c(
    "node_var", "centrality_measure", 
    "correlation", "p_value", 
    "interpretation")],
  caption = "Top 10 Centrality-Attribute Correlations",
  col.names = c("Node Variable", "Centrality Measure", "Correlation", "P-value", "Interpretation"),
  digits = 3,
  align = "c"
)
```

```{r}
relationship_header <- paste0(
  "**=== RELATIONSHIP FACTORS ===**\n\n",
  "What dyadic factors are associated with alliance formation?\n\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(relationship_header)
dyadic_summary <- attribute_results$dyadic_correlations
knitr::kable(dyadic_summary[, c("dyad_var", "correlation", "p_value")],
             caption = "Dyadic Variables Analysis",
             col.names = c("Dyadic Variable", "Correlation", "P-value"),
             digits = 3,
             align = "c")
```

## 7. testing specific ir hypotheses

### hypothesis 1: democratic peace

```{r}
# binary democracy indicator
nodal_data_binary <- nodal_data |>
  mutate(is_democracy = ifelse(regime_type == "Democracy", 1, 0))

alliance_net_binary <- add_node_vars(
  alliance_net,
  nodal_data_binary[, c("actor", "is_democracy")],
  actor = "actor")

dem_peace_test <- homophily(
  alliance_net_binary,
  attribute = "is_democracy",
  method = "categorical",
  significance_test = TRUE)

# flag both significance and magnitude so a tiny r does not get described
# as a clean "democracies prefer democracies" pattern just because p<0.05
dem_peace_summary <- paste0(
  "**Democratic Peace Hypothesis Test:**\n\n",
  "- Effect size: ", round(dem_peace_test$homophily_correlation, 3), "\n",
  "- P-value: ", fmt_p(dem_peace_test$p_value), "\n",
	"- Conclusion: ", dplyr::case_when(
	  dem_peace_test$p_value >= 0.05 ~
	    "No detectable democracy-based alliance pattern",
	  dem_peace_test$homophily_correlation >= 0.15 ~
	    "Democracy-democracy alliance ties are clearly more common",
	  dem_peace_test$homophily_correlation > 0 ~
	    "Detectable but very small same-regime pattern; the dyad count explains the significance flag more than the association size",
	  TRUE ~
	    "Detectable heterophily: democracy/non-democracy ties are more common"
	), "\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(dem_peace_summary)
```

### hypothesis 2: power politics

Do powerful countries (high military capability) primarily form alliances with other powerful countries?

```{r}
power_homophily <- homophily(
  alliance_net,
  attribute = "mil_capability",
  method = "correlation",
  significance_test = TRUE)

# gate substantive claims on |r| so a near-zero correlation does not get
# labeled "heterophily" simply because the dyad count makes every p tiny
power_politics_summary <- paste0(
  "**Power Politics Hypothesis:**\n\n",
  "- Correlation: ", round(power_homophily$homophily_correlation, 3), "\n",
  "- P-value: ", fmt_p(power_homophily$p_value), "\n",
	"- Interpretation: ", dplyr::case_when(
	  power_homophily$p_value >= 0.05 ~
	    "No detectable power-based alliance pattern",
	  power_homophily$homophily_correlation >= 0.15 ~
	    "Alliance ties are more common among similarly powerful countries",
	  power_homophily$homophily_correlation <= -0.15 ~
	    "Alliance ties are more common between high- and low-power countries (heterophily)",
	  TRUE ~
	    "Detectable but very small (|r| < 0.15); read as mild power-mixing rather than a clear homophily/heterophily pattern"
	), "\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(power_politics_summary)
```

## 8. visualizing network patterns

And as seen in other vignettes we can use the `plot()` function to visualize the network with node attributes and edge weights:

### network visualization by attributes

```{r}
# attach degree and other actor stats so we can map them to node aesthetics
alliance_net <- add_node_vars(
  alliance_net,
  summary_actor(alliance_net),
  actor = "actor"
)
```

```{r, fig.width=8, fig.height=6}
plot(alliance_net,
     node_color_by = "region",
     node_color_label = "",
     node_shape_by = "regime_type",
     node_shape_label = "",
     node_size_by = "degree",
     node_size_label = "Degree",
     node_fill = "white",
     edge_color = "grey50",
     edge_linewidth = 0.5,
     edge_alpha_label = 'Alliance Strength (scaled)',
     layout = "nicely",
     seed = 6886) +
  ggtitle("ATOP Network") +
  theme(legend.position = 'right')
```

### visualizing homophily results

```{r, fig.width=10, fig.height=6}
plot_homophily(attribute_results$homophily_analysis,
               type = "comparison") +
  labs(title = "Alliance Formation Patterns: Which Attributes Matter?",
       subtitle = "Homophily analysis summarizes similarity patterns in international cooperation")
```

### visualizing centrality patterns

```{r, fig.width=10, fig.height=8}
centrality_viz <- attribute_results$centrality_correlations |>
  filter(p_value < 0.1) |>
  mutate(
    significant = p_value < 0.05,
    node_var = factor(node_var),
    centrality_measure = factor(
      centrality_measure,
      levels = c("degree", "betweenness", "closeness"))
  )

if(nrow(centrality_viz) > 0) {
  ggplot(centrality_viz, aes(x = correlation, y = node_var, color = significant)) +
    geom_segment(
      aes(
        x = 0, xend = correlation, y = node_var, yend = node_var),
      size = 1) +
    geom_point(size = 3) +
    facet_wrap(~centrality_measure, ncol = 1) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    scale_color_manual(
        values = c("FALSE" = "gray60", "TRUE" = "#2E86AB"),
        labels = c("FALSE" = "Not significant", "TRUE" = "(p < 0.05)")
    ) +
    labs(title = "What Makes Countries Central in the Alliance Network?",
         subtitle = "Correlation between node attributes and centrality measures",
         x = "Correlation with Centrality",
         y = "Node Attribute",
         color = "") +
    theme_bw() +
    theme(
        panel.border = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "top",
        panel.grid.major.y = element_blank()
    )
} else {
  no_centrality_msg <- "**No significant centrality correlations to visualize.**\n"
  cat(no_centrality_msg)
}
```

### mixing matrix heatmap

We can use the `plot_mixing_matrix()` function to visualize the mixing patterns as a heatmap:

```{r, fig.width=8, fig.height=6}
plot_mixing_matrix(
    regime_mixing,
    show_values = TRUE,
    diagonal_emphasis = TRUE,
    text_color_threshold=.9
  ) +
  labs(title = "Regime Type Alliance Patterns",
       subtitle = "How different regime types interact in the network",
       x = "Allied with...",
       y = "Regime type")
```

The heatmap shows the alliance patterns between different regime types. The diagonal cells (emphasized with black borders) represent within-type alliances, while off-diagonal cells show cross-type alliances. Darker blue indicates higher proportions of alliances, consistent with the small but detectable regime-type homophily in alliance formation.

## 9. working with longitudinal networks

A **longitudinal network** is a network observed over multiple time periods (one snapshot per period), in contrast to a **cross-sectional network** that captures a single point in time. The attribute analysis functions in netify work with longitudinal networks. The example below creates a longitudinal alliance network and runs the same analyses across multiple time periods.

### creating a longitudinal network

```{r}
# longitudinal alliance network for 2010-2014
alliance_net_longit <- netify(
  cow_dyads,
  actor1 = 'ccode1',
  actor2 = 'ccode2',
  time = 'year',
  symmetric = TRUE,
  weight = 'cooperation'
)

alliance_net_longit
```

### adding attributes to longitudinal networks

```{r}
# nodal data for all time periods
nodal_data_longit <- cow_dyads |>
  select(year, ccode1, region1, v2x_polyarchy1, log_gdp1, cinc1) |>
  distinct() |>
  rename(
    time = year,
    actor = ccode1,
    region = region1,
    democracy = v2x_polyarchy1,
    log_gdp = log_gdp1,
    mil_capability = cinc1
  ) |>
  mutate(
    regime_type = case_when(
      democracy >= 0.6 ~ "Democracy",
      democracy >= 0.4 ~ "Hybrid",
      democracy < 0.4 ~ "Autocracy",
      TRUE ~ "Unknown"
    )
  )

alliance_net_longit <- add_node_vars(
  alliance_net_longit,
  nodal_data_longit,
  actor = "actor",
  time = "time"
)

dyad_data_longit <- cow_dyads |>
  select(year, ccode1, ccode2, log_capdist, alliance_intensity, defense_alliance) |>
  rename(
    time = year,
    actor1 = ccode1,
    actor2 = ccode2,
    geographic_distance = log_capdist
  )

alliance_net_longit <- add_dyad_vars(
  alliance_net_longit,
  dyad_data = dyad_data_longit,
  actor1 = "actor1",
  actor2 = "actor2",
  time = "time",
  dyad_vars = c("geographic_distance", "alliance_intensity", "defense_alliance"),
  dyad_vars_symmetric = c(TRUE, TRUE, TRUE)
)
```

### homophily analysis across time

```{r}
democracy_homophily_longit <- homophily(
  alliance_net_longit,
  attribute = "democracy",
  method = "correlation",
  significance_test = TRUE
)

democracy_homophily_longit

# summary of trends across time periods
homophily_trends <- democracy_homophily_longit |>
  group_by(net) |>
  summarise(
    avg_homophily = mean(homophily_correlation, na.rm = TRUE),
    significant = any(p_value < 0.05, na.rm = TRUE)
  )

knitr::kable(homophily_trends, 
             caption = "Democracy Homophily Trends Over Time",
             digits = 3)
```

### visualizing longitudinal homophily

```{r, fig.width=10, fig.height=6}
plot_homophily(democracy_homophily_longit, type = "temporal") +
  labs(title = "Democracy Homophily in Alliance Networks Over Time",
       subtitle = "Tendency for democracies to ally with other democracies")
```

If you want to see the distribution for a specific time period, you can extract that period first:

```{r, fig.width=8, fig.height=5}
# extract 2012 slice for a single-period distribution plot
alliance_2012 <- subset(alliance_net_longit, time = '2012')
democracy_homo_2012 <- homophily(
  alliance_2012,
  attribute = "democracy",
  method = "correlation"
)

plot_homophily(democracy_homo_2012, alliance_2012,
               type = "distribution",
               attribute = "democracy",
               method = "correlation") +
  labs(subtitle = "Distribution of democracy similarity scores in 2012 alliance network")
```

### mixing matrices over time

```{r}
regime_mixing_longit <- mixing_matrix(
  alliance_net_longit,
  attribute = "regime_type",
  normalized = TRUE
)

# summary statistics across each time period
mixing_summary <- regime_mixing_longit$summary_stats |>
  select(net, assortativity, diagonal_proportion) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

knitr::kable(mixing_summary,
             caption = "Regime Type Mixing Patterns Over Time",
             col.names = c("Year", "Assortativity", "Within-Type %"))
```

### dyadic correlations across time

```{r}
geo_correlation_longit <- dyad_correlation(
  alliance_net_longit,
  dyad_vars = "geographic_distance",
  method = "pearson",
  significance_test = TRUE
)

geo_summary_longit <- geo_correlation_longit |>
  select(net, correlation, p_value, n_pairs) |>
	mutate(
	  significant = ifelse(p_value < 0.05, "*", ""),
	  correlation = round(correlation, 3),
	  p_value = fmt_p(p_value)
	)

knitr::kable(geo_summary_longit,
             caption = "Geographic Distance and Alliance Formation Over Time",
             col.names = c("Year", "Correlation", "P-value", "N Dyads", "Sig."))
```

### longitudinal attribute report

```{r}
longit_attribute_results <- attribute_report(
  alliance_net_longit,
  node_vars = c("region", "regime_type", "democracy", "log_gdp"),
  dyad_vars = c("geographic_distance", "alliance_intensity"),
  include_centrality = TRUE,
  include_homophily = TRUE,
  include_mixing = TRUE,
  include_dyadic_correlations = TRUE,
  centrality_measures = c("degree", "betweenness"),
  significance_test = TRUE
)

if (!is.null(longit_attribute_results$homophily_analysis)) {
  longit_patterns <- longit_attribute_results$homophily_analysis |>
    filter(attribute == "democracy") |>
    select(net, homophily_correlation, p_value) |>
    mutate(
      trend = case_when(
        net == min(net) ~ "Start",
        net == max(net) ~ "End",
        TRUE ~ "Middle"
      )
    )

  longit_patterns
} else {
  knitr::asis_output(
    "Note: homophily for longitudinal networks is currently limited. For full coverage, analyze each time period separately.\n"
  )
}
```

## tl;dr

1. **From `homophily()`**: 
   - Whether democracies truly form more alliances with each other
   - Whether economic similarity is associated with alliance patterns
   - The strength of regional clustering in alliance formation

2. **From `mixing_matrix()`**:
   - Detailed patterns of who forms alliances with whom
   - Whether alliances cross regime type boundaries
   - How different regions form alliances globally

3. **From `dyad_correlation()`**:
   - The role of geographic distance in shaping alliance formation
   - How alliance types (defense, offense, etc.) cluster
   - Which relationship factors matter most for alliances

4. **From `attribute_report()`**:
   - What attributes make countries central/influential
   - Complete homophily patterns across all variables
   - One view of the main network-attribute relationships
   
## references

1. Leeds, B. A., Ritter, J. M., Mitchell, S. M., & Long, A. G. (2002). Alliance Treaty Obligations and Provisions, 1815–1944. International Interactions, 28(3), 237–260. DOI 10.1080/03050620213653. 

2. Miller, S. V. (2022). peacesciencer: An R Package for Quantitative Peace Science Research. Conflict Management and Peace Science, 39(6), 755–779. DOI 10.1177/07388942221077926. 
