---
title: "Measuring Who Connects with Whom: Homophily & Dyadic Analysis"
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measuring Who Connects with Whom: Homophily & Dyadic Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  dev = 'png', dpi = 150,
  cache = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

# This vignette requires peacesciencer, countrycode, and dplyr.
# Exit gracefully with an actionable message if any are missing.
pkg_check <- c("peacesciencer", "countrycode", "dplyr")
missing_pkgs <- pkg_check[!vapply(pkg_check, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
    message(
        "This vignette requires: ", paste(missing_pkgs, collapse = ", "), "\n",
        "Install with: install.packages(c(",
        paste0("'", missing_pkgs, "'", collapse = ", "), "))"
    )
    knitr::knit_exit()
}
```

## Vignette Summary

This vignette demonstrates how to play with **netify** to explore relationships between international alliance patterns and country characteristics using data from the Correlates of War (COW) project and Alliance Treaty Obligations and Provisions (ATOP) data. We'll toy around with some simple international relations ideas:

1. **Democratic Peace**: Do democracies cooperate more with each other than with non-democracies?
2. **Economic Interdependence**: Do countries with similar economic development levels cooperate more?
3. **Geographic Proximity**: Does geographic distance affect cooperation patterns?
4. **Regional Clustering**: Do countries primarily cooperate within their own regions?

We'll focus on how to do some exploratory statistical analysis with **netify**:

1. `homophily()`: Do Birds of a Feather Flock Together?

   - Tests whether similar countries tend to form alliances with each other. For example, do democracies primarily ally with other democracies? Do rich countries mainly partner with other rich countries?
   - Calculates correlations between attribute similarity and network tie presence using multiple similarity metrics. Performs permutation-based significance testing to determine if observed homophily patterns exceed random chance.
   - Function notes: 
     - **Flexible similarity metrics**: Correlation, euclidean, categorical, cosine and other methods
     - **Statistical rigor**: Permutation tests with confidence intervals
     - **Smart missing data handling**: Preserves maximum sample sizes
     - **Multi-network ready**: Works across time periods and layers

2. `mixing_matrix()`: Who Actually Partners With Whom?

   - Creates detailed "who allies with whom" tables. Shows not just that democracies prefer democracies, but exactly how much they interact with autocracies, hybrid regimes, etc. Think of it as a cross-tabulation on steroids.
   - Constructs mixing matrices showing tie distributions across attribute combinations. Calculates assortativity coefficients, modularity scores, and entropy measures to quantify mixing patterns with optional normalization schemes.
   - Function notes:
     - **Cross-dimensional analysis**: How regime types mix across regions (unique feature!)
     - **Rich summary statistics**: Assortativity, modularity, entropy, diagonal proportions  
     - **Flexible normalization**: Raw counts, proportions, or row-normalized
     - **Weighted network support**: Incorporates alliance strength, not just presence

3. `dyad_correlation()`: What Relationship Factors Drive Alliances?

   - Tests how characteristics of country pairs (like geographic distance, trade volume, or cultural similarity) predict whether they'll form alliances. Answers questions like "Do nearby countries ally more often?"
   - Correlates dyadic (pairwise) variables with network ties using multiple correlation methods. Supports partial correlation analysis to control for confounding dyadic factors while handling missing data through pairwise deletion.
   - Function notes:
     - **Partial correlation support**: Isolate specific effects while controlling for others
     - **Multiple correlation methods**: Pearson, Spearman, Kendall with significance testing
     - **Binary network options**: Analyze tie presence vs. strength separately
     - **Comprehensive diagnostics**: Descriptive stats for all variables

4. `attribute_report()`: The Complete Picture in One Function

   - A kind of Swiss Army knife for understanding how country characteristics relate to alliance patterns. Automates running relevant analyses and tells you what makes countries influential, who allies with whom, and what drives partnership formation.
   - Comprehensive wrapper combining homophily analysis, mixing matrices, dyadic correlations, and centrality-attribute relationships. Tries to automatically determine appropriate methods based on variable types.

```{r}
library(netify)
library(ggplot2)
library(peacesciencer)
library(dplyr)
library(countrycode)
```

## Data Preparation

We'll use the Correlates of War data via the `peacesciencer` package to build a network of international alliances. This data includes measures of democracy, economic development, military capabilities, geographic relationships between countries, and alliance commitments from the ATOP dataset.

### COW data

```{r}
# download peacesciencer external data if needed
peacesciencer::download_extdata()

# build dyadic dataset for a recent 5-year period
cow_dyads <- create_dyadyears(subset_years = c(2010:2014)) |>
  add_cow_mids() |>
  add_capital_distance() |>
  add_democracy() |>
  add_sdp_gdp() |>
  add_nmc() |>
  add_atop_alliance()

# build alliance cooperation measure from ATOP alliance types
cow_dyads <- cow_dyads |>
  mutate(
    alliance_score = atop_defense + atop_offense + atop_neutral + atop_nonagg + atop_consul,
    alliance_norm = alliance_score / 5,
    cooperation = alliance_norm,
    region1 = countrycode(ccode1, "cown", "region"),
    region2 = countrycode(ccode2, "cown", "region"),
    log_gdp1 = log(wbgdp2011est1 + 1),
    log_gdp2 = log(wbgdp2011est2 + 1),
    log_capdist = log(capdist + 1),
    alliance_intensity = alliance_norm,
    defense_alliance = atop_defense
  )

# filter to 2012 for cross-sectional analysis
cow_2012 <- cow_dyads |>
  filter(year == 2012)
```


```{r}
# create alliance network
alliance_net <- netify(
  cow_2012,
  actor1 = 'ccode1', actor2 = 'ccode2',
  symmetric = TRUE,
  weight = 'cooperation'
)

alliance_net
```


**Nodal data** is one row per actor (here, per country) describing actor-level attributes — distinct from dyadic (pair-level) variables.

```{r}
# prepare nodal data with country attributes
nodal_data <- cow_2012 |>
  select(
    ccode1, region1, v2x_polyarchy1,
    log_gdp1, cinc1
    ) |>
  distinct() |>
  rename(
    actor = ccode1,
    region = region1,
    democracy = v2x_polyarchy1,
    log_gdp = log_gdp1,
    mil_capability = cinc1
  ) |>
  mutate(
    regime_type = case_when(
      democracy >= 0.6 ~ "Democracy",
      democracy >= 0.4 ~ "Hybrid",
      democracy < 0.4 ~ "Autocracy",
      TRUE ~ "Unknown"
    ),
    development = case_when(
      log_gdp >= quantile(log_gdp, 0.75, na.rm = TRUE) ~ "High",
      log_gdp >= quantile(log_gdp, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    )
  )

nodal_data$country_name <- countrycode(nodal_data$actor, "cown", "country.name")

alliance_net <- add_node_vars(alliance_net, nodal_data, actor = "actor")
```

Add dyadic (relationship-level) variables. A **dyad** is a pair of actors and a **dyadic variable** describes the relationship itself (e.g., distance between two capitals) rather than either actor alone.

```{r}
# prepare dyadic data
dyad_data <- cow_2012 |>
  select(ccode1, ccode2, log_capdist, alliance_norm, atop_defense) |>
  rename(
    actor1 = ccode1,
    actor2 = ccode2,
    geographic_distance = log_capdist,
    alliance_intensity = alliance_norm,
    defense_alliance = atop_defense
  )

alliance_net <- add_dyad_vars(
  alliance_net,
  dyad_data = dyad_data,
  actor1 = "actor1",
  actor2 = "actor2",
  dyad_vars = c("geographic_distance", "alliance_intensity", "defense_alliance"),
  dyad_vars_symmetric = c(TRUE, TRUE, TRUE)
)
```

## 1. Testing the Democratic Peace with `homophily()`

The democratic peace theory posits that democracies rarely engage in conflict with one another, driven by shared liberal norms, institutional constraints on executive power, and transparency in political decision-making. Here we examine whether these same mechanisms that reduce conflict might also promote cooperation, specifically, whether democratic states demonstrate a preference for forming alliances with other democracies.

### 🔍 Using `homophily()` for Continuous Variables

**Homophily** is the tendency for connected actors to be similar to each other on some attribute -- "birds of a feather flock together." The `homophily()` function is a tool in **netify** that tests whether similar actors tend to connect more in a network. It can handle both continuous and categorical attributes.

The function uses a **permutation test** for significance: actor attributes are repeatedly reshuffled across nodes to build a null distribution of homophily scores, and the observed score is compared against that null to produce a p-value.

```{r}
# test whether countries with similar democracy levels form more alliances
democracy_homophily <- homophily(
  alliance_net,
  attribute = "democracy",
  method = "correlation",
  significance_test = TRUE
)

knitr::kable(democracy_homophily, digits=3, align='c')

democracy_summary <- paste0(
  "**Democracy Homophily Results:**\n\n",
  "- Homophily correlation: ", round(democracy_homophily$homophily_correlation, 3), "\n",
  "- Avg similarity among allies: ", round(democracy_homophily$mean_similarity_connected, 3), "\n",
  "- Avg similarity among non-allies: ", round(democracy_homophily$mean_similarity_unconnected, 3), "\n",
  "- P-value: ", round(democracy_homophily$p_value, 3), "\n",
  if(democracy_homophily$p_value < 0.05) {
    "→ Democracies significantly tend to ally with similarly democratic countries\n"
  } else {
    "→ No significant democracy-based alliance preferences detected\n"
  }
)
```

```{r, echo=FALSE, results='asis'}
cat(democracy_summary)
```


#### Understanding the Output:

- **homophily_correlation**: Measures tendency for similar values to connect (0 to 1)
- **mean_similarity_connected**: Average similarity among connected pairs
- **mean_similarity_unconnected**: Average similarity among unconnected pairs
- **p_value**: Statistical significance of the homophily pattern

The democracy homophily analysis returns a small but statistically detectable association: countries with more similar democracy scores are slightly more likely to be alliance partners. The homophily correlation of about 0.14 (p < 0.05) is a modest positive association — consistent with a democratic-peace mechanism for alliance selection, but small in magnitude. The negative similarity values (roughly -0.25 for allies vs -0.32 for non-allies) reflect the correlation method's transformed distance metric, where higher (less negative) values indicate greater similarity. The ~0.07 difference between allied and non-allied pairs means that allied countries are, on average, modestly closer in regime score than non-allied ones. Across the roughly 3,570 connected pairs (out of ~19,000 total dyads), shared political institutions appear to nudge alliance partner selection — but only nudge, leaving plenty of room for strategic, geographic, and economic considerations.

#### Visualizing Homophily Patterns

We can visualize the homophily pattern to better understand how democracy similarity relates to alliance formation:

```{r, fig.width=8, fig.height=5}
plot_homophily(democracy_homophily, alliance_net,
               type = "distribution",
               attribute = "democracy",
               method = "correlation",
               sample_size = 5000) +
  labs(subtitle = "Allied countries show greater similarity in democracy scores") +
  xlim(c(-1, 1))
```

#### Understanding the Distribution Shape

The distribution plot reveals the empirical density of pairwise similarity scores computed using the correlation method from `calculate_similarity_matrix()`. The distinctly non-normal shape arises from the specific calculation procedure:

**Details of the Similarity Calculation:**

For continuous attributes like democracy scores, when `method = "correlation"` is specified, the homophily function computes pairwise similarities as:

```r
# For each dyad (i,j), similarity is calculated as:
similarity[i,j] = cor(attr[i], attr[j])
```

However, since we're dealing with scalar attribute values (single democracy score per country) rather than vectors, the function actually computes a **transformed distance metric** that preserves the correlation interpretation. Specifically, it uses:

```r
# Standardize the attribute
z_attr = (attr - mean(attr)) / sd(attr)

# Compute pairwise "correlation-like" similarity
similarity[i,j] = 1 - abs(z_attr[i] - z_attr[j]) / max_possible_distance
```

This produces similarity scores that:

- Range from -1 to 1 (like correlations)
- Capture relative similarity in standardized attribute space
- Generate the observed multimodal distribution due to the discrete clustering of democracy scores

#### Interpretation of the Result

The observed homophily correlation of about 0.14 indicates that despite these distributional complexities, allied countries do exhibit systematically higher democracy similarity scores than non-allied pairs. The mean difference (roughly -0.25 vs -0.32) is statistically significant even though both distributions exhibit similar non-normal shapes.

However, the extensive overlap between distributions reveals that democracy similarity is just one factor among many driving alliance formation. Many democratic countries ally with non-democracies (left side of blue distribution), while many similar democracies remain unallied (right side of gold distribution). This pattern reflects the reality of international politics: shared democratic values facilitate cooperation, but geographic, security, and economic considerations often prove equally or more influential in shaping alliance networks.

### 🔍 Using `homophily()` for Categorical Variables

Now let's move onto the categorical regime type variable we made: 

```{r}
head(attr(alliance_net, 'nodal_data'))
```

```{r}
regime_homophily <- homophily(
  alliance_net,
  attribute = "regime_type",
  method = "categorical",
  significance_test = TRUE)

knitr::kable(regime_homophily, digits=3, align='c')

regime_summary <- paste0(
  "**Regime Type Homophily Results:**\n\n",
  "- Homophily score: ", round(regime_homophily$homophily_correlation, 3), "\n",
  "- Same-regime alliances: ", round(regime_homophily$mean_similarity_connected * 100, 1), "%\n",
  "- Different-regime alliances: ", round((1 - regime_homophily$mean_similarity_connected) * 100, 1), "%\n",
  "- Expected if random: ", round(regime_homophily$mean_similarity_unconnected * 100, 1), "%\n",
  "- P-value: ", round(regime_homophily$p_value, 3), "\n",
  if(regime_homophily$p_value < 0.05 && regime_homophily$homophily_correlation > 0.15) {
    "→ Countries show a clear preference for allies with similar political systems\n"
  } else if(regime_homophily$p_value < 0.05 && regime_homophily$homophily_correlation > 0) {
    "→ Countries show a modest preference for allies with similar political systems\n"
  } else {
    "→ Regime type doesn't significantly influence alliance formation\n"
  }
)
```

```{r, echo=FALSE, results='asis'}
cat(regime_summary)
```

The regime type analysis reveals a small but detectable pattern of political homophily in alliance formation. With a homophily score of about 0.12 (p < 0.05), countries show a modest preference for forming alliances with similar regime types. The similarity scores show that roughly 39% of allied pairs share the same regime type, compared to only 26% of non-allied pairs -- a roughly 13 percentage point difference. This is consistent with political regime compatibility playing some role in alliance selection, but the effect size is small.

The categorical nature of this analysis provides a clearer interpretation than continuous measures: when countries form alliances, there's about a 39% chance their partner shares the same regime type, compared to 26% for non-allied pairs. Shared political institutions and governance norms may facilitate international cooperation, but the modest effect size leaves plenty of room for cross-regime alliances driven by strategic, geographic, and economic considerations. Note that we are not testing the democratic peace idea specifically here, since we are amalgamating autocracy-autocracy and democracy-democracy pairs into the same-regime bucket.

#### Visualizing Categorical Homophily

For categorical variables like regime type, the visualization looks different. Instead of continuous similarity distributions, we see discrete categories:

```{r, fig.width=8, fig.height=5}
plot_homophily(regime_homophily, alliance_net,
               type = "distribution",
               attribute = "regime_type",
               method = "categorical",
               sample_size = 5000) +
  labs(title = "Regime Type Homophily in Alliance Networks",
       subtitle = "Do similar political systems form more alliances?")
```

Unlike the continuous democracy score, regime type similarity is binary: country pairs either share the same regime type (similarity = 1) or they don't (similarity = 0). The visualization shows two bars comparing the proportion of alliances within each category. A higher blue bar at similarity = 1 indicates that countries with the same regime type are more likely to form alliances than those with different regime types. This categorical approach provides a clearer test of the "democracies ally with democracies" hypothesis, though it loses the nuance of how similar countries are on the democracy spectrum.



## 2. Economic Interdependence and Development

International relations theory suggests that countries with similar levels of economic development tend to form more alliances. Let's test this hypothesis:

```{r}
gdp_homophily <- homophily(
  alliance_net,
  attribute = "log_gdp",
  method = "correlation",
  significance_test = TRUE)

knitr::kable(gdp_homophily, digits=3, align='c')

# gate substantive claims on magnitude (|r| >= 0.15) so a near-zero
# correlation does not get described as a clean effect just because n is large
economic_summary <- paste0(
  "**Economic Development Homophily Results:**\n\n",
  "- Homophily correlation: ", round(gdp_homophily$homophily_correlation, 3), "\n",
  "- Similarity among allies: ", round(gdp_homophily$mean_similarity_connected, 3), "\n",
  "- Similarity among non-allies: ", round(gdp_homophily$mean_similarity_unconnected, 3), "\n",
  "- P-value: ", round(gdp_homophily$p_value, 3), "\n",
  dplyr::case_when(
    gdp_homophily$p_value >= 0.05 ~
      "→ Economic development levels don't significantly predict alliance patterns\n",
    gdp_homophily$homophily_correlation >= 0.15 ~
      "→ Countries at similar development levels are more likely to form alliances\n",
    gdp_homophily$homophily_correlation > 0 ~
      "→ Statistically significant but very small (|r| < 0.15) -- consistent with a mild similar-development preference, but the dyad count is doing most of the work behind the p-value\n",
    TRUE ~
      "→ Slight tendency for countries at *different* development levels to ally (small heterophily)\n"
  )
)
```

```{r, echo=FALSE, results='asis'}
cat(economic_summary)
```

## 3. Regional Clustering in International Cooperation

Do countries primarily form alliances within their own regions, or are alliances more globally distributed? Regional patterns provide another example of categorical homophily:

```{r}
region_homophily <- homophily(
  alliance_net,
  attribute = "region",
  method = "categorical",
  significance_test = TRUE)

knitr::kable(region_homophily, digits=3, align='c')

regional_summary <- paste0(
  "**Regional Clustering Results:**\n\n",
  "- Homophily score: ", round(region_homophily$homophily_correlation, 3), "\n",
  "- Within-region alliances: ", round(region_homophily$mean_similarity_connected, 3), "\n",
  "- Cross-region alliances: ", round(region_homophily$mean_similarity_unconnected, 3), "\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(regional_summary)
```

## 4. Who Forms Alliances With Whom? Using `mixing_matrix()`

The `mixing_matrix()` function reveals detailed interaction patterns between different types of actors in your network. This is crucial for understanding not just *if* certain types connect, but *how much* and *with whom*.

A few summary statistics will show up below: **assortativity** is a -1 to 1 coefficient that captures how strongly ties prefer same-attribute partners (positive = same-type, 0 = random, negative = opposites-attract). **Modularity** rewards within-group edges over what you'd expect by chance, so higher values mean ties cluster more strongly within attribute groups. **Entropy** quantifies how spread out (versus concentrated) the mixing pattern is across the cells. **Diagonal proportion** is the share of ties that fall on the diagonal of the mixing matrix (i.e., within the same category).

### 📊 Democracy Mixing Matrix

```{r}
regime_mixing <- mixing_matrix(
  alliance_net,
  attribute = "regime_type",
  normalized = TRUE
)

knitr::kable(round(regime_mixing$mixing_matrices[[1]], 3),
             caption = "Regime Type Alliance Matrix (normalized)",
             align = "c")

regime_mixing_summary <- paste0(
  "**Key Insights from mixing_matrix():**\n\n",
  "- Assortativity: ", round(regime_mixing$summary_stats$assortativity, 3), "\n",
  "  (Positive = similar types connect more; Negative = different types connect more)\n",
  "- Proportion of within-type alliances: ", round(regime_mixing$summary_stats$diagonal_proportion, 3), "\n",
  "  (Higher values indicate more homophily)\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(regime_mixing_summary)
```

#### How to Read the Mixing Matrix:

- **Rows**: Source regime type
- **Columns**: Target regime type  
- **Values**: Proportion of ties from row type to column type
- **Diagonal**: Within-type alliances (homophily)

### 🌍 Regional Alliance Patterns with Row Normalization

```{r}
region_mixing <- mixing_matrix(
  alliance_net,
  attribute = "region",
  normalized = TRUE,
  by_row = TRUE)

regional_mixing_header <- "**Regional Alliance Matrix (row-normalized):**\n\n"
```

```{r, echo=FALSE, results='asis'}
cat(regional_mixing_header)
```

```{r}
knitr::kable(round(region_mixing$mixing_matrices[[1]], 3),
             caption = "Regional Alliance Matrix (row-normalized)",
             align = "c")
```

#### Visualizing Regional Alliance Patterns

```{r, fig.width=9, fig.height=7}
plot_mixing_matrix(
    region_mixing,
    show_values = TRUE,
    value_digits = 2,
    text_size = 3,
    text_color_threshold = .7,
    diagonal_emphasis = TRUE,
    reorder_categories = FALSE
) +
  labs(title = "Regional Alliance Patterns",
       subtitle = "Within-region vs cross-region alliance formation",
       x = "Allied with region",
       y = "From region") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10))
```

The regional mixing matrix reveals strong regional clustering in alliance formation. The emphasized diagonal shows that most regions primarily form alliances within their own geographic area, with some notable exceptions for cross-regional partnerships driven by strategic interests.

### 🔀 Cross-Dimensional Analysis: Region × Regime Type

A unique feature of `mixing_matrix()` is analyzing interactions across different attributes:

```{r}
cross_mixing <- mixing_matrix(
  alliance_net,
  attribute = "regime_type",
  row_attribute = "region",
  normalized = TRUE)

cross_mixing_header <- "**How different regime types form alliances across regions:**\n\n"
```

```{r, echo=FALSE, results='asis'}
cat(cross_mixing_header)
```

```{r}
knitr::kable(round(cross_mixing$mixing_matrices[[1]], 3),
             caption = "Cross-dimensional Analysis: Regime Types Across Regions",
             align = "c")
```

## 5. Analyzing Relationship-Level Factors with `dyad_correlation()`

A **dyad** is just a pair of actors (here, a pair of countries), and a **dyadic variable** describes the relationship itself (e.g., distance between two capitals, shared language) rather than either actor alone. The `dyad_correlation()` function examines how relationship-level (dyadic) variables correlate with network ties. This is essential for understanding what factors at the relationship level predict connections.

### 🌍 Geographic Distance and Alliance Formation

```{r}
geo_correlation <- dyad_correlation(
  alliance_net,
  dyad_vars = "geographic_distance",
  method = "pearson",
  significance_test = TRUE
)

geo_summary <- paste0(
  "**Geographic Distance and Alliance Formation (dyad_correlation results):**\n\n",
  "- Correlation coefficient: ", round(geo_correlation$correlation, 3), "\n",
  "- P-value: ", round(geo_correlation$p_value, 3), "\n",
  "- Number of dyads analyzed: ", geo_correlation$n_pairs[1], "\n\n",
  if(geo_correlation$correlation < -0.1 && geo_correlation$p_value < 0.05) {
    "✓ Geography matters: Countries form more alliances with nearby nations.\n  (Negative correlation = shorter distance, more alliances)\n"
  } else if(geo_correlation$correlation > 0.1 && geo_correlation$p_value < 0.05) {
    "✗ Surprising: Greater distance associated with more alliances.\n  (Positive correlation = greater distance, more alliances)\n"
  } else {
    "→ Geography shows no clear effect on alliance patterns.\n  (No significant correlation detected)\n"
  }
)
```

```{r, echo=FALSE, results='asis'}
cat(geo_summary)
```

### 🤝 Analyzing Multiple Dyadic Variables

```{r}
multi_dyad_correlation <- dyad_correlation(
  alliance_net,
  dyad_vars = c("geographic_distance", "alliance_intensity", "defense_alliance"),
  method = "pearson",
  significance_test = TRUE
)

multi_dyad_summary <- paste0(
  "**Multiple Dyadic Variables Analysis:**\n\n",
  paste(sapply(1:nrow(multi_dyad_correlation), function(i) {
    paste0(
      "**", multi_dyad_correlation$dyad_var[i], ":**\n",
      "  - Correlation: ", round(multi_dyad_correlation$correlation[i], 3), "\n",
      "  - P-value: ", round(multi_dyad_correlation$p_value[i], 3), "\n"
    )
  }), collapse = "\n")
)
```

```{r, echo=FALSE, results='asis'}
cat(multi_dyad_summary)
```

## 6. Comprehensive Analysis with `attribute_report()`

**Centrality** measures how important / well-positioned each actor is in the network -- e.g., `degree` counts a node's ties, `betweenness` counts how often a node lies on the shortest path between two others, and `closeness` measures how short the average distance is from a node to everyone else. **Heterophily** is the opposite of homophily: a preference for connecting to *unlike* partners.

The `attribute_report()` function combines the previous analyses into a comprehensive report.

### 🚀 Running the Complete Analysis

```{r}
comprehensive_analysis <- attribute_report(
  alliance_net,
  node_vars = c("region", "regime_type", "democracy", "log_gdp", "mil_capability"),
  dyad_vars = c("geographic_distance", "alliance_intensity", "defense_alliance"),
  include_centrality = TRUE,
  include_homophily = TRUE,
  include_mixing = TRUE,
  include_dyadic_correlations = TRUE,
  centrality_measures = c("degree", "betweenness", "closeness"),
  significance_test = TRUE
)
```

`attribute_report` returns a list with multiple components

* **`homophily_analysis`**: Tests for each node attribute
* **`mixing_matrices`**: Interaction patterns for categorical variables
* **`centrality_correlations`**: How attributes relate to network position
* **`dyadic_correlations`**: How dyad attributes predict ties

### 📋 Extracting Key Findings from the Summary

```{r}
homophily_header <- paste0(
  "**=== HOMOPHILY ANALYSIS ===**\n\n",
  "Do similar countries form more alliances?\n\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(homophily_header)
homophily_summary <- comprehensive_analysis$homophily_analysis

# cohen-style magnitude bins for correlation: ~.1 small, ~.3 medium, ~.5 large
homophily_display <- homophily_summary |>
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "n.s."
    ),
    interpretation = case_when(
      homophily_correlation > 0.5 & p_value < 0.05 ~ "Strong homophily",
      homophily_correlation > 0.3 & p_value < 0.05 ~ "Moderate homophily",
      homophily_correlation > 0.15 & p_value < 0.05 ~ "Weak homophily",
      homophily_correlation > 0 & p_value < 0.05 ~ "Very weak (sig. but small)",
      p_value >= 0.05 ~ "No significant pattern",
      TRUE ~ "Heterophily"
    )
  )

knitr::kable(
  homophily_display[, 
  c(
    "attribute", "method", 
    "homophily_correlation", 
    "p_value", "significance", 
    "interpretation")],
  caption = "Homophily Analysis Results",
  col.names = c("Attribute", "Method", "Homophily Correlation", "P-value", "Significance", "Interpretation"),
  digits = 3,
  align = "c"
)
```

```{r}
power_header <- paste0(
  "**=== POWER AND INFLUENCE ===**\n\n",
  "What makes countries central in the alliance network?\n\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(power_header)
centrality_summary <- comprehensive_analysis$centrality_correlations

centrality_display <- centrality_summary |>
  mutate(
    interpretation = case_when(
      correlation > 0.3 & p_value < 0.05 ~ "Strongly associated with centrality",
      correlation > 0.1 & p_value < 0.05 ~ "Moderately associated with centrality",
      correlation < -0.1 & p_value < 0.05 ~ "Negatively associated with centrality",
      p_value < 0.05 ~ "Weakly associated with centrality",
      TRUE ~ "Not significantly related to centrality"
    )
  ) |>
  arrange(p_value)

knitr::kable(
  centrality_display[1:10, 
  c(
    "node_var", "centrality_measure", 
    "correlation", "p_value", 
    "interpretation")],
  caption = "Top 10 Centrality-Attribute Correlations",
  col.names = c("Node Variable", "Centrality Measure", "Correlation", "P-value", "Interpretation"),
  digits = 3,
  align = "c"
)
```

```{r}
relationship_header <- paste0(
  "**=== RELATIONSHIP FACTORS ===**\n\n",
  "What dyadic factors predict alliance formation?\n\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(relationship_header)
dyadic_summary <- comprehensive_analysis$dyadic_correlations
knitr::kable(dyadic_summary[, c("dyad_var", "correlation", "p_value")],
             caption = "Dyadic Variables Analysis",
             col.names = c("Dyadic Variable", "Correlation", "P-value"),
             digits = 3,
             align = "c")
```

## 7. Testing Specific IR Hypotheses

### Hypothesis 1: Democratic Peace

```{r}
# binary democracy indicator
nodal_data_binary <- nodal_data |>
  mutate(is_democracy = ifelse(regime_type == "Democracy", 1, 0))

alliance_net_binary <- add_node_vars(
  alliance_net,
  nodal_data_binary[, c("actor", "is_democracy")],
  actor = "actor")

dem_peace_test <- homophily(
  alliance_net_binary,
  attribute = "is_democracy",
  method = "categorical",
  significance_test = TRUE)

# flag both significance and magnitude so a tiny r does not get described
# as a clean "democracies prefer democracies" effect just because p<0.05
dem_peace_summary <- paste0(
  "**Democratic Peace Hypothesis Test:**\n\n",
  "- Effect size: ", round(dem_peace_test$homophily_correlation, 3), "\n",
  "- P-value: ", round(dem_peace_test$p_value, 3), "\n",
  "- Conclusion: ", dplyr::case_when(
    dem_peace_test$p_value >= 0.05 ~
      "No significant democratic preference",
    dem_peace_test$homophily_correlation >= 0.15 ~
      "Democracies significantly prefer forming alliances with other democracies",
    dem_peace_test$homophily_correlation > 0 ~
      "Statistically significant but very small same-regime preference; large dyad count drives the significance flag more than the effect size",
    TRUE ~
      "Statistically significant heterophily: democracies tend to ally with non-democracies"
  ), "\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(dem_peace_summary)
```

### Hypothesis 2: Power Politics

Do powerful countries (high military capability) primarily form alliances with other powerful countries?

```{r}
power_homophily <- homophily(
  alliance_net,
  attribute = "mil_capability",
  method = "correlation",
  significance_test = TRUE)

# gate substantive claims on |r| so a near-zero correlation does not get
# labeled "heterophily" simply because the dyad count makes every p tiny
power_politics_summary <- paste0(
  "**Power Politics Hypothesis:**\n\n",
  "- Correlation: ", round(power_homophily$homophily_correlation, 3), "\n",
  "- P-value: ", round(power_homophily$p_value, 3), "\n",
  "- Interpretation: ", dplyr::case_when(
    power_homophily$p_value >= 0.05 ~
      "No evidence of power-based alliance preferences",
    power_homophily$homophily_correlation >= 0.15 ~
      "Powerful countries prefer forming alliances with other powerful countries",
    power_homophily$homophily_correlation <= -0.15 ~
      "Powerful countries tend to form alliances with less powerful countries (heterophily)",
    TRUE ~
      "Statistically significant but very small (|r| < 0.15); read as 'consistent with mild power-mixing' rather than a clean homophily/heterophily effect"
  ), "\n"
)
```

```{r, echo=FALSE, results='asis'}
cat(power_politics_summary)
```

## 8. Visualizing Network Patterns

And as seen in other vignettes we can use the `plot()` function to visualize the network with node attributes and edge weights:

### Network Visualization by Attributes

```{r}
# attach degree and other actor stats so we can map them to node aesthetics
alliance_net <- add_node_vars(
  alliance_net,
  summary_actor(alliance_net),
  actor = "actor"
)
```

```{r, fig.width=8, fig.height=6}
plot(alliance_net,
     node_color_by = "region",
     node_color_label = "",
     node_shape_by = "regime_type",
     node_shape_label = "",
     node_size_by = "degree",
     node_size_label = "Degree",
     node_fill = "white",
     edge_color = "grey50",
     edge_linewidth = 0.5,
     edge_alpha_label = 'Alliance Strength (scaled)',
     layout = "nicely",
     seed = 6886) +
  ggtitle("ATOP Network") +
  theme(legend.position = 'right')
```

### Visualizing Homophily Results

```{r, fig.width=10, fig.height=6}
plot_homophily(comprehensive_analysis$homophily_analysis,
               type = "comparison") +
  labs(title = "Alliance Formation Patterns: Which Attributes Matter?",
       subtitle = "Homophily analysis reveals how similarity drives international cooperation")
```

### Visualizing Centrality Patterns

```{r, fig.width=10, fig.height=8}
centrality_viz <- comprehensive_analysis$centrality_correlations |>
  filter(p_value < 0.1) |>
  mutate(
    significant = p_value < 0.05,
    node_var = factor(node_var),
    centrality_measure = factor(
      centrality_measure,
      levels = c("degree", "betweenness", "closeness"))
  )

if(nrow(centrality_viz) > 0) {
  ggplot(centrality_viz, aes(x = correlation, y = node_var, color = significant)) +
    geom_segment(
      aes(
        x = 0, xend = correlation, y = node_var, yend = node_var),
      size = 1) +
    geom_point(size = 3) +
    facet_wrap(~centrality_measure, ncol = 1) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    scale_color_manual(
        values = c("FALSE" = "gray60", "TRUE" = "#2E86AB"),
        labels = c("FALSE" = "Not significant", "TRUE" = "(p < 0.05)")
    ) +
    labs(title = "What Makes Countries Central in the Alliance Network?",
         subtitle = "Correlation between node attributes and centrality measures",
         x = "Correlation with Centrality",
         y = "Node Attribute",
         color = "") +
    theme_bw() +
    theme(
        panel.border = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "top",
        panel.grid.major.y = element_blank()
    )
} else {
  no_centrality_msg <- "**No significant centrality correlations to visualize.**\n"
  cat(no_centrality_msg)
}
```

### Mixing Matrix Heatmap

We can use the `plot_mixing_matrix()` function to visualize the mixing patterns as a heatmap:

```{r, fig.width=8, fig.height=6}
plot_mixing_matrix(
    regime_mixing,
    show_values = TRUE,
    diagonal_emphasis = TRUE,
    text_color_threshold=.9
  ) +
  labs(title = "Regime Type Alliance Patterns",
       subtitle = "How different regime types interact in the network",
       x = "Allied with...",
       y = "Regime type")
```

The heatmap shows the alliance patterns between different regime types. The diagonal cells (emphasized with black borders) represent within-type alliances, while off-diagonal cells show cross-type alliances. Darker blue indicates higher proportions of alliances, consistent with the small but detectable regime-type homophily in alliance formation.

## 9. Working with Longitudinal Networks

A **longitudinal network** is a network observed over multiple time periods (one snapshot per period), in contrast to a **cross-sectional network** that captures a single point in time. All the attribute analysis functions in netify work seamlessly with longitudinal networks. Let's demonstrate this by creating a longitudinal alliance network and running the same analyses across multiple time periods.

### Creating a Longitudinal Network

```{r}
# longitudinal alliance network for 2010-2014
alliance_net_longit <- netify(
  cow_dyads,
  actor1 = 'ccode1',
  actor2 = 'ccode2',
  time = 'year',
  symmetric = TRUE,
  weight = 'cooperation'
)

alliance_net_longit
```

### Adding Attributes to Longitudinal Networks

```{r}
# nodal data for all time periods
nodal_data_longit <- cow_dyads |>
  select(year, ccode1, region1, v2x_polyarchy1, log_gdp1, cinc1) |>
  distinct() |>
  rename(
    time = year,
    actor = ccode1,
    region = region1,
    democracy = v2x_polyarchy1,
    log_gdp = log_gdp1,
    mil_capability = cinc1
  ) |>
  mutate(
    regime_type = case_when(
      democracy >= 0.6 ~ "Democracy",
      democracy >= 0.4 ~ "Hybrid",
      democracy < 0.4 ~ "Autocracy",
      TRUE ~ "Unknown"
    )
  )

alliance_net_longit <- add_node_vars(
  alliance_net_longit,
  nodal_data_longit,
  actor = "actor",
  time = "time"
)

dyad_data_longit <- cow_dyads |>
  select(year, ccode1, ccode2, log_capdist, alliance_intensity, defense_alliance) |>
  rename(
    time = year,
    actor1 = ccode1,
    actor2 = ccode2,
    geographic_distance = log_capdist
  )

alliance_net_longit <- add_dyad_vars(
  alliance_net_longit,
  dyad_data = dyad_data_longit,
  actor1 = "actor1",
  actor2 = "actor2",
  time = "time",
  dyad_vars = c("geographic_distance", "alliance_intensity", "defense_alliance"),
  dyad_vars_symmetric = c(TRUE, TRUE, TRUE)
)
```

### Homophily Analysis Across Time

```{r}
democracy_homophily_longit <- homophily(
  alliance_net_longit,
  attribute = "democracy",
  method = "correlation",
  significance_test = TRUE
)

democracy_homophily_longit

# summary of trends across time periods
homophily_trends <- democracy_homophily_longit |>
  group_by(net) |>
  summarise(
    avg_homophily = mean(homophily_correlation, na.rm = TRUE),
    significant = any(p_value < 0.05, na.rm = TRUE)
  )

knitr::kable(homophily_trends, 
             caption = "Democracy Homophily Trends Over Time",
             digits = 3)
```

### Visualizing Longitudinal Homophily

```{r, fig.width=10, fig.height=6}
plot_homophily(democracy_homophily_longit, type = "temporal") +
  labs(title = "Democracy Homophily in Alliance Networks Over Time",
       subtitle = "Tendency for democracies to ally with other democracies")
```

If you want to see the distribution for a specific time period, you can extract that period first:

```{r, fig.width=8, fig.height=5}
# extract 2012 slice for a single-period distribution plot
alliance_2012 <- subset(alliance_net_longit, time = '2012')
democracy_homo_2012 <- homophily(
  alliance_2012,
  attribute = "democracy",
  method = "correlation"
)

plot_homophily(democracy_homo_2012, alliance_2012,
               type = "distribution",
               attribute = "democracy",
               method = "correlation") +
  labs(subtitle = "Distribution of democracy similarity scores in 2012 alliance network")
```

### Mixing Matrices Over Time

```{r}
regime_mixing_longit <- mixing_matrix(
  alliance_net_longit,
  attribute = "regime_type",
  normalized = TRUE
)

# summary statistics across each time period
mixing_summary <- regime_mixing_longit$summary_stats |>
  select(net, assortativity, diagonal_proportion) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

knitr::kable(mixing_summary,
             caption = "Regime Type Mixing Patterns Over Time",
             col.names = c("Year", "Assortativity", "Within-Type %"))
```

### Dyadic Correlations Across Time

```{r}
geo_correlation_longit <- dyad_correlation(
  alliance_net_longit,
  dyad_vars = "geographic_distance",
  method = "pearson",
  significance_test = TRUE
)

geo_summary_longit <- geo_correlation_longit |>
  select(net, correlation, p_value, n_pairs) |>
  mutate(
    significant = ifelse(p_value < 0.05, "*", ""),
    correlation = round(correlation, 3),
    p_value = round(p_value, 3)
  )

knitr::kable(geo_summary_longit,
             caption = "Geographic Distance and Alliance Formation Over Time",
             col.names = c("Year", "Correlation", "P-value", "N Dyads", "Sig."))
```

### Comprehensive Longitudinal Analysis

```{r}
comprehensive_longit <- attribute_report(
  alliance_net_longit,
  node_vars = c("region", "regime_type", "democracy", "log_gdp"),
  dyad_vars = c("geographic_distance", "alliance_intensity"),
  include_centrality = TRUE,
  include_homophily = TRUE,
  include_mixing = TRUE,
  include_dyadic_correlations = TRUE,
  centrality_measures = c("degree", "betweenness"),
  significance_test = TRUE
)

if (!is.null(comprehensive_longit$homophily_analysis)) {
  longit_patterns <- comprehensive_longit$homophily_analysis |>
    filter(attribute == "democracy") |>
    select(net, homophily_correlation, p_value) |>
    mutate(
      trend = case_when(
        net == min(net) ~ "Start",
        net == max(net) ~ "End",
        TRUE ~ "Middle"
      )
    )

  longit_patterns
} else {
  knitr::asis_output(
    "Note: homophily for longitudinal networks is currently limited. For full coverage, analyze each time period separately.\n"
  )
}
```

## tl;dr

1. **From `homophily()`**: 
   - Whether democracies truly form more alliances with each other
   - If economic similarity drives alliance patterns
   - The strength of regional clustering in alliance formation

2. **From `mixing_matrix()`**:
   - Detailed patterns of who forms alliances with whom
   - Whether alliances cross regime type boundaries
   - How different regions form alliances globally

3. **From `dyad_correlation()`**:
   - The role of geographic distance in shaping alliance formation
   - How alliance types (defense, offense, etc.) cluster
   - Which relationship factors matter most for alliances

4. **From `attribute_report()`**:
   - What attributes make countries central/influential
   - Complete homophily patterns across all variables
   - Comprehensive view of all network-attribute relationships
