---
title: "Ego Networks"
description: "Build and compare ego-centered networks around focal actors."
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ego Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = FALSE,
    fig.width = 7,
    fig.height = 5
)

```

This vignette shows how `netify` handles ego networks—the network surrounding a specific actor and their immediate connections.

## what are ego networks?

An **ego network** focuses on one actor (the "ego") and includes:

- The **ego**: Your focal actor
- The **alters**: Actors directly connected to the ego  
- **Alter-alter ties**: Connections between the ego's neighbors

Think of it as zooming in on one node and asking: "Who are their connections, and how are those connections related to each other?"

## examples

Pakistan's diplomatic neighborhood includes not just major powers (US, China) but regional rivals (India) and neighbors (Afghanistan). The connections *between* these alters matter—does Pakistan broker between disconnected partners or navigate a densely connected region?

A senator's ego network reveals their coalition partners, but also whether those partners work together or represent distinct constituencies the senator must balance.

An activist organization's ego network shows both allies and the broader movement structure—are they bridging disconnected groups or embedded in a tight cluster?

## why ego networks matter

Ego networks let you apply network thinking without needing complete network data. You can study:

- How actors manage competing relationships
- Whether similar actors cluster together (homophily)
- How an actor's position changes across relationships
- Whether an actor brokers between otherwise disconnected groups

```{r}
library(netify)
library(dplyr)
library(ggplot2)
```

## getting started: extract an ego network

We'll use data from the Integrated Crisis Early Warning System (ICEWS) to demonstrate:

```{r}
data(icews)

# verbal cooperation between countries
netlet <- netify(
    icews,
    actor1 = "i", actor2 = "j",
    time = "year",
    weight = "verbCoop",
    nodal_vars = c("i_polity2", "i_log_gdp", 'i_region'),
    dyad_vars = c("matlCoop", "verbConf")
)
```

### extract your first ego network

The `ego_netify()` function makes extraction straightforward:

```{r}
pakistan_ego_net <- ego_netify(
    netlet,
    ego = "Pakistan"
    )

print(pakistan_ego_net)
head(summary(pakistan_ego_net)[, c("net", "num_actors", "density", "num_edges")])
```

## key features

### 1. network statistics

```{r}
ngbd_summ <- summary(pakistan_ego_net)
head(ngbd_summ)
```

Pakistan's network shows high density (0.79-0.97)—most countries that interact with Pakistan also interact with each other. The network size fluctuates between 24-49 actors across years, but that high density persists. This isn't a hub-and-spoke pattern where Pakistan connects otherwise isolated actors; it's a dense web where multilateral dynamics likely matter.

```{r}
plot_graph_stats(ngbd_summ) +
    scale_x_discrete(
        breaks = seq(2002, 2014, by = 4)
    ) +
    theme(
        axis.text.x=element_text(angle=45, hjust=1)
    )
```

Notice the consistently high transitivity (~0.84-0.97). Pakistan's partners tend to cooperate with each other, creating a tightly clustered neighborhood rather than structural holes Pakistan could exploit.

### 2. actor-level analysis

```{r}
ngbd_actor_summ <- summary_actor(pakistan_ego_net)
head(ngbd_actor_summ)
```

Beyond degree (number of connections), look at eigenvector centrality—it captures importance based on connections to other important actors. China's high and growing eigenvector centrality reflects not just bilateral ties with Pakistan but its connections to other key players in Pakistan's neighborhood.

```{r}
plot_actor_stats(ngbd_actor_summ,
    across_actor=FALSE,
    specific_actors=c('United States', 'China')
    ) +
    scale_x_discrete(
        breaks = seq(2002, 2014, by = 4)
    ) +
    theme(
        axis.text.x=element_text(angle=45, hjust=1)
    )
```

The diverging trajectories are striking. China's rising centrality measures show growing connectedness in Pakistan's neighborhood, while the US shows more variable engagement despite maintaining presence.

### 3. visualization

By default, `plot()` highlights the ego:

```{r, fig.width=8, fig.height=6}
plot(pakistan_ego_net)
```

### 4. advanced visualization

For clearer visualization of alter-alter relationships, remove the ego's edges—we already know everyone connects to Pakistan:

```{r}
# attach actor stats so we can map them to node aesthetics
pakistan_ego_net <- add_node_vars(
    pakistan_ego_net,
    summary_actor(pakistan_ego_net),
    "actor", "time"
)

# strip ties incident to ego so the alter-alter structure stands on its own
pakistan_no_ego_edges <- remove_ego_edges(pakistan_ego_net)
```

```{r, fig.width=10, fig.height=8}
plot(pakistan_no_ego_edges,
    layout = "hierarchical",  
    node_size_by = "i_log_gdp",
    node_size_label = "Log(GDP)",
    highlight = c("Pakistan", "United States", "China"),
    highlight_color = c(
        "Pakistan" = '#01411cff', 
        "United States" = "#0A3161",
        "China" = "#EE1C25",
        "Other" = 'grey40'),
    edge_linewidth = 0.05,
    mutate_weight = log1p,
    edge_alpha_label = 'Log(Verb.\n Coop.)',
    time_filter = as.character(seq(2002, 2014, 4))
    ) +
    labs(
        title = "Pakistan's Diplomatic Neighborhood",
        subtitle = "Node size = GDP; Edges show cooperation between Pakistan's partners") +
    theme(legend.position = 'right')
```

The visualization reveals how Pakistan's partners form a dense web of relationships among themselves, with major powers like the US and China occupying central positions.

## comparing ego networks

### 5. compare across actors

One strength of ego network analysis is systematic comparison:

```{r}
powers <- c("United States", "China", "Russian Federation", "India", "Pakistan")
ego_networks <- lapply(powers, function(country) {
    ego_netify(netlet, ego = country)
})
names(ego_networks) <- powers

summaries <- lapply(ego_networks, summary)
comparison_df <- bind_rows(summaries, .id = "country")

ggplot(comparison_df,
    aes(x = net, y = density, color = country, group = country)) +
    geom_line(linewidth = 1.2) +
    labs(title = "Ego Network Density Over Time",
         subtitle = "How interconnected are different countries' diplomatic neighborhoods?",
         x = "Year", y = "Density") +
    theme_bw() +
    theme(
        panel.border = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "top"
    )
```

Pakistan and India maintain the highest density—their partners are highly interconnected. The US shows lower density, suggesting more of a hub-and-spoke pattern. Russia's volatile pattern may reflect shifting alliances during this period.

### 6. test for homophily

Do birds of a feather flock together? Let's test if similar regime types cluster:

```{r}
pakistan_homophily <- homophily(
    pakistan_ego_net,
    attribute = "i_polity2",
    method = "correlation"
)

plot_homophily(pakistan_homophily, type = "temporal") +
    labs(title = "Do Democracies Cluster in Pakistan's Network?",
         subtitle = "Testing regime type homophily over time")
```

The results show weak and fluctuating homophily: regime type is not consistently associated with cooperation patterns in Pakistan's neighborhood. Geopolitics appears more central than ideological alignment in this example.

### 7. control neighborhood boundaries

Sometimes you want to focus on only the strongest relationships:

```{r}
# default includes all connections
pakistan_all <- ego_netify(netlet, ego = "Pakistan")

# only strong connections (threshold = 50 cooperation events)
pakistan_strong <- ego_netify(netlet, ego = "Pakistan", threshold = 50)

all_size <- mean(summary(pakistan_all)$num_actors)
strong_size <- mean(summary(pakistan_strong)$num_actors)

tibble(
  Network = c("All connections", "Strong only (>50)"),
  `Average Size` = round(c(all_size, strong_size), 1)
) |>
  knitr::kable()
```

### 8. different relationship types

Networks of cooperation and conflict often follow different logics:

```{r}
conflict_net <- netify(
    icews,
    actor1 = "i", actor2 = "j",
    time = "year",
    weight = "verbConf"
)

# cooperation vs conflict ego networks for pakistan
pak_coop_ego <- ego_netify(netlet, ego = "Pakistan")
pak_conf_ego <- ego_netify(conflict_net, ego = "Pakistan")

coop_stats <- summary(pak_coop_ego)
conf_stats <- summary(pak_conf_ego)

comparison <- tibble(
  Metric = c("Average actors", "Average density"),
  Cooperation = c(
    round(mean(coop_stats$num_actors), 1),
    round(mean(coop_stats$density), 3)
  ),
  Conflict = c(
    round(mean(conf_stats$num_actors), 1),
    round(mean(conf_stats$density), 3)
  )
)

knitr::kable(comparison)
```

Pakistan's conflict network is smaller but still substantial. The lower average density in conflict (about 0.65 vs 0.90) suggests conflicts are more bilateral while cooperation tends to be multilateral.

### 9. track network evolution

How stable are relationships over time?

```{r}
pakistan_comparison <- compare_networks(pakistan_ego_net, what = "edges")

# zoom in on one transition
changes_2010_2011 <- pakistan_comparison$edge_changes$`2010_vs_2011`

stability_ratio <- round(
  changes_2010_2011$maintained / 
  (changes_2010_2011$maintained + changes_2010_2011$removed), 3)

tibble(
  Transition = "2010 to 2011",
  Added = changes_2010_2011$added,
  Removed = changes_2010_2011$removed,
  Maintained = changes_2010_2011$maintained,
  `Stability Ratio` = stability_ratio
) |>
  knitr::kable()
```

A stability ratio of 0.416 indicates moderate turnover -- about 58% of relationships don't persist year-to-year.

### 10. compare network structures: rising vs established powers

How do the US and China structure their diplomatic neighborhoods differently?

```{r}
us_ego <- ego_netify(netlet, ego = "United States")
china_ego <- ego_netify(netlet, ego = "China")

structural_comp <- compare_networks(
    list("US" = us_ego, "China" = china_ego),
    what = 'structure'
)

comp_summary <- structural_comp$summary
avg_props <- comp_summary |>
    group_by(network) |>
    summarise(
        `Avg. Nodes` = round(mean(num_actors), 0),
        `Avg. Density` = round(mean(density), 3),
        `Avg. Transitivity` = round(mean(transitivity), 3),
        .groups = 'drop'
    )

knitr::kable(avg_props)
```

The US maintains a larger but less dense network—more of a hub-and-spoke pattern. China's denser network suggests its partners are more interconnected, potentially reflecting regional concentration.

Let's look at the temporal patterns:

```{r}
us_summary <- summary(us_ego)
china_summary <- summary(china_ego)

temporal_comparison <- bind_rows(
    us_summary |> mutate(network = "US"),
    china_summary |> mutate(network = "China")
)

temporal_comparison$year <- as.integer(temporal_comparison$net)

ggplot(temporal_comparison, aes(x = year, y = density, color = network)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2) +
    scale_color_manual(values = c("US" = "#377eb8", "China" = "#e41a1c")) +
    labs(title = "Ego Network Density: US vs China",
         subtitle = "China's increasing density reflects growing multilateral engagement",
         x = "Year", y = "Density") +
    theme_bw() +
    theme(
        panel.border = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "top"
    )
```

Now let's examine how similar their cooperation patterns are:

```{r}
comp_2012 <- compare_networks(
    list(
        "US_2012" = subset(us_ego, time = "2012"),
        "China_2012" = subset(china_ego, time = "2012")
    ),
    what = "edges",
    method = "all"
)

edge_stats <- comp_2012$summary
edge_changes <- comp_2012$edge_changes[[1]]

tibble(
  Metric = c("Edge correlation", "Jaccard similarity", 
             "Unique to US", "Unique to China", "Shared"),
  Value = c(
    round(edge_stats$correlation, 3),
    round(edge_stats$jaccard, 3),
    edge_changes$removed,
    edge_changes$added,
    edge_changes$maintained
  )
) |>
  knitr::kable()
```

The very high correlation (0.991) but moderate Jaccard similarity (0.645) tells an interesting story: when both countries engage with a partner, they do so in similar ways, but the US maintains many more unique relationships (2,222 vs 1,104).

Finally, who's in these networks?

```{r}
# compare node composition
node_comp <- compare_networks(
    list("US" = us_ego, "China" = china_ego),
    what = "nodes"
)

# the 75% overlap reflects shared major partners
# but the us's 24 unique partners vs china's 8 shows its broader global reach
knitr::kable(node_comp$summary)
```

## tl;dr

```r
# extract ego network
ego_net <- ego_netify(netlet, ego = "Actor Name")

# basic analysis
summary(ego_net)                    # network-level stats
summary_actor(ego_net)              # actor-level stats
plot(ego_net)                       # visualize
plot(remove_ego_edges(ego_net))     # focus on alter-alter ties

# advanced analysis
homophily(ego_net, attribute = "democracy")     # test homophily
compare_networks(ego_net, what = "edges")       # temporal comparison
compare_networks(list(ego1, ego2), what = "edges")  # cross-sectional comparison
```

## references

Boschee, E., Lautenschlager, J., O'Brien, S., Shellman, S., Starz, J., & Ward, M. (2015). ICEWS Coded Event Data. Harvard Dataverse. DOI 10.7910/DVN/28075.
