---
title: "Conflict Application: Event Data"
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conflict Application: Event Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  dev = 'png', dpi = 150,
  cache = FALSE,
  echo = TRUE
)
```

##  Vignette Summary 

Below is a demonstration of how to use the **netify** R package to build and analyze various conflict networks. We will walk through data preparation, network construction, summary statistics, visualizations, and special use cases like weighted networks, longitudinal networks, and bipartite networks.

A common data type used to create conflict networks is in the structure of event data, where actors are repeated across rows but there is no specific variable that denotes the 'edge' as shown in our previous example.

We show how to use `netify` and UCDP data as just one example of potential applications to intrastate event data. The first step is to go to https://ucdp.uu.se/downloads/ and download the data you want to use. For this tutorial we have downloaded UCDP GED event data version 23.1 and subset the data for the case of Mexico. We also have saved this example internal to the package to that users can easily run the code below.


```{r}
library(netify)
library(ggplot2)
```

(1) Create: generate an aggregated, weighted network of conflict between actors in Mexico. The default here, when the user does not supply a 'weight' value, is for the function to simply count up interactions and return this as the weight. When this is the case, the user will get the warning shown in the example below.

```{r}
# Load UCDP GED data on Mexico
data(mexico)

# Construct unweighted network (number_of_events by default)
mex_network <- netify(
  mexico,
  actor1 = 'side_a',
  actor2 = 'side_b',
  symmetric = TRUE,
  sum_dyads = TRUE,
  diag_to_NA = TRUE,
  missing_to_zero = TRUE
)

# Summaries at the graph and actor levels
summary(mex_network)

actor_stats <- summary_actor(mex_network)
plot_actor_stats(actor_stats)

```

(2) Explore: next we can quickly use the basic plot function in the `netify` package to plot network of violent interactions in Mexico.

```{r, fig.width=6, fig.height=6}
# select 10 random names for plotting
select_names <- rownames(mex_network)
set.seed(12345)
random_indices <- sample(length(select_names), 10)
random_names <- select_names[random_indices]

plot(mex_network,
     add_text = TRUE)
    
plot(mex_network,
     select_text = random_names,
     select_text_display = random_names)
```

The graphs above reveals the need to transform the data to increase interpretability. To do so, we can log the the values of the edges. The `netify` plot function has a built-in parameter for this where the user can pass the function they want to use to transform the weights of the data (see the general `transform_weights` function in the package to transform weights outside of plotting.)

```{r, fig.width=6, fig.height=6}
plot(mex_network,
     select_text = random_names,
     select_text_display = random_names, 
     # log(x+1) to better see range of connects
     weight_transform = log1p
     )
```

Let's make the same graph as above but this time make sure we remove any isolates from the graph as well as refine the edges. In addition, there are some points that are overlapping and difficult to see. To get ideas about how to improve the plot, we might like to inspect each component of the plot. This is an option when the plot parameter, `return_components = TRUE`. We can also directly modify these components, as shown below by adding a label specification.

```{r, fig.width=6, fig.height=6}
# investigate each component of the plot
comp <- plot(
  mex_network,
  remove_isolates = TRUE,
  select_text = random_names,
  select_text_display = random_names,              
  weight_transform = log1p,
  return_components = TRUE
  )

print(comp)

# modify component
comp$base + 
  netify_edge(comp) + 
  labs(alpha='Log(Event Count)') +
  reset_scales() +
  netify_node(comp) +
  netify_text(comp) +
  comp$theme
```

There is, however, a more straightforward way to do this which is via the `edge_alpha_label` parameter inside the netify plot function:

```{r, fig.width=6, fig.height=6}
# and even simpler
plot(mex_network, 
     select_text = random_names,
     select_text_display = random_names,
     weight_transform = log1p,
     edge_alpha_label = "Log(Event Count)")
```

For conflict data, we often like to see which actors are involved in the most conflict. To do this, however, we'd need to calculate degree centrality. To do this we can use the `add_node_vars()` function (there is also a `add_dyad_vars()` function) which takes a netlet object (the network of study), nodal data, and a string to specify the column name in the nodal data that matches with each actor in the network. Below, we use `summary_actor()` to create the nodal data frame. 

```{r, fig.width=6, fig.height=6}
# add degree centrality 
mex_network = add_node_vars(
  mex_network, 
  summary_actor(mex_network), 
  actor='actor' )

plot(mex_network, 
     weight_transform = log1p,
     edge_alpha_label = "Log(Event Count)",
     point_size_var='degree',
     point_size_label='Degree',
     point_alpha=.5,
    check_overlap=TRUE
    )
```


## References

- Sundberg, Ralph and Erik Melander (2013) Introducing the UCDP Georeferenced Event Dataset. Journal of Peace Research 50(4).