---
title: "Event Data"
description: "Build one-mode and two-mode networks from event data."
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Event Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    dev = "png", dpi = 150,
    cache = FALSE,
    echo = TRUE
)
```

## tl;dr

Event data come in two flavors. **One-mode** event data (actor-on-actor: who fought who, who cited whom) becomes a unipartite network. **Two-mode** event data (actor-on-venue: firms in jurisdictions, attendees at meetings) becomes a bipartite network. This vignette walks through both. Along the way we cover transforming weights, single-actor highlighting, applying a ggplot theme, and how to export a figure with `ggsave()`.

```{r}
library(netify)
library(ggplot2)
```

## one-mode event data: ucdp conflict in mexico

A common data type used to build conflict networks is **event data**, where actors are repeated across rows but no single column explicitly encodes an edge weight. `netify()` handles this case by counting interactions when no `weight` is provided.

We use UCDP GED data on Mexico as one example of an application to intrastate event data. The data come from <https://ucdp.uu.se/downloads/> (version 23.1, subset to Mexico); the subset ships internally so the code below runs as-is.

(1) **Create**: generate an aggregated, weighted network of conflict between actors in Mexico. When the user does not supply a `weight` value, `netify()` counts interactions and returns the count as the edge weight, with an informational message indicating that this has happened.

```{r}
# load ucdp ged data on mexico
data(mexico)

# construct count-weighted network (number_of_events by default)
mex_network <- netify(
    mexico,
    actor1 = "side_a",
    actor2 = "side_b",
    symmetric = TRUE,
    sum_dyads = TRUE,
    diag_to_NA = TRUE,
    missing_to_zero = TRUE
)

# summaries at the graph and actor levels
summary(mex_network)

actor_stats <- summary_actor(mex_network)
plot_actor_stats(actor_stats)
```

### tl;dr sidebar: na versus zero in event-count weights

Event counts have a sharp interpretation gap that does not show up in, say, a 0/1 alliance network: the difference between **"these two actors were both in the data and recorded zero events together"** and **"we have no information on whether these two actors interacted at all"** is the difference between a substantive zero and a missing observation. The default `missing_to_zero = TRUE` collapses both into `0`. For routine event-data uses this is what you want — a missing UCDP dyad really does mean "no recorded violent event between these two sides in this slice of the data," which is closer to a substantive zero than a Bayesian "we don't know." But for epidemiology / contact-tracing / sparse rosters, the asymmetry matters and you'll want `missing_to_zero = FALSE`.

A 5-actor toy makes the bookkeeping concrete. Suppose four contact-tracing interviews surfaced four edges among five named individuals; the remaining dyads were never asked about:

```{r}
tiny <- data.frame(
    a1 = c("p1", "p1", "p2", "p3"),
    a2 = c("p2", "p3", "p3", "p4"),
    n_contacts = c(2, 1, 3, 1),
    stringsAsFactors = FALSE
)

# preserve unobserved dyads as na
net_na <- netify(
    tiny,
    actor1 = "a1", actor2 = "a2",
    weight = "n_contacts",
    symmetric = TRUE,
    missing_to_zero = FALSE,
    nodelist = c("p1", "p2", "p3", "p4", "p5")
)

# the raw matrix carries explicit nas for the never-asked dyads
get_raw(net_na)

# summary() reports the missingness fraction in its own column when
# missing_to_zero = false
summary(net_na)[, c("num_actors", "density", "num_edges",
                    "prop_edges_missing", "prop_unknown_edges")]
```

`prop_unknown_edges` only shows up when `missing_to_zero = FALSE` — that is the cue that the netlet is actually carrying NA semantics downstream. Build the same netlet with the default and the NAs are replaced with `0`s and the column disappears; that is fine for UCDP / GED event counts (the rest of this vignette uses the default for exactly that reason) but it is the wrong choice when "we did not measure this dyad" is a meaningful state.

(2) **Explore**: next we use the basic `plot()` method to visualize the network of violent interactions in Mexico. The simplest call labels every actor:

```{r, fig.width=6, fig.height=6}
set.seed(6886)
plot(mex_network, add_text = TRUE)
```

For a less cluttered view, pass a vector of actors to `select_text` / `select_text_display`:

```{r, fig.width=6, fig.height=6}
# select 10 random names for plotting
select_names <- rownames(mex_network)
set.seed(6886)
random_indices <- sample(length(select_names), 10)
random_names <- select_names[random_indices]

plot(mex_network,
    select_text = random_names,
    select_text_display = random_names
)
```

The graphs above reveal the need to transform the edge weights to increase interpretability. Logging the values is a common move. The `netify` plot function has a built-in parameter for this — pass any function to `mutate_weight` (see also the general `mutate_weights()` function for transforming weights outside of plotting). The `edge_alpha_label` parameter just relabels the resulting legend:

```{r, fig.width=6, fig.height=6}
set.seed(6886)
plot(mex_network,
    select_text = random_names,
    select_text_display = random_names,
    # log(x+1) to better see the range of connections
    mutate_weight = log1p,
    edge_alpha_label = "Log(Event Count)"
)
```

### highlighting a single actor

Often there is one actor you really want to call out — a focal group, a state agency, a single firm. `plot.netify()` supports this directly via `highlight=`, which takes either a character vector of focal actors or a named color vector mapping actor names to colors:

```{r, fig.width=6, fig.height=6}
focal <- "Government of Mexico"
set.seed(6886)
plot(mex_network,
    mutate_weight = log1p,
    edge_alpha_label = "Log(Event Count)",
    highlight = focal,
    highlight_color = c("Government of Mexico" = "#0A3161", "Other" = "grey70"),
    select_text = focal,
    select_text_display = focal
)
```

The focal actor is now visually distinct from the rest of the network — useful for storytelling and for slide decks.

### applying a theme

Drop on `theme_publication_netify()` — a ggplot theme with a larger base font, italic legend titles, and stripped axes (the right default for force-directed layouts):

```{r, fig.width=6, fig.height=6}
# add degree centrality from summary_actor()
actor_stats <- summary_actor(mex_network)
actor_stats$degree_group <- ifelse(
    actor_stats$degree > median(actor_stats$degree, na.rm = TRUE),
    "higher degree",
    "lower degree"
)
mex_network <- add_node_vars(
    mex_network,
    actor_stats,
    actor = "actor"
)

set.seed(6886)
plot(mex_network,
    mutate_weight = log1p,
    edge_alpha_label = "Log(Event Count)",
    node_size_by = "degree",
    node_size_label = "Degree",
    node_alpha = .5,
    check_overlap = TRUE
) +
    theme_publication_netify()
```

`theme_publication_netify()` is built for network layouts (no axes). For actor-level summaries or heatmaps — where you *do* want axes — use `theme_publication_netify_ts()` (the `_ts` is for "tables / time-series", anything that isn't a node-edge layout).

### saving the figure

Once you have a plot you like, export it with `ggsave()`. A vector format (PDF or SVG) holds up at any size; for the web or social, a PNG at 300 dpi is fine.

```{r, eval=FALSE}
# the last plot rendered is captured by ggsave() by default
ggsave("mexico_conflict.pdf", width = 7, height = 6)

# pass a named plot object
p <- plot(mex_network, mutate_weight = log1p) + theme_publication_netify()
ggsave("mexico_conflict.png", plot = p, width = 7, height = 6, dpi = 300)
```

### a note on color choice

Colors matter for accessibility — roughly 8% of men and 0.5% of women have some form of color vision deficiency. When you encode a variable in color, prefer a colorblind-safe palette. Two easy options:

- **ColorBrewer qualitative palettes** like `Set2`, `Dark2`, and `Paired` are colorblind-safe by design. Use `scale_color_brewer(palette = "Set2")` or just pass `node_color_palette = "Set2"` to `plot.netify()`.
- **viridis** (`viridis`, `magma`, `plasma`, `cividis`) gives a perceptually uniform, colorblind-safe gradient for continuous variables.

```{r, eval=FALSE}
# example: colorblind-safe categorical fill via plot.netify()
plot(mex_network,
    node_color_by = "degree_group",
    node_color_palette = "Set2"     # colorbrewer colorblind-safe
) +
    theme_publication_netify()

# example: colorblind-safe continuous color, manual
plot(mex_network, node_color_by = "degree") +
    scale_color_viridis_c(option = "cividis") +
    theme_publication_netify()
```

### advanced: assemble the plot from components

For most uses the calls above are enough. If you want full control over how the layers are stacked — say, to insert a custom annotation between the edge layer and the node layer — set `return_components = TRUE` and reassemble the pieces yourself.

```{r, fig.width=6, fig.height=6}
# investigate each component of the plot
set.seed(6886)
comp <- plot(
    mex_network,
    remove_isolates = TRUE,
    select_text = random_names,
    select_text_display = random_names,
    mutate_weight = log1p,
    return_components = TRUE
)

comp

# modify component
comp$base +
    netify_edge(comp) +
    labs(alpha = "Log(Event Count)") +
    reset_scales() +
    netify_node(comp) +
    netify_text_repel(comp) +
    comp$theme
```

### a note on missingness

When you build a network from event data, `netify()` by default fills unobserved dyads with 0 (`missing_to_zero = TRUE`). For some applications — contact tracing, animal interactions, partially-observed survey rosters — the difference between *observed-no-interaction* (0) and *not-observed-at-all* (NA) matters. Set `missing_to_zero = FALSE` to keep NAs as NAs. Similarly, `actor_time_uniform = FALSE` lets actor composition vary across periods (useful for open-cohort longitudinal data); the netlet then stores an `actor_pds` data.frame recording each actor's entry / exit window. See `?netify` for details.

## two-mode (bipartite) event data: offshore jurisdictions

Event data are not limited to one-mode actor-on-actor interactions. People attending venues, students enrolling in classes, firms registered in jurisdictions — anywhere two distinct sets of entities are linked by observed events — calls for a **bipartite** network.

A concrete data-journalism example: leaks like the Panama Papers or Paradise Papers connect *entities* (companies, trusts) to *jurisdictions* (tax havens). Each entity-jurisdiction pairing is an event ("firm X was incorporated in haven Y"). The result is a two-mode network with firms on one side and jurisdictions on the other.

```{r}
# toy offshore-leak data: firms incorporated in jurisdictions
set.seed(6886)
firms <- paste0("firm_", LETTERS[1:8])
havens <- c("BVI", "Cayman", "Bermuda", "Jersey", "Panama", "Mauritius")

offshore <- data.frame(
    firm = sample(firms, 40, replace = TRUE),
    haven = sample(havens, 40, replace = TRUE,
        prob = c(0.30, 0.25, 0.15, 0.12, 0.10, 0.08))
)
head(offshore)
```

To build the bipartite netlet, set `mode = "bipartite"` and point `actor1` / `actor2` at the two entity columns. `sum_dyads = TRUE` aggregates repeated firm-haven incorporations into an edge weight.

```{r}
bp <- netify(
    offshore,
    actor1 = "firm", actor2 = "haven",
    mode = "bipartite",
    sum_dyads = TRUE
)
bp
```

Because the row actors (firms) and column actors (havens) are different sets, the resulting adjacency is rectangular and asymmetric. A **heatmap** is a natural display here -- one cell per firm-haven pair, color encoding the number of incorporations.

Use `plot(bp, style = "heatmap")` for a quick plot:

```{r, fig.width=6, fig.height=4}
plot(bp, style = "heatmap")
```

The code below extracts the same adjacency with `get_raw()` and draws the tile plot directly with `ggplot2::geom_tile()`:

```{r, fig.width=6, fig.height=4}
# turn a netify bipartite adjacency into a tidy long data.frame
heatmap_long <- function(net) {
    m <- get_raw(net)
    data.frame(
        row = factor(rep(rownames(m), times = ncol(m)),
            levels = rownames(m)),
        col = factor(rep(colnames(m), each = nrow(m)),
            levels = colnames(m)),
        value = as.numeric(m)
    )
}

bp_long <- heatmap_long(bp)

ggplot(bp_long, aes(x = col, y = row, fill = value)) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_viridis_c(option = "cividis", name = "Incorporations") +
    labs(x = "Jurisdiction", y = "Firm",
         title = "Offshore footprint by jurisdiction") +
    theme_publication_netify_ts(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Note the use of `theme_publication_netify_ts()` here instead of `theme_publication_netify()` — heatmaps need their axes, so the `_ts` variant (which keeps axes on) is the right choice. The `cividis` viridis palette is colorblind-safe and prints well in grayscale.

Bipartite networks use a different family of summaries than unipartite ones: there is no transitivity or mutuality (a triangle would have to span the bipartition), and degree / centralization are reported separately for row and column actors. Run `summary(bp)` for the full set.

```{r}
summary(bp)
```

To save the heatmap for a story or a slide deck:

```{r, eval=FALSE}
ggsave("offshore_heatmap.pdf", width = 7, height = 5)
```

## cookbook: twitter `@mentions`

Event data is not just for conflict. Anything that arrives as rows of `(source, target, time)` -- a tweet that mentions another handle, an email between coworkers, a code review that requests a reviewer -- can be netified the exact same way. Here is a tiny self-contained Twitter `@mentions` example so you can copy/paste and adapt.

```{r}
# tiny synthetic edge list: who mentioned whom, on which day
mentions <- data.frame(
    from = c("@alice", "@alice", "@bob",  "@bob",  "@carol",
             "@dave",  "@alice", "@eve",  "@eve",  "@frank"),
    to   = c("@bob",   "@carol", "@alice","@dave", "@dave",
             "@carol", "@eve",   "@bob",  "@alice","@alice"),
    day  = c(1, 1, 1, 2, 2, 2, 3, 3, 3, 3)
)

# directed (a mention has a sender and receiver); sum repeat mentions as
# the edge weight by letting netify count them for us
mention_net <- netify(
    mentions,
    actor1 = "from", actor2 = "to",
    time = "day",
    symmetric = FALSE,
    sum_dyads = TRUE
)

mention_net
```

A quick `plot()` already gives you the picture:

```{r, fig.width=6, fig.height=5}
plot(mention_net)
```

And the usual `summary*` helpers tell you who is most active and who is most mentioned:

```{r}
summary(mention_net)

# directed networks split degree into in / out / total
mention_actor <- summary_actor(mention_net)
head(mention_actor[, c("actor", "time", "degree_in", "degree_out", "degree_total")])
```

Same recipe, different domain -- the only thing that changes is the column names you hand to `netify()`.

## references

- Sundberg, Ralph and Erik Melander (2013) Introducing the UCDP Georeferenced Event Dataset. *Journal of Peace Research* 50(4).

- Boschee, E., Lautenschlager, J., O'Brien, S., Shellman, S., Starz, J., & Ward, M. (2015). ICEWS Coded Event Data. Harvard Dataverse. DOI 10.7910/DVN/28075.
