---
title: "Foundations"
description: "Build a netify object from dyadic political-science data."
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Foundations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    dev = "png", dpi = 150,
    cache = FALSE,
    echo = TRUE
)

```

## package overview

You supply dyadic data, and `netify` turns it into network objects that work with summary, plotting, and modeling tools. The package is built for social-science network data, especially peace science data, but the same workflow works for many dyadic data sets.

This vignette walks through the main workflow: create a network, inspect it, plot it, add attributes, and pass it to other network packages.

The main workflow has three parts:

  (1) **Create**: Build network objects from dyadic data and attach nodal or dyadic variables.
  (2) **Explore**: Summarize networks at the graph and actor levels, then visualize the results.
  (3) **Model**: Prepare the network for other R network packages.

The table below lists common functions for each part of the workflow.

| Create | Explore | Model |
|---------|---------|---------|
| `netify()`  | `peek()`  | `netify_to_amen()`  |
| `add_node_vars()` | `summary_actor()`   | `netify_to_igraph()`  |
| `add_dyad_vars()`  | `summary()`  | `netify_to_statnet()`  |
| `subset_netify()`  | `plot_actor_stats()`  |   | 
|              |   `plot_graph_stats()`|        |
|              |   `plot()`|        |

`netify` begins with the user's data input. Our core function, `netify()`, handles several different data inputs including data frames and edgelists (see documentation for more detail).

- The package can also create different types of networks including:

  - cross sectional networks 
  - longitudinal (with static and varying actor composition)
  - bipartite networks
  - multilayer

- As well as create networks with different edge types:
  - weighted 
  - binary 
  - symmetric or non-symmetric 

## step 1: create

Begin by loading packages and supplying the data. We will use the `peacesciencer` package to grab some familiar data.


```{r, eval=TRUE, results='hide', message=FALSE}
# load packages
library(netify)
library(peacesciencer)
library(dplyr)
library(ggplot2)

# organize external data for peacesciencer
peacesciencer::download_extdata()

# build a dyadic panel with peacesciencer
cow_dyads <- create_dyadyears(
    subset_years = c(1995:2014)
) |>
    # add mid data
    add_cow_mids() |>
    # add capital distance
    add_capital_distance() |>
    # add democracy
    add_democracy() |>
    # add gdp
    add_sim_gdp_pop(keep = c("pwtrgdp", "pwtpop"))
```

Next, create a `netlet` object from the COW data frame with `netify()`. The main arguments are:

- `input` is, in this use case, a dyadic data.frame that should have at the following variables used to specify actors:

  - `actor1`: names the first actor column
  - `actor2`: names the second actor column
  
- `time` names the period variable for longitudinal data.
- `output_format` sets the storage shape when you need one explicitly (`"cross_sec"`, `"longit_array"`, or `"longit_list"`); otherwise `netify()` picks from the input structure.

```{r, eval=TRUE}
mid_long_network <- netify(
    input = cow_dyads,
    actor1 = "ccode1", actor2 = "ccode2", time = "year",
    weight = "cowmidonset",
    actor_time_uniform = FALSE,
    sum_dyads = FALSE, symmetric = TRUE,
    diag_to_NA = TRUE, missing_to_zero = TRUE,
    nodal_vars = c("v2x_polyarchy1", "v2x_polyarchy2"),
    dyad_vars = c("capdist"),
    dyad_vars_symmetric = c(TRUE)
)
```

You have created a network object.

You can also add nodal and dyadic data after creating the network with `add_node_vars()` and `add_dyad_vars()`.

For example, add logged GDP for each actor-year:

```{r, eval=TRUE}
# build one actor-year row for each node
node_data <- bind_rows(
    cow_dyads |>
        select(actor = ccode1, year, pwtrgdp = pwtrgdp1),
    cow_dyads |>
        select(actor = ccode2, year, pwtrgdp = pwtrgdp2)
) |>
    distinct(actor, year, .keep_all = TRUE) |>
    mutate(pwtrgdp_log = log(pwtrgdp + 1))

# attach the nodal variable
mid_long_network <- add_node_vars(
    netlet = mid_long_network,
    node_data = node_data,
    actor = "actor",
    time = "year"
)

# build a dyadic distance variable
cow_dyads$log_capdist <- log(cow_dyads$capdist + 1)

# attach the dyadic variable
mid_long_network <- add_dyad_vars(
    netlet = mid_long_network,
    dyad_data = cow_dyads,
    actor1 = "ccode1",
    actor2 = "ccode2",
    time = "year",
    dyad_vars = "log_capdist",
    dyad_vars_symmetric = TRUE
)
```

## step 2: explore and summarize

We made a network, so let's look at it. First, we might want to take a `peek` at the network object to see if the matrix looks the way we'd expect it to look. This function lets you glance at a specific slice of the network if it is longitudinal or the entire network if it is cross-sectional. (To actually subset the netlet object and make a new object use netify's `subset` function.)

```{r}
peek(mid_long_network,
    from = 5, to = 5,
    time = c("2009", "2010")
)
```

Next, let's examine a few basic summary statistics about the network using our`summary()` function.

```{r}
# build network-level summary statistics
mid_long_summary <- summary(mid_long_network)
```

We can also make a quick visualization of network statistics over time using the summary statistics data frame.

```{r, fig.width=6, fig.height=6}
plot_graph_stats(mid_long_summary)
```

These graph statistics are useful for understanding changes over time at the network level. We might also want to look at actor-level statistics over time. The built-in `summary_actor()` function calculates actor-level degree, strength, and centrality statistics for each actor in each time period. The `plot_actor_stats()` function can show distributions across actors or trajectories for selected actors:

```{r}
# summarize every actor-year
summary_actor_mids <- summary_actor(mid_long_network)
head(summary_actor_mids)
```

We can look at the distribution of the statistic for all actors over time:

```{r, fig.width=6, fig.height=6, message = FALSE}
# plot actor-level distributions
plot_actor_stats(
    summary_actor_mids,
    across_actor = TRUE
)
```

Or we might like to select a specific statistic to focus on across actors over time:

```{r, fig.width=6, fig.height=6}
# focus on closeness
plot_actor_stats(
    summary_actor_mids,
    across_actor = TRUE,
    specific_stats = "closeness"
)
```

For actor-specific statistics over time, subset to a few actors so the plot stays legible.

```{r, fig.width=6, fig.height=6}
# selected high-gdp countries
top_5 <- c("2", "710", "740", "255", "750")

plot_actor_stats(
    summary_actor_mids,
    across_actor = FALSE,
    specific_actors = top_5
)
```

We can also zoom into a specific time slice of the network:

```{r, fig.width=6, fig.height=6}
summary_df_static <- summary_actor_mids[summary_actor_mids$time == 2011, ]

plot_actor_stats(
    summary_df_static,
    across_actor = FALSE,
    specific_actors = top_5
)
```

Instead of looking at summary statistics, we also might want to simply visualize the entire network. We can do this by plotting the netify object.

By default, `plot()` uses `auto_format = TRUE`, which automatically adjusts aesthetics based on your network's properties — node sizes scale down for larger networks, edge transparency increases for denser networks, and text labels appear for small networks (≤15 nodes). You can disable this with `auto_format = FALSE` for full manual control, or simply override any individual parameter.

```{r, fig.width=6, fig.height=6}
# default plot with auto_format
plot(mid_long_network,
    static_actor_positions = TRUE,
    remove_isolates = FALSE
)

# override specific defaults
plot(
    mid_long_network,
    edge_color = "grey",
    node_size = 2
)
```

You can also map actor-level summary statistics onto the network plot.

```{r}
# attach actor statistics
mid_long_network <- add_node_vars(
    mid_long_network,
    summary_actor_mids,
    actor = "actor", time = "time",
    node_vars = c("degree", "prop_ties", "eigen_vector"),
)
```

Print the object to inspect its metadata.


```{r}
# inspect the netlet object
print(mid_long_network)
```


The nodal attributes are stored on the object.

```{r}
# inspect nodal data
head(
    attr(
        mid_long_network,
        "nodal_data"
    )
)

head(attributes(mid_long_network)$nodal_data)
```

And now return to our network graph by highlighting specific nodal attributes. To map visual properties to data, we use the `node_*_by` naming convention (e.g., `node_size_by`, `node_color_by`). The legacy `point_*_var` names (e.g., `point_size_var`) also work identically if you prefer that style.

```{r, fig.width=6, fig.height=6}
# vary node size by degree
plot(
    mid_long_network,
    edge_color = "grey",
    node_size_by = "degree"
)

# vary node color by polyarchy
plot(
    mid_long_network,
    edge_color = "grey",
    node_size_by = "degree",
    node_color_by = "v2x_polyarchy1",
    node_color_label = "Polyarchy",
    node_size_label = "Degree"
) +
    scale_color_gradient2()
```

We might also prefer to add labels, but only a select few:

```{r, fig.width=6, fig.height=6}
library(countrycode)
cowns <- countrycode(
    c(
        "United States", "China", "Russia",
        "France", "Germany", "United Kingdom"
    ),
    "country.name", "cown"
)
cabbs <- countrycode(cowns, "cown", "iso3c")

plot(
    mid_long_network,
    edge_color = "grey",
    node_size_by = "degree",
    node_color_by = "v2x_polyarchy1",
    node_color_label = "Polyarchy",
    node_size_label = "Degree",
    select_text = cowns,
    select_text_display = cabbs,
    text_size = 3
) +
    scale_color_gradient2()

# draw labels without points
plot(
    mid_long_network,
    edge_color = "grey",
    add_points = FALSE,
    add_label = TRUE,
    label_size_var = "degree",
    label_color = "white",
    label_fill_var = "v2x_polyarchy1",
    label_fill_label = "Polyarchy",
    label_size_label = "Degree"
) + guides(size = "none")
```

### extracting data back to a data frame

If you need to convert a netify object back into a dyadic data frame — for example, to run regressions or export to other software — use `unnetify()`:

```{r}
mid_long_df <- unnetify(mid_long_network)
head(mid_long_df[, 1:5])
```

This returns one row per dyad with all nodal and dyadic attributes merged in. Use `remove_zeros = TRUE` to keep only non-zero edges for a more compact result.

## step 3: model

After creating and exploring the network, you can pass it to other modeling packages. The examples below use a cross-sectional network.

First, prepare the data:

```{r, message = FALSE}
# prepare cross-sectional data
cow_cross <- cow_dyads |>
    group_by(ccode1, ccode2) |>
    summarize(
        cowmidonset = ifelse(any(cowmidonset > 0), 1, 0),
        capdist = mean(capdist),
        polity21 = mean(polity21, na.rm = TRUE),
        polity22 = mean(polity22, na.rm = TRUE),
        pwtrgdp1 = mean(pwtrgdp1, na.rm = TRUE),
        pwtrgdp2 = mean(pwtrgdp2, na.rm = TRUE),
        pwtpop1 = mean(pwtpop1, na.rm = TRUE),
        pwtpop2 = mean(pwtpop2, na.rm = TRUE)
    ) |>
    ungroup() |>
    mutate(
        capdist = log(capdist + 1)
    )

# subset to actors with average population above 10 million
actor_to_keep <- cow_cross |>
    select(ccode1, pwtpop1) |>
    filter(pwtpop1 > 10) |>
    distinct(ccode1)

# filter cow_cross by actor_to_keep
cow_cross <- cow_cross |>
    filter(ccode1 %in% actor_to_keep$ccode1) |>
    filter(ccode2 %in% actor_to_keep$ccode1)

# build cross-sectional netlet
mid_cross_network <- netify(
    cow_cross,
    actor1 = "ccode1", actor2 = "ccode2",
    weight = "cowmidonset",
    sum_dyads = FALSE, symmetric = TRUE,
    diag_to_NA = TRUE, missing_to_zero = FALSE,
    nodal_vars = c(
        "polity21", "polity22", "pwtrgdp1",
        "pwtrgdp2", "pwtpop1", "pwtpop2"
    ),
    dyad_vars = c("capdist"),
    dyad_vars_symmetric = c(TRUE)
)
```

Next, pass the netify object to `amen`.
This section requires the `amen` package (`install.packages("amen")`).

```{r}
library(amen)

# prepare amen inputs
mid_cross_amen <- netify_to_amen(mid_cross_network)

# inspect amen inputs
str(mid_cross_amen)

# fit amen model
mid_amen_mod <- ame(
    Y = mid_cross_amen$Y,
    Xdyad = mid_cross_amen$Xdyad,
    family = "bin",
    R = 0,
    symmetric = TRUE,
    seed = 6886,
    nscan = 50,
    burn = 10,
    odens = 1,
    plot = FALSE,
    print = FALSE
)
```

We can apply the same process to ERGMs.
This section requires the `ergm` package (`install.packages("ergm")`).

```{r}
library(ergm)

# netify_to_statnet converts to a network object,
# which is what ergm uses
mid_cross_ergm <- netify_to_statnet(mid_cross_network)

# inspect statnet attributes
mid_cross_ergm

# replace missing nodecov values for the example
set.vertex.attribute(
    mid_cross_ergm, "polity21", 
    ifelse(
        is.na(get.vertex.attribute(mid_cross_ergm, "polity21")), 
        0, get.vertex.attribute(mid_cross_ergm, "polity21")))

set.vertex.attribute(
    mid_cross_ergm, "pwtrgdp2",
    ifelse(
        is.na(get.vertex.attribute(mid_cross_ergm, "pwtrgdp2")),
        0, get.vertex.attribute(mid_cross_ergm, "pwtrgdp2")) )

set.vertex.attribute(
    mid_cross_ergm, "pwtpop2",
    ifelse(
        is.na(get.vertex.attribute(mid_cross_ergm, "pwtpop2")),
        0,  get.vertex.attribute(mid_cross_ergm, "pwtpop2")) )

# fit ergm model
ergm_model <- ergm(
    formula = mid_cross_ergm ~
        edges +
        nodecov("polity21") +
        nodecov("pwtrgdp2") +
        nodecov("pwtpop2")
)
```



## references

- Csárdi G, Nepusz T, Traag V, Horvát S, Zanini F, Noom D, Müller K (2024). igraph: Network Analysis and Visualization in R. doi:10.5281/zenodo.7682609, R package version 2.0.3, https://CRAN.R-project.org/package=igraph.

- Davies, Shawn, Therese Pettersson & Magnus Öberg (2023). Organized violence 1989-2022 and the return of conflicts between states?. Journal of Peace Research 60(4).

- Handcock M, Hunter D, Butts C, Goodreau S, Krivitsky P, Morris M (2018). ergm: Fit, Simulate and Diagnose Exponential-Family Models for Networks. The Statnet Project (https://statnet.org/). R package version 3.9.4, https://CRAN.R-project.org/package=ergm.

- Hoff, Peter D. "Dyadic data analysis with amen." arXiv preprint arXiv:1506.08237 (2015).

- Högbladh Stina, 2023, “UCDP GED Codebook version 23.1”, Department of Peace and Conflict Research, Uppsala University

- Miller S (2022). “peacesciencer: An R Package for Quantitative Peace Science Research.” Conflict Management and Peace Science, 39(6), 755–779. doi: 10.1177/07388942221077926.

- Statnet Development Team (Pavel N. Krivitsky, Mark S. Handcock, David R. Hunter, Carter T. Butts, Chad Klumb, Steven M. Goodreau, and Martina Morris) (2003-2023). statnet: Software tools for the Statistical Modeling of Network Data. URL https://statnet.org/

- Sundberg, Ralph and Erik Melander (2013) Introducing the UCDP Georeferenced Event Dataset. Journal of Peace Research 50(4).
