---
title: "Foundations (lite)"
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Foundations (lite)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    dev = "png", dpi = 150,
    cache = FALSE,
    echo = TRUE,
    collapse = TRUE,
    comment = "#>"
)
```

This is a minimal end-to-end tour of `netify` that uses **only the data bundled with the package** — no `peacesciencer`, no `countrycode`, no external downloads. If you want the full IR-data walkthrough with covariates from outside sources, see the **Foundations** vignette.

We'll cover the four things `netify` is for:

1. **Build** a network object from dyadic data 🛠
2. **Explore** it with summary statistics and a quick plot 🔎
3. **Test** a couple of basic inferential questions 📊
4. **Bridge** out to other packages when you want to model 🔁

```{r}
library(netify)
library(ggplot2)
```

## 1. Build 🛠

The bundled `icews` dataset has ICEWS event-data slices for 152 countries from 2002 to 2014 with the four "quad" variables (verbal/material × cooperation/conflict) plus a handful of nodal covariates.

```{r}
data(icews)
head(icews[, c("i", "j", "year", "verbCoop", "matlConf", "i_polity2", "i_region")])
```

Turn it into a netify object with one call:

```{r}
verb_coop <- netify(
    icews,
    actor1 = "i", actor2 = "j", time = "year",
    symmetric = FALSE,
    weight   = "verbCoop",
    nodal_vars = c("i_polity2", "i_log_gdp", "i_region"),
    dyad_vars  = c("matlCoop", "verbConf")
)

print(verb_coop)
```

The `print()` summary tells you the network is unipartite, directed, longitudinal (13 periods), 152 actors, with nodal and dyadic features attached. The numeric table shows density, missingness, mean edge weight, mutual-dyad proportion, and transitivity averaged across time.

A quick glossary, since several of these terms appear before they're defined elsewhere:

- **unipartite**: one kind of actor (countries here). The opposite is **bipartite** -- two kinds, e.g., students-to-clubs.
- **directed** / **symmetric**: a directed tie has a sender and a receiver (`i -> j` is different from `j -> i`); a symmetric tie does not.
- **density**: share of possible ties that are actually present (0 = no ties, 1 = complete graph).
- **transitivity**: probability that two of your friends are also friends with each other -- "friend of a friend is a friend." Higher = more clustering.
- **mutual-dyad proportion (`mutual`)**: among ordered pairs where at least one tie exists, the share where both `i -> j` and `j -> i` are present.
- **reciprocity**: correlation between the adjacency matrix `A` and its transpose `A^T`; behaves like `mutual` for binary networks but generalizes to weighted ones.

## 2. Explore 🔎

`summary()` returns one row per time period with graph-level statistics:

```{r}
gs <- summary(verb_coop)
head(gs[, c("net", "num_actors", "density", "reciprocity", "mutual", "transitivity")])
```

Note both `reciprocity` (the correlation between $A$ and $A^T$, useful for weighted networks) and `mutual` (the classic mutual-dyad proportion). Use whichever fits your audience.

Actor-level stats — degree, prop_ties, centrality, strength — come from `summary_actor()`. Quick definitions of what shows up in the columns:

- **degree_in / degree_out / degree_total**: count of incoming, outgoing, or total ties for an actor.
- **prop_ties_***: those degree counts divided by the number of possible partners.
- **strength**: sum / average / sd of the *weights* of an actor's ties (parallel to degree but weight-aware).
- **betweenness**: how often an actor sits on the shortest path between two others -- a broker score.
- **closeness**: how short the actor's average distance is to everyone else.
- **authority_score / hub_score** (HITS): authorities are nodes that many hubs point to; hubs are nodes that point to many authorities. Useful in directed networks (e.g., who's getting cited vs. who's doing the citing).

```{r}
as_ <- summary_actor(verb_coop)
head(as_[, c("actor", "time", "degree_in", "degree_out",
             "betweenness", "authority_score", "hub_score")])
```

Plot it. The default uses `auto_format = TRUE` and adapts to network size — for 152 actors over 13 years it'll suppress text labels and tone down edge alpha automatically:

```{r, fig.width = 7, fig.height = 5}
plot(verb_coop,
     time_filter = c("2004", "2008", "2012"),
     node_color_by = "i_region",
     edge_alpha = 0.1) +
  theme(legend.position = "bottom")
```

## 3. Test (basic inferential) 📊

A common first question is *homophily*: do similar countries cooperate more? Test it:

```{r}
hom <- homophily(verb_coop,
                 attribute = "i_polity2",
                 method = "correlation",
                 significance_test = TRUE)
head(hom)
```

For a categorical attribute, `mixing_matrix()` gives the full who-with-whom table:

```{r}
mm <- mixing_matrix(verb_coop, attribute = "i_region", normalized = TRUE)
round(mm$mixing_matrices[[1]], 2)
mm$summary_stats[1, ]
```

`compare_networks()` is the all-purpose comparison tool. For a longitudinal netify it returns pairwise temporal comparisons:

```{r}
temp_cmp <- compare_networks(verb_coop, method = "correlation")
head(temp_cmp$summary)
```

You can also slice by a nodal attribute and compare the resulting subnetworks:

```{r}
by_region <- compare_networks(
  subset(verb_coop, time = "2010"),
  by = "i_region",
  method = "correlation"
)
by_region$by_group$n_actors_per_group
```

## 4. Bridge 🔁

netify intentionally stops at descriptives + basic inference. For statistical models, hand off to a downstream package:

| Want to fit... | Use this | netify exporter |
|---|---|---|
| Latent factor / AME | [amen](https://github.com/pdhoff/amen) | `to_amen(netlet)` |
| Longitudinal AME (varying actors) | [netify-dev/lame](https://github.com/netify-dev/lame) | `to_amen(netlet, lame = TRUE)` |
| Multilayer dynamic bilinear | [netify-dev/dbn](https://github.com/netify-dev/dbn) | `to_dbn(netlet)` |
| Social relations model | [netify-dev/srm](https://github.com/netify-dev/srm) | `to_amen(netlet)` |
| ERGM | [ergm/statnet](https://statnet.org/) | `to_statnet(netlet)` |
| Community detection / graph algorithms | [igraph](https://igraph.org/r/) | `to_igraph(netlet)` |
| Roll your own dyadic regression | — | `unnetify(netlet)` for a long data frame |

Example: convert to igraph and use a function netify doesn't expose:

```{r, eval = requireNamespace("igraph", quietly = TRUE)}
ig <- to_igraph(verb_coop)             # list of igraph objects, one per year
ig_2010 <- ig[["2010"]]
length(igraph::cluster_walktrap(ig_2010))
```

Or flatten back to a dyadic data frame:

```{r}
df <- unnetify(subset(verb_coop, time = "2010"), remove_zeros = TRUE)
head(df[, c("from", "to", "verbCoop", "matlCoop", "i_polity2_from", "i_polity2_to")])
```

## tl;dr

```r
net <- netify(df, actor1 = "i", actor2 = "j", time = "year", weight = "x")
summary(net)            # graph-level stats
summary_actor(net)      # actor-level stats
plot(net)               # ggplot-based network visual
homophily(net, attribute = "v")   # do similar actors connect?
compare_networks(net)             # how do periods/layers/groups differ?
to_amen(net)  # or to_dbn / to_statnet / to_igraph when you're ready to model
```

For the long version with peacesciencer data and a full IR walkthrough, see `vignette("foundations", package = "netify")`. Have fun!

## A note for non-time use cases

Most of netify's docs talk about "longitudinal" networks because that's the bread-and-butter use case in social science. But the underlying `longit_list` structure (a list of matrices, each with its own actor set) works for **any per-slice partition**, not just time. Common non-time examples:

- **Per-subject networks**: brain connectivity matrices, one per fMRI subject
- **Per-replicate networks**: synthetic networks drawn from a generative process
- **Per-condition networks**: networks observed under different experimental conditions
- **Per-document networks**: term co-occurrence networks, one per document

For these, build a list of matrices (one per partition) and use either `new_netify(list_of_matrices)` or `netify(df, ..., time = "partition_id")`. All of netify's per-period operations — `summary()`, `summary_actor()`, faceted `plot()`, `subset(time = ...)` — work identically; the variable name just happens to be `time`. If the "time" framing gets confusing in your context, alias it locally:

```r
# build a per-subject network library
subjects <- new_netify(list_of_subject_matrices)
# treat the "time" dimension as subject
per_subject_stats <- summary(subjects)
```

This generality is why the package treats `longit_list` as a general partition format rather than a strictly temporal one.
