---
title: "Netify: Manual Plotting with ggplot2"
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Netify: Manual Plotting with ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  dev = 'png', dpi = 150,
  cache = FALSE,
  echo = TRUE
)
```


This vignette provides an overview of how to create customizable plots using `ggplot2` while still using `netify` to prepare the data. 

Lets load the necessary libraries.

```{r}
library(netify)
library(ggplot2)
```

We'll also use the `ggnewscale` package to create multiple legends when necessary in the same plot (e.g., if you want to have legends for a color aesthetic for both nodes and edges).

```{r}
if(!'ggnewscale' %in% rownames(installed.packages())){
  install.packages('ggnewscale', repos='https://cloud.r-project.org') }
library(ggnewscale)
```

##  Preparing data

First lets create a `netlet` object from some dyadic data (ICEWS data) using the `netify` package.

```{r}
# load icews data
data(icews)

# choose attributes
nvars = c( 'i_polity2', 'i_log_gdp', 'i_log_pop' )
dvars = c( 'matlCoop', 'verbConf', 'matlConf' )

# create a netify object
netlet = netify(
    icews, 
    actor1='i', actor2='j',
    time = 'year',
    symmetric=FALSE, weight='verbCoop',
    mode='unipartite', sum_dyads=FALSE,
    actor_time_uniform=TRUE, actor_pds=NULL,
    diag_to_NA=TRUE, missing_to_zero=TRUE,
    nodal_vars = nvars, 
    dyad_vars = dvars
)

# subset to a few actors
actors_to_keep = c(
  'Australia', 'Brazil',
  'Canada', 'Chile', 'China',
  'Colombia', 'Egypt', 'Ethiopia',
  'France', 'Germany', 'Ghana', 
  'Hungary', 'India', 'Indonesia', 
  'Iran, Islamic Republic Of', 
  'Israel', 'Italy', 'Japan', 'Kenya', 
  "Korea, Democratic People's Republic Of",
  'Korea, Republic Of', 'Nigeria', 'Pakistan', 
  'Qatar', 'Russian Federation', 'Saudi Arabia',
  'South Africa', 'Spain', 'Sudan', 
  'Syrian Arab Republic', 'Thailand', 
  'United Kingdom', 'United States', 
  'Zimbabwe' )
netlet = subset_netify(
  netlet, 
  actors = actors_to_keep
)

# print
netlet
```

This is a longitudinal, weighted network with nodal and dyadic attributes. In a few more steps we will show how to highlight these attributes in the plot.

Next, we use the `net_plot_data` function to create a data frame for `ggplot2`. `net_plot_data` extracts and sets up node and edge data from a `netify` object according to specified plotting arguments. It returns a list of different components but the most important one for users is the `net_dfs` element. This element contains two objects: `edge_data` and `nodal_data`. These are data frames that can be passed to `ggplot2`.

```{r}
# create a data frame for plotting
plot_data = net_plot_data(netlet)

# get relevant dfs
net_dfs = plot_data$net_dfs

# check structure of what's here
str(net_dfs)

# check the first few rows of the edge data
head(net_dfs$edge_data)

# check the first few rows of the nodal data
head(net_dfs$nodal_data)
```

The `x` and `y` in `nodal_data` and the `x1`, `y1`, `x2`, and `y2` in `edge_data` are the coordinates of the nodes and edges, respectively. These are the coordinates that will be used to plot the network. 

## Creating a plot

Now that we have the data, we can create a plot using `ggplot2`. We'll use the `geom_segment` and `geom_point` (or, `geom_label`, `geom_text`, and the `ggrepel package equivalents) functions to plot the edges and nodes, respectively.

```{r, fig.width=9, fig.height=8}
ggplot() + 
	geom_segment(
		data = net_dfs$edge_data,
		aes(
			x=x1, 
			y=y1, 
			xend=x2, 
			yend=y2
		),
		color='lightgrey',
		alpha=.2
	) +
	geom_point(
		data = net_dfs$nodal_data,
		aes(
			x=x, 
			y=y,
			size=i_log_pop, 
			color=i_polity2
		)
	) +
  labs(
    color='Polity', 
    size='Log(Pop.)'
  ) +
	scale_color_gradient(low='#a6bddb', high='#014636') +
	facet_wrap(~time, scales='free') +
	theme_netify()
```

### Changing the layout

By default layouts for node positions are drawn from the `layout_nicely` algorithm in the `igraph` package. Users can specify other layouts as, for example, say that you wanted to use the `mds` algorithm instead:

```{r}
# create a df using mds instead 
plot_data_mds = net_plot_data(netlet, 
    list(
        layout='mds'
        )
    )

# see new x-y coordinates
lapply(plot_data_mds$net_dfs, head)
```


### Add Edge Information

So far, we have focused on using color to convey information about nodal attributes in the network (population size and polity score). Now, let's add more edge information to the plot. For example, we can include information about the `matlConf` dyadic attribute. Imagine we want to highlight edges of verbal cooperation that occur at the same time as when higher than average levels of material conflict occur in the network. First, let's create the variable in the edge data.

```{r}
if(!'dplyr' %in% rownames(installed.packages())){
  install.packages('dplyr', repos='https://cloud.r-project.org') }
library(dplyr)

# create high_matlConf variable
net_dfs$edge_data = net_dfs$edge_data |>
    group_by(time) |>
    mutate(
        high_matlConf = matlConf > mean(matlConf, na.rm=TRUE)
    ) |>
    ungroup() |>
    as.data.frame()

# check
head(net_dfs$edge_data)
```    

Now that we have the new variable in the data.frame, we can plot by it but note that we now need a color aesthetic for both points and segments, even though `ggplot2` only supports one legend by aesthetic by default. We can get around this by using the `new_scale_color` function from the ``ggnewscale` package.

```{r, fig.width=9, fig.height=8}
# color line segments by this new variable
ggplot() + 
	geom_segment(
		data = net_dfs$edge_data,
		aes(
			x=x1, 
			y=y1, 
			xend=x2, 
			yend=y2,
            color=high_matlConf
		),
		alpha=.2
	) +
    scale_color_manual(
      name='', 
      values=c('grey', 'red'),
      labels=c('Below Avg. Matl. Conf', 'Above Avg.')
      ) +
    new_scale_color() +
	geom_point(
		data = net_dfs$nodal_data,
		aes(
			x=x, 
			y=y,
			size=i_log_pop, 
			color=i_polity2
		)
	) +
	scale_color_gradient(
	  name='Polity',
	  low='#a6bddb', high='#014636') +
  labs(
    size='Log(Pop.)'
  ) +
	facet_wrap(~time, scales='free') +
	theme_netify() +
  theme(
    legend.position='right'
  )
```

## References

- Boschee, Elizabeth; Lautenschlager, Jennifer; O'Brien, Sean; Shellman, Steve; Starz, James; Ward, Michael, 2015, ``ICEWS Coded Event Data'', doi:10.7910/DVN/28075 , Harvard Dataverse.

- Pedersen, T. L. (2020). ggnewscale: Multiple Fill and Colour Scales in 'ggplot2'. R package version 0.4.3. https://CRAN.R-project.org/package=ggnewscale

- Wickham, H. (2016). ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York.