---
title: "Multilayer Networks"
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multilayer Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE,
    fig.width = 7,
    fig.height = 5
)
```

This vignette demonstrates how to analyze multilayer networks in `netify`. Multilayer networks capture multiple types of relationships between the same actors -- a crucial feature for understanding complex political systems where actors interact through various channels simultaneously.

## Why Multilayer Networks Matter for Social Science

Traditional network analysis forces us to choose: study trade OR conflict, cooperation OR competition, formal OR informal ties. But political reality is multilayered. States that trade may also fight. Legislators who cosponsor bills may attack each other on social media. Ignoring these multiple, simultaneous relationships leads to incomplete and potentially misleading conclusions.

Multilayer networks solve this problem by allowing us to analyze multiple relationship types together, revealing:

- **Substitution effects**: Do actors use one type of relationship instead of another?
- **Complementarities**: Do certain relationships tend to co-occur?
- **Cross-layer influence**: Does position in one network affect behavior in another?
- **Structural coupling**: How do different relationship types shape each other?

### Quick Start Example

We'll use data from the Integrated Crisis Early Warning System (ICEWS) throughout here.

```{r}
library(netify)
library(tidyverse)

# Load the ICEWS event data
data(icews)

# Create a simple multilayer network for 2010
icews_2010 <- icews[icews$year == 2010, ]

# Verbal cooperation network
verbal_net <- netify(
    icews_2010,
    actor1 = "i", actor2 = "j",
    weight = "verbCoop"
)

# Material cooperation network
material_net <- netify(
    icews_2010,
    actor1 = "i", actor2 = "j",
    weight = "matlCoop"
)

# Combine into multilayer network
multi_net <- layer_netify(
    list(verbal = verbal_net, material = material_net)
)

# Quick visualization
plot(multi_net, add_text = FALSE)
```

**What we see**: Even this simple example reveals the "cheap talk" problem in international relations. The verbal cooperation network (left) is dense with connections, while material cooperation (right) is sparse. Most diplomatic promises don't translate into material support.

## Building Multilayer Networks: A Substantive Example

Let's build a more complete multilayer network to study international cooperation patterns during a critical period -- 2002, shortly after 9/11 when cooperation dynamics were rapidly shifting.

```{r}
# Focus on 2002 - post-9/11 cooperation dynamics
icews_2002 <- icews[icews$year == 2002, ]

# Create networks with relevant attributes
verbal_coop_net <- netify(
    icews_2002,
    actor1 = "i", actor2 = "j",
    symmetric = TRUE,
    weight = "verbCoop",
    nodal_vars = c("i_polity2", "i_log_gdp", "i_log_pop", "i_region"),
    dyad_vars = c("verbConf", "matlConf")
)

material_coop_net <- netify(
    icews_2002,
    actor1 = "i", actor2 = "j",
    symmetric = TRUE,
    weight = "matlCoop",
    nodal_vars = c("i_polity2", "i_log_gdp", "i_log_pop", "i_region"),
    dyad_vars = c("verbConf", "matlConf")
)

# Combine into multilayer network
multilayer_net <- layer_netify(
    netlet_list = list(verbal_coop_net, material_coop_net),
    layer_labels = c("Verbal", "Material")
)

print(multilayer_net)
```

## Visualizing Multilayer Structures

### Basic Visualization with Interpretation

```{r, fig.width=10, fig.height=5}
plot(multilayer_net,
    add_text = FALSE,
    node_size_by = "i_log_gdp",
    node_color_by = "i_region"
) +
    theme(
        legend.position = "right"
    )
```

This visualization reveals the fundamental asymmetry in international cooperation. The verbal layer shows the "diplomatic commons" -- nearly all countries engage in diplomatic discourse. The material layer reveals the "cooperation core" -- a much smaller set of countries that back words with resources. Node sizes (GDP) show that while large economies participate in both networks, material cooperation isn't simply a function of capacity.

### Comparing Network Structures

```{r}
# Get structural comparison
struct_comp <- compare_networks(multilayer_net, what = "structure")
print(struct_comp)
```

**What these numbers mean**:

- **4x density difference**: For every material cooperation relationship, there are four verbal ones
- **3.6x degree difference**: The average country verbally cooperates with 123 others but materially supports only 34
- **Different clustering**: Higher transitivity in verbal cooperation (0.606 vs 0.311) suggests diplomatic communities form more readily than material support clusters

This structural evidence supports the "cheap talk" hypothesis but also reveals something deeper: material cooperation follows different organizational logics than diplomacy.

## Testing Political Science Theories with Multilayer Networks

### Theory 1: Democratic Peace in Different Domains

Do democracies cooperate more with each other? And does this vary by cooperation type?

```{r}
# Test homophily across layers
multilayer_homophily <- homophily(
    multilayer_net,
    attribute = "i_polity2",
    method = "correlation"
)

print(multilayer_homophily)
```

Democratic homophily exists in verbal cooperation (p < 0.001) but not material cooperation (p = 0.13). This suggests the "democratic peace" operates primarily through diplomatic channels. Material support crosses regime boundaries, likely driven by strategic rather than ideological considerations.

### Theory 2: Regional Security Complexes

Do security concerns create regional cooperation clusters? Let's examine regional patterns:

```{r}
# Regional mixing patterns
multilayer_mixing <- mixing_matrix(
    multilayer_net,
    attribute = "i_region"
)

print(multilayer_mixing$summary_stats)
```

Material cooperation shows stronger regional clustering (assortativity = 0.218) than verbal cooperation (0.156). This supports regional security complex theory -- when countries commit resources, they prioritize neighbors. Verbal cooperation, being cheaper, can afford to be more globally distributed.

### Theory 3: Conflict-Cooperation Dynamics

How does conflict in one domain relate to cooperation in another?

```{r}
# Analyze conflict-cooperation relationships
cross_layer_corr <- dyad_correlation(
    multilayer_net,
    dyad_vars = c("verbConf", "matlConf"),
    edge_vars = c("Verbal", "Material")
)

# Show key correlations
cross_layer_corr |>
    select(dyad_var, edge_var, correlation, p_value) |>
    filter(row_number() <= 4) # Show first 4 unique combinations
```

Verbal conflict strongly correlates with verbal cooperation (r = 0.456) -- countries that talk also argue. But material conflict shows weaker correlations with cooperation, suggesting material interactions follow different dynamics than diplomatic ones.

## Identifying Strategic Actors

### Who Bridges the Cooperation Gap?

Some countries specialize in material cooperation despite limited diplomatic engagement:

```{r}
# Find actor-level patterns
actor_stats <- summary_actor(multilayer_net)

# Identify material cooperation specialists
centrality_comparison <- actor_stats |>
    select(actor, layer, degree, betweenness, strength_sum) |>
    pivot_wider(names_from = layer, values_from = c(degree, betweenness, strength_sum))

material_specialists <- centrality_comparison |>
    mutate(
        material_bias = betweenness_Material / (betweenness_Verbal + 1)
    ) |>
    filter(betweenness_Material > quantile(betweenness_Material, 0.75, na.rm = TRUE)) |>
    arrange(desc(material_bias)) |>
    head(5)

print("Material cooperation brokers:")
print(material_specialists[, c("actor", "betweenness_Verbal", "betweenness_Material", "material_bias")])
```

Japan leads as a material cooperation broker despite low verbal centrality -- reflecting its post-WWII strategy of economic rather than diplomatic leadership. Countries in conflict or transition (Colombia, DRC, Ukraine) also appear as material brokers, likely as aid conduits.

### Power Dynamics Across Layers

```{r, fig.height=10}
# Compare major powers
plot_actor_stats(
    actor_stats,
    across_actor = FALSE,
    specific_actors = c("United States", "China", "Russian Federation", "Japan", "Germany")
)
```

**Power Analysis**: 

- The US dominates both layers, confirming hegemonic status
- Russia shows high verbal but lower material engagement -- influence through diplomacy rather than resources
- China (in 2002) shows moderate engagement, presaging its later rise
- Japan's material cooperation exceeds its diplomatic footprint
- Germany balances both dimensions, reflecting EU leadership

## Understanding Multiplexity: The Architecture of International Relations

Multiplexity analysis reveals which relationships involve multiple types of interaction -- a key indicator of relationship depth and stability:

```{r}
# Analyze edge patterns
verbal_melt <- melt(verbal_coop_net, remove_zeros = FALSE, na.rm = TRUE)
material_melt <- melt(material_coop_net, remove_zeros = FALSE, na.rm = TRUE)

edge_comparison <- merge(
    verbal_melt,
    material_melt,
    by = c("Var1", "Var2"),
    suffixes = c("_verbal", "_material")
) |>
    filter(value_verbal > 0 | value_material > 0)

# Categorize relationships
multiplexity_summary <- edge_comparison |>
    mutate(
        relationship_type = case_when(
            value_verbal > 0 & value_material > 0 ~ "Multiplex",
            value_verbal > 0 ~ "Verbal Only",
            value_material > 0 ~ "Material Only"
        )
    )

# Summary and visualization
table(multiplexity_summary$relationship_type)

ggplot(multiplexity_summary, aes(x = relationship_type)) +
    geom_bar() +
    labs(
        title = "The Multiplexity Gap in International Cooperation",
        subtitle = "Most relationships remain purely verbal",
        x = "Relationship Type",
        y = "Number of Country Pairs"
    ) +
    theme_minimal()
```

**Multiplexity Insights**:

- Only 26% of relationships are multiplex (involving both verbal and material cooperation)
- 124 "material only" relationships deserve attention -- these might represent covert cooperation or technical assistance below the diplomatic radar
- The 2,452 multiplex relationships form the stable core of international cooperation

### Who Has Multiplex Relationships?

```{r}
# Identify multiplex relationships
multiplex_pairs <- multiplexity_summary |>
    filter(relationship_type == "Multiplex") |>
    mutate(total_coop = value_verbal + value_material) |>
    arrange(desc(total_coop)) |>
    head(10)

print("Strongest multiplex relationships:")
print(multiplex_pairs[, c("Var1", "Var2", "value_verbal", "value_material")])
```

US-Russia tops the list despite tensions, showing how former adversaries maintain multiple channels. US relationships with allies (Israel, Japan) and strategic competitors (China) also show high multiplexity, indicating relationship importance transcends simple friend/foe categories.

## Advanced Analysis: Layer Dependencies

### How Do Layers Relate?

```{r}
# Comprehensive layer comparison
layer_comparison <- compare_networks(multilayer_net, what = "edges", method = "all")
print(layer_comparison)
```

**Key Metrics Explained**:

- **Correlation (0.472)**: Moderate positive relationship -- verbal cooperation somewhat predicts material cooperation
- **Jaccard (0.258)**: Only 26% edge overlap -- most relationships exist in just one layer
- **Hamming (0.305)**: About 30% of possible edges differ between layers
- **QAP test (p < 0.001)**: The correlation is statistically significant, not due to chance

### Visualizing Layer Relationships

```{r, fig.width=8, fig.height=6}
# Cross-layer scatter
ggplot(edge_comparison, aes(x = log1p(value_verbal), y = log1p(value_material))) +
    geom_point(alpha = 0.2, color = "darkblue") +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    geom_density_2d(color = "gray", alpha = 0.5) +
    labs(
        title = "The Verbal-Material Cooperation Gap",
        subtitle = "Most relationships cluster along axes, not diagonal",
        x = "Log(Verbal Cooperation + 1)",
        y = "Log(Material Cooperation + 1)"
    ) +
    theme_minimal() +
    annotate("text", x = 7, y = 1, label = "Talk without action", fontface = "italic") +
    annotate("text", x = 1, y = 5, label = "Quiet support", fontface = "italic") +
    annotate("text", x = 6, y = 5, label = "Full cooperation", fontface = "italic")
```

## Temporal Dynamics in Multilayer Networks

Understanding how different layers evolve over time reveals changing cooperation dynamics:

```{r, fig.width=12, fig.height=6}
# Create temporal multilayer network
years_example <- c(2001, 2002, 2003) # Around 9/11
icews_subset <- icews[icews$year %in% years_example, ]

verbal_temporal <- netify(
    icews_subset,
    actor1 = "i", actor2 = "j",
    time = "year",
    symmetric = TRUE,
    weight = "verbCoop"
)

material_temporal <- netify(
    icews_subset,
    actor1 = "i", actor2 = "j",
    time = "year",
    symmetric = TRUE,
    weight = "matlCoop"
)

temporal_multilayer <- layer_netify(
    list(verbal_temporal, material_temporal),
    layer_labels = c("Verbal", "Material")
)

# Visualize evolution
plot(temporal_multilayer,
    add_text = FALSE,
    facet_type = "wrap",
    facet_ncol = 3
)
```

The post-9/11 period shows network evolution. Both layers densify from 2001 to 2003, but material cooperation grows more selectively, suggesting crisis-driven coalition building.

## Practical Recommendations

### When to Use Multilayer vs. Separate Analysis

**Use multilayer when:**

- Relationships theoretically influence each other
- You suspect substitution or complementarity effects
- Actor positions might differ across relationship types
- You need to identify relationship depth through multiplexity

**Use separate analysis when:**

- Relationships are theoretically independent
- Different actor sets across layers
- Computational constraints exist
- Initial exploration before multilayer modeling

### Choosing Analysis Methods

1. **For regime type effects**: Use `homophily()` with "correlation" method for continuous variables like polity scores
2. **For regional/categorical patterns**: Use `mixing_matrix()` to see full interaction patterns
3. **For relationship strength**: Use `compare_networks()` with `what="edges"` and `method="all"` for comprehensive comparison
4. **For individual actors**: Use `summary_actor()` and examine layer-specific centralities

## tl;dr

Multilayer network analysis in `netify` reveals how different types of political relationships interact, overlap, and diverge. By moving beyond single-layer analysis, we can:

1. Test more nuanced theories about international behavior
2. Identify actors with specialized roles across different domains
3. Understand the architecture of complex political systems
4. Track how relationship portfolios evolve over time
5. Most importantly ... have fun!

