---
title: "Introduction to the netify Package"
author: "Ha Eun Choi, Cassy Dorff, Colin Henry, and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the netify Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

The goal of this vignette is to provide a wholistic overview of how to use the `netify` package.

`netify` is a set of network analysis tools. The package creates network objects, summarizes graphs, and analyzes these objects. `netify` also provides a bridge to other major network modeling packages so that users can seamlessly continue their work with available statistical modeling packages.

`netify` provides a suite of primary functions to help achieve these goals:

- `netify()`
- `add_dyad()`
- `add_nodal()`
- `peek()`
- `summary()`

In addition, `netify` includes additional functions that help users bridge between `netify` and other network packages:

- `prep_for_amen()`
- `prep_for_igraph()`
- `prep_for_network()`

### Install

In this document, we provide a brief tutorial on how to get started using the `netify` package. `netify` can be installed via devtools:

```{r, eval=FALSE}
library(devtools)
library(netify)
```

```{r, eval=TRUE, echo=FALSE}
suppressMessages(library(netify))
```

## Starting with dyadic data

The `netify` package allows users to begin with dyadic data wherein each row of the data represents an event or an exchange between actors, entities, or groups.  In dyadic data, each row typically contains information about which actors are involved in the event. We can transform this data into a matrix where each actor represents rows and columns and exchanges between the actors fill the values of the matrix.

To learn more about network analysis (with an emphasis on conflict networks) see below for suggested reading:

1. Cranmer, S. J., Desmarais, B. A., & Morgan, J. W. (2020). Inferential network analysis. Cambridge University Press.

3. Dorff, C., & Ward, M. D. (2013). Networks, dyads, and the social relations model. Political Science Research and Methods, 1(2), 159-178.

2. Dorff, C., Gallop, M., & Minhas, S. (2020). Networks of violence: Predicting conflict in Nigeria. The Journal of Politics, 82(2), 476-493.

3. Larson, J. M. (2021). Networks of conflict and cooperation. Annual Review of Political Science, 24, 89-107.

4. Minhas, S., Dorff, C., Gallop, M. B., Foster, M., Liu, H., Tellez, J., & Ward, M. D. (2022). Taking dyads seriously. Political Science Research and Methods, 10(4), 703-721.


### What data is in the netify package?

We include data examples drawn from the `peacesciencer` package and the ICEWS data. 

The *`peacesciencer`* package provides tools for constructing datasets for peace science scholarship. 

- See \url{https://svmiller.com/peacesciencer/ms.html} for more information. Users will need to install this package to run the examples below.

The *ICEWS* (Integrated Crisis Early Warning Systems) data example is a sample of interstate conflict and cooperation counts between 2002 to 2014. The dataset is for illustration purposes only. It is not to be used for research or inference. 

- The data is dyad-year data where the information is structured by yearly edgelists. 
- Boschee, Elizabeth; Lautenschlager, Jennifer; O'Brien, Sean; Shellman, Steve; Starz, James, 2018, "ICEWS Weekly Event Data", https://doi.org/10.7910/DVN/QI2T9A, Harvard Dataverse, V359.


To begin, let's create a dyadic dataset to work with. 

```{r, eval=TRUE, echo=FALSE}
library(peacesciencer)
library(tidyverse)

# create dyadic data set over time
cow_dyads <- create_dyadyears( 
     subset_years = c(1995:2014)
     ) %>%
    # add mids
    add_cow_mids() %>%
    # add capital distance
    add_capital_distance() %>%
    # add democracy
    add_democracy()
```

## Using `netify()`

### cross-sectional networks

Dyadic data like this represents exchanges between actors or entities and can be thought of as network data. To analyze dyadic data as a network, we can first transform dyadic data into an adjacency matrix using `netify()`.

Using `netify()`, we can convert dyadic data into different network formats:

- a cross-sectional, symmetric, and weighted network
- a cross-sectional, unsymmetric, and weighted network
- a cross-sectional, symmetric, and unweighted network
- a cross-sectional, unsymmetric, and unweighted network

For illustration purposes, let's think of a cross-sectional network as one time point of the network, or as an aggregated network (for example, several years of dyadic data aggregated into one time period). Later, we will show you how to create a longitudinal network.

The function `netify()` takes a single data.frame: `dyad_data`. It returns a netlet object. 


```{r, eval = FALSE}
netify(
  dyad_data, 
  actor1=NULL, actor2=NULL, time=NULL,
  symmetric=TRUE, 
  mode='unipartite',  
  weight=NULL, 
  sum_dyads=TRUE, 
  actor_time_uniform=TRUE,
  actor_pds=NULL,
  diag_to_NA=TRUE,
  missing_to_zero=TRUE,
  output_format = ifelse(is.null(time), "cross_sec", "longit_list"),
  nodal_vars=NULL,dyad_vars=NULL, dyad_vars_symmetric=NULL
  )
```

`dyad_data` is a dyadic data.frame that should have at least the following variables used to specify actors:

  - `actor1`: character indicating actor 1 variable in the data
  - `actor2`: character indicating actor 2 variable in the data

`netify_type` is a type of netlet object ('cross-sec', 'longit_list', or 'longit_array'). 

If the data is symmetric, then the order of the actors does not matter. If the data is not symmetric, then we can think of actor 1 as the 'sender' and actor 2 as the 'receiver' in a given exchange.

Other arguments:

- `weight`: character indicating weighted edge variable in the data, default is `weight=NULL`
- `sum_dyads`: logical indicating whether to sum up the `weight` value when there exists repeating dyads.
- `symmetric`: logical indicating whether ties are symmetric, default is `TRUE`
- `nodal_vars`: a vector of which variables from nodeData should be merged
- `dyad_vars`: a vector of which variables from nodeData should be merged
- `dyad_vars_symmetric`: logical vector indicating whether ties are symmetric, default is `TRUE`
- `isolates`: logical indicating whether to include isolates, default is `FALSE`
- `diag_to_NA`: logical indicating whether diagonals should be set to NA, default is `TRUE`
- `missing_to_zero` logical indicating whether missing values should be set to zero, default is `TRUE`
- `nodal_data`: nodal data attribute
- `dyad_data`: dyad data attribute
- `graph_data`: graph data attribute

Let's transform our dyadic dataset, `cow_dyads`, into an adjacency matrix for analysis. It is important we give the object a name. In this case, we are creating an aggregated matrix across all years and summing repeating dyads.


```{r}
mid_network <- netify(
   cow_dyads,
   actor1='ccode1', actor2='ccode2',
   weight='cowmidonset', 
   sum_dyads=TRUE, symmetric=TRUE,
   diag_to_NA=TRUE, missing_to_zero=TRUE)
   mid_network 
```

The above code creates a basic network using conflict between states (MIDs) as the weight variable that characterizes the links or edges (links) in our network. The resulting network object is at the core of the `netify` package and is what we will pass to most of the other functions.

Once we have successfully made a new adjacency matrix from our dyadic data, `netify` returns a helpful message. The output tells us a number of important features about the network: if is symmetric, the number of actors, and the number of edges (links or ties) between actors. Note that we have provided a weight value (you must supply a weight if you chose to sum up dyadic interactions). In this case, the output tells us the network is 'cross-sectional' because we have aggregated the years into one time slice. If there are no repeating dyads, then the function returns the weight for each i,j. If you have repeating dyads like we do in this example and you want to sum the weight variable for every time i,j interact in the data (for example, if they appear multiple times across 'years'), you need to set `sum_dyads=TRUE`).

What if we wanted a network that represented conflict not as a count of interactions but as a binary outcome? First, we subset the original data to create an edgelist like format. Here we only include edges when they are not zero (i.e., they have a conflict). Next we use netify to create a matrix. Here we will not specify a weight value and we will not sum dyads.

```{r}
cow_dyad_conflict_onset <- cow_dyads %>%
        filter(
        cowmidonset > 0
        )

any_mid_network <- netify(
   cow_dyad_conflict_onset,
   actor1='ccode1', actor2='ccode2',
   sum_dyads=FALSE, symmetric=TRUE,
   diag_to_NA=TRUE, missing_to_zero=TRUE)
any_mid_network
```


Now that we have a basic network to inspect, let's take a look at the resulting network object. The quickest and easiest way to do this is using the `peek` function. 

```{r}
peek(mid_network, 20)
peek(any_mid_network, 20)
```

### Longitudinal networks

Sometimes we have data that contains information about time and we perserve the temporal information to create longitudinal networks. If we want to create an object wherein each network represents a single time period in our data then we will need to create a *longitudinal* network. We can do this using the `netify()` function.

Relevant arguments:

- `netify_type` is a type of netlet object ('cross-sec', 'longit_list', or 'longit_array'). For a longitudinal network, 'longit_list', or 'longit_array' should be given. 
- `sum_dyads`: a logical indicating whether or not to assume the actors in the network (or nodes) are the same across an entire time series. This is more relevant for time series data.
- `time`: character indicating the variable representing time in the data. If none is provided, the function assumes the network is cross-sectional.
- `actor_time_uniform`: logical indicating whether to assume actors are the same across the full time series observed in the data. `TRUE` means that actors are the same across the full time series observed in the data. `FALSE` means that actors come in and out of the observed data and their "existence" should be determined by the data, meaning that their first year of existence will be determined by the time point of their first event and their last year of existence by the time point of their last event.
- `actor_pds`: a data.frame indicating start and end time point for every actor, this can be created using `get_actor_time_info.R`, unless provided this will estimated for the user based on their choice of `actor_time_uniform`.


```{r, eval=FALSE}
netify(
  dyad_data,
  actor1, actor2,
  sum_dyads,
	time,
	actor_time_uniform=TRUE,
	actor_pds=NULL,
	nodal_vars=NULL,dyad_vars=NULL, dyad_vars_symmetric=NULL,
  diag_to_NA=TRUE, missing_to_zero=TRUE
)
```

Two key arguments help us understand what kind of longitudinal network we want to create. `time` denotes which variable from the data represents time. `actor_time_uniform` indicates whether or not the actors in the network change over time. If the actors do not change, then we set `actor_time_uniform=TRUE`, and
`netify` returns a three-dimensional array in which the first two dimensions correspond to actors and the third corresponds to time.

If the actors change, then we set `actor_time_uniform = FALSE` and the function returns a list object. Each object in the list will be a matrix at a given time point. The dimensions of each matrix inside the list will depend on how many actors (or nodes) "exist" for that time slice.

Let's try to create an over-time network using the `year` variable in `cow_dyads`. For longitudinal networks, we also need to consider whether or not the nodes, or actors, change overtime in our network. In our case, countries exit and enter the international system so we will set `actor_time_uniform=FALSE`.

Our dataset also includes information about actors and dyadic relationships. The netify function also allows users to add nodal and dyadic variables into the network object. Below, we add covariate information for regime type to the actors (nodes) in the network using the `nodal_vars` parameter. We also add information about the geographical distance between each country in a given dyads using the `dyad_vars` parameter. 

```{r, eval=TRUE}
mid_long_network <- netify(
  cow_dyads,
  actor1='ccode1', actor2='ccode2', time='year',
  weight='cowmidonset', 
  actor_time_uniform=FALSE,
  sum_dyads=FALSE, symmetric=TRUE,
  diag_to_NA=TRUE, missing_to_zero=FALSE,
  nodal_vars = c('v2x_polyarchy1', 'v2x_polyarchy2'),
  dyad_vars = c('capdist'),
  dyad_vars_symmetric = c(TRUE, FALSE)
  )
mid_long_network

```

We now have a longitudinal network and this is reflected in the output which notes there are 23 time periods. Our netlet object is now a list where in each element reflects a time period of the network.

At this point we might wonder, what would have happened if we set `actor_time_uniform=FALSE`? We want to set this parameter to TRUE if the number of actors, or nodes, in the network changes over time and false if it does not. Since we previously told the function that the actors change over time, let's see if the number of actors changes for each timeslice.

```{r, eval=TRUE}
head(lapply(mid_long_network, dim))
```

Our actors are  changing over time! And the resulting object is an organized list of networks over time.

```{r, eval=TRUE}
is.list(mid_long_network)
```

You can inspect these outputs on your own. Or you can use `netify`'s helper functions. Note, to choose a specific time point, such as year, you must pass it to the parameter using quotations. For example, to take a peek at our network at different time points:

```{r, error=TRUE}
# use the index location
peek(mid_long_network, when_to_peek = c(20,23), what_to_peek = 10)
peek(mid_long_network, when_to_peek = c('2012'), what_to_peek = 30)
```

Finally, we might want to supply our own set of actors with custom start and end years. To do this, use the `actors_pds` argument which allows us to pass a data.frame to `netify` with actors' start and end years.

```{r, error=TRUE}
actor_comp <- data.frame(
    actor = c(2, 365, 220, 710),
    min_time = c(1980, 1980, 1991, 1980),
    max_time = c(2001, 2001, 2001, 2001)
)

# pass actor composition to netify
mid_network_actors <- netify(
  cow_dyads,
  actor1='ccode1', actor2='ccode2', time='year',
  weight='cowmidonset', 
  actor_pds=actor_comp,
  sum_dyads=FALSE, symmetric=TRUE,
  diag_to_NA=TRUE, missing_to_zero=TRUE)
mid_network_actors
```

## Summary Functions

Once we have made a `netify` object to fit our needs, we can inspect that object in three main ways using the package functions. 

  (1) As already seen, once the object is printed, we receive a message with some basic information about their network. 
  (2) We have also seen how to inspect the object by turning to the `peek` function. 
  (3) Below we explore the use of netify's `summary` functions. 

The `summary()` function returns valuable summary information about the network:

- numActors: Number of actors in the network
- dens: network density
- recip: reciprocity
- sdSendEff: sender effects
- sdRecEff: sender effects
- srCov: sender receiver covariance

```{r}
sum_event_mat <- summary(mid_network)
sum_event_mat
```

Users can also supply their own summary functions.  The output is an easy to digest dataframe.

```{r}
transitivity = function(mat){
  E <- mat-mean(mat,na.rm=TRUE)
  D <- 1*(!is.na(E))
  E[is.na(E)] <- 0
  trans <- sum(diag(E%*%t(E)%*%E))/( sum(diag(D%*%t(D)%*%D))*sd(c(mat),na.rm=TRUE)^3)
  return(trans)
}

cycles = function(mat){
  E <- mat-mean(mat,na.rm=TRUE)
  D <- 1*(!is.na(E))
  E[is.na(E)] <- 0
  cycles <- sum(diag(E%*%E%*%E))/( sum(diag(D%*%D%*%D))*sd(c(mat),na.rm=TRUE)^3)
  return(cycles)
}

sum_event_mat <- summary(mid_network, cycles=cycles, transitivity=transitivity)
sum_event_mat
```

The second summary function is called `summary_actor`. The `summary_actor()` function returns useful actor level statistics for the network object. This function returns a data frame object where N equals the number of actors in the network.


```{r, eval = FALSE}
network_sum <- summary_actor(mid_network)
head(network_sum)
```

## Visualization

Netify provides its own plotting functions for both network- and actor-level statistics for netify objects.

The first plotting function is called `plot_networkStats`. The `plot_networkStats()` function returns a graph object summarizing network-level statistics when it is longitudinal. It has three main parameter: `netlet`, `logitudinal`, and `net_stat`. Since it plots a network-level visualization, `logitudianl` should always set to `TRUE`. For `net_stat`, you can pick summary statistics of your interest from the netlet object you created with the `summary` function. The default is "dens".

```{r, echo=FALSE}
# note: needs another look
nets_time_sum <- summary(mid_long_network)
```

Next, we can use ggplot to visualize these statistics.  

```{r}
library(ggplot2)
plot_networkStats(nets_time_sum, longitudinal = TRUE, net_stat=c("sdSendEff", "sdRecEff")) +
  labs(title = "Network-level Statistics") +
  theme_bw()
```

The second plotting function is called `plot_actorStats`. The `plot_actorStats()` function returns a actor-level visualization. It has five main parameter: `netlet`, `logitudinal`, `net_stat`, `actor_names`, and `toplist_n`. 

Using `actor_names` and `toplist_n` you can pick actors in your network and the number of actors to be listed as top n actors in your visualization. More specifically, `actor_names` is character indicating actor name(s) in the network, and `toplist_n` is numeric indicating the number of actors to be listed as top n actors. 


```{r, eval = FALSE}
# compare USA (2), China (710), Russia (365)
plot_actorStats(network_sum, longitudinal=FALSE, net_stat=c('average_degree_in','average_degree_out'),
actor_names=c('2', '710', '365'), toplist_n=NULL)
```

```{r, eval = FALSE}
plot_actorStats(network_sum, longitudinal=FALSE, net_stat='average_degree_in', toplist_n=10)
```


## Using netify as a bridge to other network packages

Netify provides prep functions that allow users to reshape the `netlet` object to make it ready for analysis with the following two packages:

- `prep_for_igraph`: [igraph](https://igraph.org/r/)
- `prep_for_network`: [network](https://cran.r-project.org/web/packages/network/index.html)
- `prep_for_amen`:[amen](https://pdhoff.github.io/amen/)

You can simply provide a netlet object into these prep functions to get your data ready for `igraph` and network functions. 

```{r}
library(igraph)
# return to***
# do i have to subset and then make a netify again? or can peek do that?
# netplot is a matrix not netlet obj
net_plot <- peek(mid_network, 20)

igraph_obj <- prep_for_igraph(net_plot)
igraph_obj

rm_iso = which(degree(igraph_obj)==0)
igraph_obj = delete.vertices(igraph_obj, rm_iso)
graph_attr(igraph_obj, "layout") <- layout.random

plot(igraph_obj, edge.arrow.size=.1,vertex.label=NA)
```

Many researchers and users will want to use the network package to do analysis using `statnet` or `ERGM` models and packages. The netify package includes functions that take the network object (`netlet`) and quickly prepare the data for use in these other packages. See `network` and related packages documentation for more information. 

```{r}
network_obj <- prep_for_network(mid_long_network)
network_obj
```

Similarly, users can employ the `prep_for_amen` function which will take the netlet object and transform it into a list object with the necessary components to pass to the AME function. 

```{r}
amen_obj <- prep_for_amen(mid_long_network)
amen_obj
```

