---
title: "Pipeline: netify to lame and dbn"
description: "Prepare netify objects for lame and dbn models."
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipeline: netify to lame and dbn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = FALSE,
    fig.width = 7,
    fig.height = 5
)
```

Latent space models for network data require array inputs that are easy to get wrong by hand. The `amen` package handles cross-sectional and single-layer longitudinal networks, `lame` adds support for longitudinal networks with changing actor sets, and `dbn` fits Dynamic Bilinear Network models to multilayer longitudinal networks. Each package expects a different shape: 3D arrays, lists of matrices, or 4D arrays. `netify()` and `layer_netify()` build the network object; `to_amen()`, `to_lame()`, and `to_dbn()` export the shape needed by the modeling package.

The conversion chunks use `netify`. The model-fitting chunks require `lame` or `dbn`; install those packages before running them. You can still run the conversion chunks to inspect the arrays that `netify` creates.

## setup

The examples below use ICEWS event data, which captures directed interactions between 152 countries from 2002 to 2014 across four relation types: verbal cooperation, material cooperation, verbal conflict, and material conflict.

```{r}
library(netify)
library(ggplot2)

# load the icews event data
data(icews)

# preview
head(icews[, c("i", "j", "year", "verbCoop", "matlCoop", "verbConf", "matlConf")], 4)
```

## part 1: single-layer pipeline (netify to amen)

### step 1: create the network

The first step is always `netify()`. We create a single directed, weighted network representing verbal cooperation over time.

```{r}
verbal_net = netify(
    icews,
    actor1 = "i", actor2 = "j", time = "year",
    symmetric = FALSE,
    weight = "verbCoop",
    missing_to_zero = FALSE,
    nodal_vars = c("i_polity2", "i_log_gdp"),
    dyad_vars = c("matlCoop", "verbConf"),
    dyad_vars_symmetric = c(FALSE, FALSE),
    output_format = "longit_array"
)

verbal_net
```

Key choices:

- **`missing_to_zero = FALSE`**: Preserving NAs distinguishes "no observed interaction" from "structurally impossible interaction." This matters for latent space models where missing data and structural zeros have different likelihoods.
- **`output_format = "longit_array"`**: We use array format because our actor set is constant across time.
- **`nodal_vars` and `dyad_vars`**: These covariates will be carried through to the amen output automatically.

### step 2: inspect the network

Before modeling, it's worth checking basic properties.

```{r}
# quick summary
net_summary = summary(verbal_net)
head(net_summary[, c("net", "num_actors", "density", "num_edges", "reciprocity")])
```

```{r, fig.width=8, fig.height=4}
# visualize a single year
verbal_2010 = subset(verbal_net, time = "2010")
plot(verbal_2010, add_text = FALSE)
```

### step 3: convert to amen format

`to_amen()` converts a single-layer netify object into the list structure that `amen::ame()` expects.

```{r}
amen_data = to_amen(verbal_net)

# inspect the resulting array shapes
data.frame(
    object = c("Y", "Xdyad", "Xrow", "Xcol"),
    dims = c(
        paste(dim(amen_data$Y), collapse = " x "),
        paste(dim(amen_data$Xdyad), collapse = " x "),
        paste(dim(amen_data$Xrow), collapse = " x "),
        paste(dim(amen_data$Xcol), collapse = " x ")
    )
)
```

The output is:

- **Y**: `[n_actors x n_actors x n_time]`, the adjacency array
- **Xdyad**: `[n_actors x n_actors x n_dyad_vars x n_time]`, dyadic covariates
- **Xrow**: `[n_actors x n_nodal_vars x n_time]`, sender covariates
- **Xcol**: `[n_actors x n_nodal_vars x n_time]`, receiver covariates

These can be plugged directly into `amen::ame()`:

```{r, eval = FALSE}
library(amen)

# cross-sectional model for a single year
ame_fit = ame(
    Y = amen_data$Y[, , "2010"],
    Xdyad = amen_data$Xdyad[, , , "2010"],
    Xrow = amen_data$Xrow[, , "2010"],
    Xcol = amen_data$Xcol[, , "2010"],
    R = 2,
    symmetric = FALSE,
    family = "nrm",
    nscan = 1000, burn = 500, odens = 1,
    plot = FALSE, print = FALSE
)
```

## part 2: longitudinal pipeline with lame

For longitudinal modeling workflows that need per-period matrices, use `to_lame()`. This is a thin specialization of `to_amen()` that (a) can return ragged per-period matrices when actor sets vary, (b) emits a ready-to-run padding + `lame::lame()` snippet, and (c) provides `from_lame_fit()` for round-tripping posterior predictions back into a netify for plotting. The ICEWS example below has the same actor set in each year, but the format also supports open-cohort panels.

### when to use lame format

Use `to_lame()` when:

- Actors enter and exit the network over time (e.g., new states forming, organizations dissolving)
- You want to fit longitudinal latent space models with `lame`

### create a per-period list for lame

```{r}
# use longit_list format for varying actor compositions
verbal_list = netify(
    icews,
    actor1 = "i", actor2 = "j", time = "year",
    symmetric = FALSE,
    weight = "verbCoop",
    missing_to_zero = FALSE,
    actor_time_uniform = FALSE,
    nodal_vars = c("i_polity2", "i_log_gdp"),
    dyad_vars = c("matlCoop"),
    dyad_vars_symmetric = c(FALSE)
)

verbal_list
```

### convert to lame format

```{r}
nl = to_lame(verbal_list, lame = TRUE)

# inspect the shape of the returned per-period matrices
data.frame(
    description = c("number of periods", "first period dims", "last period dims"),
    value = c(
        length(nl$Y),
        paste(dim(nl$Y[[1]]), collapse = " x "),
        paste(dim(nl$Y[[length(nl$Y)]]), collapse = " x ")
    )
)

# check actor counts across time
actor_counts = sapply(nl$Y, nrow)
actor_counts
```

Each element of the list can have a different set of actors. `to_lame()` also bundles a copy-paste-ready snippet that pads the ragged list into a 3D array with `lame::list_to_array()` and then fits `lame::lame()`:

```{r}
cat(nl$ame_call)
```

Run that snippet to fit the model. Conceptually:

```{r, eval = FALSE}
library(lame)

U <- unique(unlist(lapply(nl$Y, rownames)))
padded <- lame::list_to_array(actors = U, Y = nl$Y, Xdyad = nl$Xdyad,
    Xrow = nl$Xrow, Xcol = nl$Xcol)

lame_fit = lame::lame(
    Y = padded$Y, Xdyad = padded$Xdyad,
    Xrow = padded$Xrow, Xcol = padded$Xcol,
    R = 2, family = "normal",
    method = "mcmc", nscan = 1000, burn = 500
)
```

### round-tripping fitted values back to netify

Once a fit is in hand, `from_lame_fit()` pulls the posterior-mean linear predictor (or a residual / probability / per-cell quantile) back into a cross-sectional netify so it can be plotted or compared against the observed network. In the example below, `lame_fit` is the object created by running the model-fitting snippet above:

```{r, eval = FALSE}
# posterior-mean linear predictor
pred_net = from_lame_fit(lame_fit, value = "fitted")
plot(pred_net, style = "heatmap")

# 90% per-cell credible interval bounds on the probability scale
# (only for binary families; here for illustration)
lo = from_lame_fit(lame_fit, value = "prob_lower", alpha = 0.05)
hi = from_lame_fit(lame_fit, value = "prob_upper", alpha = 0.05)
```

`from_lame_fit()` auto-detects the link function: probit for `lame::ame_als()` / `amen::ame()` fits, logit for `lame::lame()` Gibbs fits, with an explicit `link` slot taking precedence.

## part 3: multilayer pipeline (netify to dbn)

The `dbn` package models multilayer longitudinal networks using Dynamic Bilinear Network models. It expects a 4D array with dimensions `[n, n, p, T]` where `p` is the number of relation types (layers).

### step 1: create individual layer networks

```{r}
# verbal cooperation layer (symmetric - mutual diplomatic engagement)
verbal_coop = netify(
    icews,
    actor1 = "i", actor2 = "j", time = "year",
    symmetric = TRUE,
    weight = "verbCoop",
    missing_to_zero = FALSE,
    output_format = "longit_array",
    nodal_vars = c("i_polity2", "i_log_gdp"),
    dyad_vars = c("verbConf"),
    dyad_vars_symmetric = c(TRUE)
)

# material cooperation layer (directed - trade/aid flows)
material_coop = netify(
    icews,
    actor1 = "i", actor2 = "j", time = "year",
    symmetric = FALSE,
    weight = "matlCoop",
    missing_to_zero = FALSE,
    output_format = "longit_array",
    dyad_vars = c("matlConf"),
    dyad_vars_symmetric = c(FALSE)
)
```

### step 2: combine into a multilayer network

`layer_netify()` supports mixed directedness, meaning layers can have different symmetry settings. This is common in IR applications where some relations are inherently symmetric (alliance status) while others are directed (trade exports).

```{r}
multi_net = layer_netify(
    list(verbal_coop, material_coop),
    layer_labels = c("Verbal", "Material")
)

multi_net
```

Notice that the print output shows the mixed directedness across layers.

### step 3: convert to dbn format

`to_dbn()` extracts the multilayer longitudinal netify object into the exact 4D array format that `dbn` expects.

```{r}
dbn_data = to_dbn(multi_net)

# what did we get?
cat("Y dimensions:", paste(dim(dbn_data$Y), collapse = " x "), "\n")
cat("  Actors:", dim(dbn_data$Y)[1], "\n")
cat("  Layers:", paste(dimnames(dbn_data$Y)[[3]], collapse = ", "), "\n")
cat("  Time periods:", dim(dbn_data$Y)[4], "\n")
```

The output structure:

- **Y**: `[n_actors x n_actors x n_layers x n_time]`, the 4D adjacency array
- **Xdyad**: `[n_actors x n_actors x n_dyad_vars x n_time]`, dyadic covariates
- **Xrow**: `[n_actors x n_nodal_vars x n_time]`, sender covariates
- **Xcol**: `[n_actors x n_nodal_vars x n_time]`, receiver covariates

```{r, eval = FALSE}
# fit a dynamic bilinear network model
library(dbn)

dbn_fit = dbn(
    Y = dbn_data$Y,
    R = 2,
    nscan = 1000, burn = 500, odens = 1,
    plot = FALSE, print = FALSE
)
```

### single-layer networks work too

`to_dbn()` also handles single-layer longitudinal networks by adding a layer dimension of size 1:

```{r}
dbn_single = to_dbn(verbal_coop)
cat("Single-layer Y dimensions:", paste(dim(dbn_single$Y), collapse = " x "), "\n")
cat("Layer name:", dimnames(dbn_single$Y)[[3]], "\n")
```

## part 4: working with mixed directedness

A common challenge in political science is that data naturally includes both symmetric and directed relations. For example:

- **Symmetric**: UNGA voting agreement, alliance status, shared IGO membership
- **Directed**: trade exports, foreign aid, arms transfers

`netify` now handles this directly.

### creating mixed-directedness multilayer networks

```{r}
# symmetric layer: average verbal cooperation
verbal_symm = netify(
    icews,
    actor1 = "i", actor2 = "j", time = "year",
    symmetric = TRUE,
    weight = "verbCoop",
    missing_to_zero = FALSE,
    output_format = "longit_array"
)

# directed layer: material cooperation
material_dir = netify(
    icews,
    actor1 = "i", actor2 = "j", time = "year",
    symmetric = FALSE,
    weight = "matlCoop",
    missing_to_zero = FALSE,
    output_format = "longit_array"
)

# combine -- mixed directedness is allowed
mixed_net = layer_netify(
    list(verbal_symm, material_dir),
    layer_labels = c("Verbal_Symm", "Material_Dir")
)

# check attributes
cat("Symmetric attribute:", attr(mixed_net, "symmetric"), "\n")
cat("Layer names:", attr(mixed_net, "layers"), "\n")
```

When you subset to a single layer, the correct symmetry is preserved:

```{r}
# subset to the symmetric layer
verbal_only = subset(mixed_net, layers = "Verbal_Symm")
cat("Verbal layer symmetric:", attr(verbal_only, "symmetric"), "\n")

# subset to the directed layer
material_only = subset(mixed_net, layers = "Material_Dir")
cat("Material layer symmetric:", attr(material_only, "symmetric"), "\n")
```

### converting mixed-directedness to dbn

`to_dbn()` handles mixed-directedness multilayer objects:

```{r}
dbn_mixed = to_dbn(mixed_net)
cat("Y dimensions:", paste(dim(dbn_mixed$Y), collapse = " x "), "\n")
```

Note that the 4D array represents all layers uniformly. The per-layer symmetry information is metadata that you would pass to your model separately. For dbn, you would typically specify this in the model call.

## part 5: the missing_to_zero decision

The `missing_to_zero` parameter deserves special attention for modeling pipelines. By default, `netify()` sets `missing_to_zero = TRUE`, which fills unobserved dyads with zeros. That is appropriate when an unrecorded dyad means no event occurred, but it is not appropriate when unrecorded dyads were never observed.

### why it matters

```{r, fig.width=8, fig.height=4}
observed_contacts = data.frame(
    i = c("a", "b"),
    j = c("b", "c"),
    contacts = c(2, 1),
    stringsAsFactors = FALSE
)
all_people = c("a", "b", "c", "d")

net_zeros = netify(
    observed_contacts,
    actor1 = "i", actor2 = "j",
    symmetric = FALSE,
    weight = "contacts",
    nodelist = all_people,
    missing_to_zero = TRUE
)

net_na = netify(
    observed_contacts,
    actor1 = "i", actor2 = "j",
    symmetric = FALSE,
    weight = "contacts",
    nodelist = all_people,
    missing_to_zero = FALSE
)

# count zeros vs nas
raw_zeros = get_raw(net_zeros)
raw_na = get_raw(net_na)

cat("With missing_to_zero = TRUE:\n")
cat("  Zeros:", sum(raw_zeros == 0, na.rm = TRUE), "\n")
cat("  NAs:", sum(is.na(raw_zeros)), "\n")

cat("\nWith missing_to_zero = FALSE:\n")
cat("  Zeros:", sum(raw_na == 0, na.rm = TRUE), "\n")
cat("  NAs:", sum(is.na(raw_na)), "\n")
```

For latent space models, the distinction matters:

- **Zero**: "We observed this dyad and there was no interaction," which informs the model that these actors are distant in latent space
- **NA**: "We don't know whether these actors interacted," so the model marginalizes over possible values

**Recommendation for modeling pipelines**: Always use `missing_to_zero = FALSE` unless you are confident that all unobserved dyads represent genuine zeros.

## quick reference: choosing your export function

| Scenario | Function | Output |
|----------|----------|--------|
| Single-layer, cross-sectional | `to_amen()` | list(Y, Xdyad, Xrow, Xcol) |
| Single-layer, longitudinal (constant actors) | `to_amen()` | 3D arrays |
| Single-layer, longitudinal (varying actors) | `to_lame(lame=TRUE)` | per-period lists + padding / fit snippet |
| Round-trip AME/LAME fit back to netify | `from_lame_fit()` | netify (fitted / residual / prob / quantiles) |
| Multilayer, longitudinal | `to_dbn()` | 4D array `[n, n, p, T]` |
| Single-layer, longitudinal (dbn format) | `to_dbn()` | 4D array `[n, n, 1, T]` |

## tl;dr

The `netify` pipeline eliminates manual array construction by providing a four-step workflow: create the network with `netify()` (using `missing_to_zero = FALSE` for modeling applications), combine layers with `layer_netify()` when multilayer structure is needed, export with `to_amen()` / `to_lame()` / `to_dbn()` depending on the target modeling package, and pass the resulting output (or the bundled `ame_call` snippet) directly to the model fitting function. After fitting, `from_lame_fit()` round-trips posterior predictions back into a netify so they can be plotted alongside the observed network.
