---
title: "Pipeline: netify to ergm (statnet)"
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipeline: netify to ergm (statnet)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    dev = "png", dpi = 150,
    cache = FALSE,
    echo = TRUE,
    collapse = TRUE,
    comment = "#>"
)

# this vignette needs network + ergm; gracefully exit if not installed
pkg_check <- c("network", "ergm")
missing_pkgs <- pkg_check[!vapply(pkg_check, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
    message(
        "This vignette requires: ", paste(missing_pkgs, collapse = ", "), "\n",
        "Install with: install.packages(c(",
        paste0("'", missing_pkgs, "'", collapse = ", "), "))"
    )
    knitr::knit_exit()
}
```

ERGMs (Exponential Random Graph Models) are the workhorse of inferential network analysis in the [statnet](https://statnet.org/) ecosystem. `netify` provides a clean bridge: build your network with `netify()`, attach attributes with `add_node_vars()` / `add_dyad_vars()`, then convert to the `network` format that `ergm` expects with `to_statnet()`.

This vignette covers:

1. The cross-sectional pipeline (single network → single ergm fit)
2. The longitudinal pipeline (per-time ergm fits)
3. The multilayer pipeline (per-layer ergm fits — `to_statnet()` now iterates layers automatically)
4. Round-tripping ergm-simulated networks back into netify for descriptive checks

```{r}
library(netify)
library(ggplot2)
library(network)
library(ergm)

data(icews)
```

## 1. Cross-sectional pipeline

The simplest case: one snapshot, one model.

```{r}
# build a single-year netify object with nodal and dyadic attributes
icews_2010 <- icews[icews$year == 2010, ]

verb_coop <- netify(
    icews_2010,
    actor1 = "i", actor2 = "j",
    symmetric = FALSE,
    weight   = "verbCoop",
    nodal_vars = c("i_polity2", "i_log_gdp"),
    dyad_vars  = "matlCoop"
)

# convert to statnet 'network' object
sn_2010 <- to_statnet(verb_coop)
sn_2010
```

`to_statnet()` carries nodal attributes through as vertex attributes and dyadic attributes through as edge attributes.

### Before you fit: three sanity checks

ERGMs fail in cryptic ways when the underlying network is malformed. Three checks catch most "invalid output from statistic" headaches before you ever call `ergm()`:

```{r}
# 1. NAs in nodal covariates referenced by nodecov / nodematch
nd <- attr(verb_coop, "nodal_data")
na_cols <- names(nd)[vapply(nd, function(c) any(is.na(c)), logical(1))]
na_cols <- setdiff(na_cols, c("actor", "time", "layer"))
na_cols

# 2. isolates -- ergm fits but some terms (e.g. gwdegree) degenerate
m <- get_raw(verb_coop)
bin <- (m != 0) & !is.na(m)
deg_total <- rowSums(bin) + colSums(bin)
isolates <- rownames(m)[deg_total == 0]
length(isolates)

# 3. symmetry: the netify flag must match how your model treats ties
attr(verb_coop, "symmetric")
```

If `na_cols` is non-empty, drop the affected actors with `drop_na_actors(verb_coop, cols = na_cols)` or impute before converting. Now fit an ergm:

```{r, eval = FALSE}
# Note: this chunk is not evaluated by default to keep vignette build fast.
# Replace eval = FALSE with eval = TRUE to actually run.
set.seed(6886)
m <- ergm(
    sn_2010 ~ edges +
        nodecov("i_polity2") +
        nodecov("i_log_gdp")
)
summary(m)
```

One footgun worth flagging: `nodecov("i_polity2")` will refuse to fit if any vertex has an `NA` polity score. Either subset to actors with complete covariates or impute before passing to `to_statnet()`. `netify` emits a one-shot inform when it detects NAs in a nodal attribute and stashes the offending column names on the resulting network object so you can introspect:

```{r, eval = FALSE}
attr(sn_2010, "netify_na_cols")   # character vector of NA-bearing nodal vars
```

If non-empty, drop or impute those columns before fitting `ergm()` formulas that reference them with `nodecov()` / `nodematch()`. The `drop_na_actors()` helper does this in one call and works for cross-sectional, longitudinal, and bipartite netlets:

```{r, eval = FALSE}
clean <- drop_na_actors(verb_coop, cols = attr(sn_2010, "netify_na_cols"))
sn_2010_clean <- to_statnet(clean)
```

### Dyadic edge covariates: the `_e` suffix

Any dyadic covariate you passed to `netify()` (e.g. `dyad_vars = "matlCoop"`) is attached to the resulting `network` object in two places:

- as a network-level attribute under its original name (the full `n x n` matrix), accessible via `network::get.network.attribute(sn, "matlCoop")`, and
- as a per-edge attribute under `<var>_e` (here, `matlCoop_e`), populated only on edges that actually exist.

The trailing `_e` disambiguates the per-edge edgelist from the network-level matrix. For `ergm::edgecov()`, pass the **original (matrix) name** — `edgecov()` resolves its argument as a network-level matrix attribute, so the `_e` per-edge alias will not work there:

```{r, eval = FALSE}
m <- ergm(sn_2010 ~ edges + edgecov("matlCoop"))
```

The `_e` per-edge attribute is exposed for descriptive uses (for example, coloring edges by covariate in `network::plot.network`). `to_statnet()` emits a one-shot inform listing both forms the first time it attaches dyadic covariates.

Goodness-of-fit, mcmc.diagnostics, and other postestimation tools live in `ergm` itself.

## 2. Longitudinal pipeline (per-time fits)

For a longitudinal netify, `to_statnet()` returns a named list — one `network` object per time period:

```{r}
verb_longit <- netify(
    icews[icews$year %in% 2010:2012, ],
    actor1 = "i", actor2 = "j", time = "year",
    symmetric = FALSE,
    weight = "verbCoop",
    nodal_vars = "i_polity2"
)

sn_list <- to_statnet(verb_longit)
length(sn_list)
names(sn_list)
class(sn_list[[1]])
```

Fit a separate ergm per period:

```{r, eval = FALSE}
set.seed(6886)
fits <- lapply(sn_list, function(n) {
    ergm(n ~ edges + nodecov("i_polity2"))
})
```

For *coevolution* models (where ties change as a function of past ties), look at `tergm` from the statnet suite — `netify` provides the data, `tergm` does the modeling. A typical tergm 4.x call against the same per-period list looks like this:

```{r, eval = FALSE}
# Longitudinal ERGM via tergm 4.x
library(tergm)
set.seed(6886)

nets <- to_statnet(verb_longit)   # named list of network objects
fit <- tergm(
    nets ~ Form(~ edges + nodecov("i_polity2")) +
           Persist(~ edges),
    estimate = "CMLE",
    times = seq_along(nets)
)
summary(fit)
```

In the tergm 4.x split formulation, `Form()` models the *formation* of new edges between periods, and `Persist()` models the *persistence* of existing edges from one period to the next; both formulas accept the same ergm terms (`edges`, `nodecov`, `nodematch`, `mutual`, `gwesp`, etc.). (Note: `i_polity2` has missing values for some country-years — `to_statnet()` flags this and stashes the offending column names on `attr(net, "netify_na_cols")` so you can drop or impute before fitting.)

## 3. Multilayer pipeline (per-layer fits)

Multilayer ergm modeling is its own research literature. The simple pragmatic approach is to fit one ergm per layer. `to_statnet()` handles this automatically: pass a multilayer netify, get back a named list keyed by layer.

```{r}
verb <- netify(icews_2010, actor1 = "i", actor2 = "j",
               symmetric = FALSE, weight = "verbCoop")
matl <- netify(icews_2010, actor1 = "i", actor2 = "j",
               symmetric = FALSE, weight = "matlCoop")
multi <- layer_netify(list(verbal = verb, material = matl))

sn_multi <- to_statnet(multi)
length(sn_multi)
names(sn_multi)
```

Each element is a `network` object you can plug straight into `ergm()`.

## 4. Round-tripping simulated networks back to netify

A useful descriptive check after fitting an ergm is to simulate from the fit, compute descriptives on the simulated networks, and compare to the observed network. `simulate.ergm()` returns `network` objects, which means you can pipe them straight back into a netify for any of netify's descriptive tools:

```{r, eval = FALSE}
set.seed(6886)
sims <- simulate(m, nsim = 100)  # list of network objects

# convert each simulated network back into a netify for comparison
sim_nets <- lapply(sims, function(s) to_netify(s))

# compare observed vs simulated at the structural level
all_nets <- c(list(observed = verb_coop), sim_nets)
struct_comp <- compare_networks(all_nets, what = "structure")
```

This gives you observed-vs-simulated comparisons for density, reciprocity, transitivity, mean degree, etc. — useful for assessing whether your ergm captured the descriptive features you care about.

## tl;dr

```r
# build → attach attrs → export → model
net <- netify(df, actor1 = "i", actor2 = "j", symmetric = FALSE, weight = "x")
net <- add_node_vars(net, attrs, actor = "id")
sn  <- to_statnet(net)               # single network, longit list, or multilayer list
m   <- ergm(sn ~ edges + ...)        # modeling happens in ergm/statnet
```

For modeling beyond ergm:

- Latent factor / additive-multiplicative: [`pipeline_lame_dbn`](pipeline_lame_dbn.html) vignette
- Social relations model: [netify-dev/srm](https://github.com/netify-dev/srm) via `to_amen(net)`
- Social influence regression (Minhas & Hoff): [netify-dev/sir](https://github.com/netify-dev/sir)

Have fun!

## References

1. Butts, C.Butts, C. T. (2008). network: A Package for Managing Relational Data in R. Journal of Statistical Software, 24(2), 1–36. https://doi.org/10.18637/jss.v024.i02

2. Cranmer, S. J., Desmarais, B. A., & Morgan, J. W. (2021). Inferential Network Analysis. Cambridge University Press. https://doi.org/10.1017/9781316662915

3. Hunter, D. R., Handcock, M. S., Butts, C. T., Goodreau, S. M., & Morris, M. (2008). ergm: A Package to Fit, Simulate and Diagnose Exponential-Family Models for Networks. Journal of Statistical Software, 24(3), 1–29. https://doi.org/10.18637/jss.v024.i03

4. Krivitsky, P. N., & Handcock, M. S. (2014). A Separable Model for Dynamic Networks. Journal of the Royal Statistical Society: Series B (Statistical Methodology), 76(1), 29–46. https://doi.org/10.1111/rssb.12014

5. Snijders, T. A. B., Pattison, P. E., Robins, G. L., & Handcock, M. S. (2006). New Specifications for Exponential Random Graph Models. Sociological Methodology, 36(1), 99–153. https://doi.org/10.1111/j.1467-9531.2006.00176.x
