---
title: "Netify: Workshop PSSI"
author: "Dorff, Keltner, Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Netify: Workshop PSSI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  dev = 'png', dpi = 150,
  cache = FALSE,
  echo = TRUE
)
```

# Netify Workshop PSSI

Below is a demonstration of how to use the **netify** R package to build and analyze various conflict networks. We will walk through data preparation, network construction, summary statistics, visualizations, and special use cases like weighted networks, longitudinal networks, and bipartite networks.

## Setup

This section checks for necessary packages and installs them if they are missing. We also load the **netify** package from GitHub.

```{r setup}
# Load netify
library(netify)

# Check/install additional packages for the vignette
if(!'tidyverse' %in% rownames(installed.packages())){
  install.packages('tidyverse', repos='https://cloud.r-project.org') }

# Load other libraries needed
library(tidyverse)
```

## Case Examples Using UCDP GED

We will use the [UCDP GED](https://ucdp.uu.se/downloads/) data to demonstrate **netify**. Specifically, we’ll look at conflict events in Mexico and Myanmar (Burma).

### 1. Basic Aggregated Network (Mexican Case)

We first import the GED dataset, filter for Mexico, and construct a simple network (unweighted) based on the number of conflict events between actors.

```{r basic_agg_network}
# Load UCDP GED data on Mexico
data(mexico)

# Construct unweighted network (number_of_events by default)
mex_network <- netify(
  mexico,
  actor1 = 'side_a',
  actor2 = 'side_b',
  symmetric = TRUE,
  sum_dyads = TRUE,
  diag_to_NA = TRUE,
  missing_to_zero = TRUE
)

mex_network_log <- transform_weights(
  mex_network, 
  transform_fn = log,
  add_constant = 1,  # Add 1 before log to handle zeros
  new_name = "log_events"
)

mex_bin <- transform_weights(
  mex_network, 
  transform_fn = function(x) ifelse(x > 0, 1, 0),
  new_name = "conflict_present")


# Summaries at the graph and actor levels
summary(mex_network)

actor_stats <- summary_actor(mex_network)
plot_actor_stats(actor_stats)

# Visualize the network
plot(mex_network, add_text = TRUE)

# Show a subset of 10 random actor names for clarity
select_names <- rownames(mex_network)
set.seed(12345)
random_indices <- sample(length(select_names), 10)
random_names <- select_names[random_indices]

plot(mex_network,
     select_text = random_names,
     select_text_display = random_names)

plot(mex_network,
     select_text = random_names,
     select_text_display = random_names, 
     # log(x+1) to better see range of connects
     weight_transform = log1p
     )
```

### 2. Weighted Network (Mexican Case)

Next, we create a weighted network using the civilian deaths variable.

```{r weighted_network}
# Weighted by civilian deaths
mex_network_civ <- netify(
  mexico,
  actor1 = 'side_a',
  actor2 = 'side_b',
  weight = 'deaths_civilians',
  symmetric = TRUE,
  sum_dyads = TRUE,
  diag_to_NA = TRUE,
  missing_to_zero = TRUE
)

# Summaries and visualization
summary(mex_network_civ)
plot(mex_network_civ, add_text = TRUE)

# Subset of labels for clarity
plot(mex_network_civ,
     select_text = random_names,
     select_text_display = random_names)

# Example: community detection (optimal clustering)
i_opt_memb <- function(x){
  ig <- netify_to_igraph(x)
  memb <- igraph::cluster_optimal(ig)$membership
  return(memb)
}
sum_mex <- summary_actor(
  mex_network_civ, 
  other_stats = list(i_opt_memb = i_opt_memb)
)

# Example: add nodal-level data (e.g., number_of_sources)
size_by_source_count <- mexico |>
  pivot_longer(cols = c(side_a, side_b), 
               names_to = "actor_type", 
               values_to = "actor") |>
  group_by(actor) |>
  summarize(size_by_source_count = sum(number_of_sources, na.rm = TRUE)) |>
  ungroup()

mex_network_civ <- add_node_vars(
  netlet = mex_network_civ,
  node_data = size_by_source_count,
  actor = 'actor',
  node_vars = c('size_by_source_count')
)

# Plot with color and size based on external data
plot(mex_network_civ, 
     point_color_var = 'size_by_source_count',
     point_size=4
)

# Filter large nodes for labeling
large_nodes <- size_by_source_count |> 
  filter(size_by_source_count >= 1000) |> 
  select(actor) |> unlist()

plot(
  mex_network_civ,
  point_color_var = 'size_by_source_count',
  point_size=4,
  select_text = large_nodes,                   
  select_text_display = large_nodes
)
```

### 3. Longitudinal Network (Mexican Case)

We can also create time-varying networks. Here, we use the **year** field to construct a series of networks.

```{r longitudinal_network}
# Create a time-varying network
mex_network_long <- netify(
  mexico,
  actor1 = 'side_a',
  actor2 = 'side_b',
  time = 'year',      
  weight = 'deaths_civilians',
  symmetric = TRUE,
  sum_dyads = TRUE,
  diag_to_NA = TRUE,
  missing_to_zero = TRUE
)  

# Summaries over time at actor level
summary_actor_mex <- summary_actor(mex_network_long)

# Look at distribution of actor stats
plot_actor_stats(summary_actor_mex, across_actor = TRUE)

# Focus on closeness measure
plot_actor_stats(summary_actor_mex, 
                 across_actor = TRUE,
                 specific_stats = 'closeness')

# Visualize
plot.netify(mex_network_long,
            static_actor_positions = FALSE,
            remove_isolates = TRUE)

# Example: subset for selected years
mex_network_short <- subset_netify(mex_network_long,
                                   time = c("2006", "2008", 
                                           "2010", "2012"))
mex_network_short
plot(mex_network_short)
```

## Myanmar UCDP GED Dataset

Let’s now repeat similar steps for conflict data in Myanmar (Burma).

```{r myanmar_data}
# UCDP data for Myanmar
data(myanmar)

# Create basic network
my_network <- netify(
  myanmar,
  actor1 = 'side_a',
  actor2 = 'side_b',
  symmetric = TRUE,
  sum_dyads = TRUE,
  diag_to_NA = TRUE,
  missing_to_zero = TRUE
)

summary(my_network)

# Weighted by civilian deaths
my_network_civ <- netify(
  myanmar,
  actor1 = 'side_a',
  actor2 = 'side_b',
  weight = 'deaths_civilians',
  symmetric = TRUE,
  sum_dyads = TRUE,
  diag_to_NA = TRUE,
  missing_to_zero = TRUE
)

# Plot with labels
plot(my_network_civ, add_text = TRUE)
```

### Adding Nodal Attributes (Myanmar Case)

We define various attribute categories (government, non-government, ethnic/ideological, armed, etc.) and add those as node-level data.

```{r nodal_attributes}
# Define actor categories
gov_actors <- c("Government of Myanmar (Burma)", "BMA", "NUG")
nongov_actors <- c(
  "KNU", "MTA", "DKBA", "MDA", "RCSS", "Buddhists (Myanmar)", "ARSA",
  "Civilians", "KNPP", "KIO", "RSO", "UWSA", "NMSP", "ABSDF", "NSCN-K",
  "PSLF", "God's Army", "MDA - LM", "CNF", "Muslims (Myanmar)", "SSPP",
  "DKBA 5", "MNDAA", "ULA"
)
ethnic_actors <- c(
  "KNU", "RCSS", "KIO", "UWSA", "NMSP", "KNPP", "NSCN-K", "CNF", "PSLF", "SSPP",
  "MTA", "DKBA", "MDA", "RSO", "God's Army", "MDA - LM", "DKBA 5", "MNDAA", "ULA"
)
ideological_actors <- c(
  "Government of Myanmar (Burma)", "NUG", "ARSA", "ABSDF", "Civilians",
  "Buddhists (Myanmar)", "Muslims (Myanmar)", "BMA"
)
armed_actors <- c(
  "KNU", "MTA", "DKBA", "MDA", "RCSS", "ARSA", "NMSP", "KNPP", "KIO", "RSO",
  "UWSA", "ABSDF", "NSCN-K", "PSLF", "God's Army", "DKBA 5", "MNDAA", "ULA",
  "CNF", "SSPP", "NUG", "MDA - LM"
)
non_armed_actors <- c(
  "Government of Myanmar (Burma)", "BMA", "Civilians", 
  "Buddhists (Myanmar)", "Muslims (Myanmar)"
)

# Create a data frame for nodal attributes
actors <- unique(c(myanmar$side_a, myanmar$side_b))
node_data <- data.frame(
  actor = actors,
  gov_or_nongov = ifelse(actors %in% gov_actors, 1, 
                         ifelse(actors %in% nongov_actors, 0, NA)),
  ethnic_or_ideological = ifelse(actors %in% ethnic_actors, 1, 
                                 ifelse(actors %in% ideological_actors, 0, NA)),
  armed_or_not = ifelse(actors %in% armed_actors, 1, 
                        ifelse(actors %in% non_armed_actors, 0, NA))
)

node_data <- node_data |> mutate_all(as.factor)

# Add these attributes to the network
my_network_civ <- add_node_vars(my_network_civ, 
                            node_data = node_data, 
                            actor = 'actor', 
                            node_vars = 'gov_or_nongov')
my_network_civ <- add_node_vars(my_network_civ, 
                            node_data = node_data, 
                            actor = 'actor', 
                            node_vars = 'ethnic_or_ideological')
my_network_civ <- add_node_vars(my_network_civ, 
                            node_data = node_data, 
                            actor = 'actor', 
                            node_vars = 'armed_or_not')

# Plot with different color encoding
plot(
  my_network_civ,
  title = "Myanmar Network",
  point_color_var = "gov_or_nongov", 
  layout = "circle",
  add_text = TRUE
)

plot(
  my_network_civ,
  title = "Myanmar Network",
  point_color_var = "ethnic_or_ideological", 
  layout = "circle",
  add_text = TRUE
)
```

### Bipartite Network (Myanmar Case)

Finally, we illustrate a bipartite network example, focusing on interactions between Buddhist and Christian actors in Myanmar.

```{r bipartite_network}
# Define actors by religion
buddhist_actors <- c("PSLF", "DKBA", "God's Army", "Government of Myanmar (Burma)")
christian_actors <- c("KNU", "KNPP", "KIO", "MDA", "NSCN-K", "CNF")

selected_actors <- c(buddhist_actors, christian_actors)

# Subset data for only these actors
myanmar_bipartite <- subset(myanmar, 
                            side_a %in% selected_actors & side_b %in% selected_actors)
if (nrow(myanmar_bipartite) == 0) {
  stop("No data for the specified Buddhist-Christian interactions.")
}

# Tag each side as Buddhist or Christian
myanmar_bipartite$actor_type_a <- ifelse(
  myanmar_bipartite$side_a %in% buddhist_actors, "Buddhist", "Christian")
myanmar_bipartite$actor_type_b <- ifelse(
  myanmar_bipartite$side_b %in% buddhist_actors, "Buddhist", "Christian")

# Construct bipartite network
bipartite_network <- netify(
  myanmar_bipartite,
  actor1 = 'side_a',
  actor2 = 'side_b',
  mode = 'bipartite',
  weight = 'deaths_civilians',
  symmetric = FALSE,
  sum_dyads = TRUE,
  diag_to_NA = TRUE,
  missing_to_zero = TRUE
)

# Plot bipartite network
plot(
  bipartite_network,
  title = "Bipartite Network: Buddhist vs Christian Actors in Myanmar",
  add_text = TRUE,
  layout = "bipartite"
)
```

## Conclusion

In this vignette, we explored how to install and use **netify** to create different types of conflict networks (unweighted, weighted, longitudinal, bipartite), how to summarize and visualize them, and how to incorporate additional node attributes for deeper analysis.

For more details on **netify**, consult the package documentation and function help files.
