---
title: "Netify Workshop"
author: "Cassy Dorff and Shahryar Minhas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Netify Workshop}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  Package Overview 

This vignette provides a high level overview of our package from start to finish. The best use of this vignette is to introduce the main components of the package to larger audiences. This overview covers our primary functions with data examples and minimal writing. 

`netify` goals: 

  (1) Create: Netify your data! We Make it easy for users to create networks from raw data in $R$ as well as add additional features, such as nodal and dyadic variables, to the network object.
  (2) Explore: Explore characteristics of the network you created, like summary statistics at both the network and actor levels. Visualize your network.
  (3) Advance:  Advance your network analysis to the next stage by preparing it for use in other network packages and modeling approaches. 
  
`netify` provides a suite of primary functions to help achieve these goals:

| Create💡  | Explore 🔎 | Advance ️🚀|
|---------|---------|---------|
| `netify()`  | `peek()`  | `prep_for_amen()`  |
| `add_nodal()` | `summary()`   | `prep_for_igraph()`  |
| `add_dyad()`  | `summary_actor()`  | `prep_for_network()`  |
| `subset_netlet()`  | `plot_actorStats()`  |   | 
|              |   `plot.netify()`|        |

`netify` can take dyadic data or edgelists to get started. 

- The package can also create different types of networks including:

  - cross sectional networks (with static and varying actor composition)
  - longitudinal 
  - bipartite networks
  - multilayer

- As well as create networks with different edge types:
  - weighted 
  - binary 
  - symmetric or non-symmetric 

## Step 1: Create 💡

Begin by loading packages and supplying the data.


```{r, eval=TRUE, results='hide', warning=FALSE, message=FALSE}
# load packages
library(netify)

# install extra packages for this vignette
if('tidyverse' %in% rownames(installed.packages())){
  install.packages('tidyverse', repos='https://cloud.r-project.org') }
if('peacesciencer' %in% rownames(installed.packages())){
  install.packages('peacesciencer', repos='https://cloud.r-project.org') }

# load necessary packages for this vignette
library(peacesciencer)
library(tidyverse)

# organize external data for peacesciencer
peacesciencer::download_extdata()

# create dyadic data set over time using peacesciencer
cow_dyads <- create_dyadyears( 
     subset_years = c(1995:2014)
     ) %>%
    # add mids
    add_cow_mids() %>%
    # add capital distance
    add_capital_distance() %>%
    # add democracy
    add_democracy() %>%
    # add gdp
    add_sdp_gdp()
```

Next, create a `netlet` object from the above COW data frame using our package's core function `netify.` There are a number of useful parameters, but the most important ones to highlight are:

- `dyad_data` is a dyadic data.frame that should have at least the following variables used to specify actors:

  - `actor1`: character indicating actor 1 variable in the data
  - `actor2`: character indicating actor 2 variable in the data
  
- `netify_type` is a type of netlet object (`cross-sec`, `longit_list`, or `longit_array`).

```{r, eval=TRUE}
mid_long_network <- netify(
  cow_dyads,
  actor1='ccode1', actor2='ccode2', time='year',
  weight='cowmidonset', 
  actor_time_uniform=FALSE,
  sum_dyads=FALSE, symmetric=TRUE,
  diag_to_NA=TRUE, missing_to_zero=FALSE,
  nodal_vars = c('v2x_polyarchy1', 'v2x_polyarchy2'),
  dyad_vars = c('capdist'),
  dyad_vars_symmetric = c(TRUE, FALSE)
  )
```

**Congratulations** you have created a network object! 🎉

We can also add nodal and dyadic data after we've created the network via the `add_nodal()` and `add_dyad()` functions. 

Let's assume that we had information about each actor in the network that we'd like to add as a nodal variable after we already made the network object. For example, lets add a variable measuring gdp for each node in the network over time:

```{r, eval=TRUE}
# create a vector of nodal data
node_data <- unique(cow_dyads[,c("ccode1", "year", "wbgdppc2011est2")])

# add nodal variable to netlet object
mid_long_network <- add_nodal(
    netlet = mid_long_network,
    node_data = node_data,
    actor = "ccode1",
    time = "year"
)
```

## Step 2: Explore 🔎

We made a network, so let's look at it. First, we might want to take a `peek` at the network object to see if the matrix looks the way we'd expect it to look. This function lets you glance at a specific slice of the network if it is longitudinal or the entire network if it is cross-sectional.

```{r}
peek(mid_long_network, when_to_peek = c('2012'), what_to_peek = 30)
```

Next, let's examine a few basic summary statistics about the network using our  `summary()` function.

```{r}
# create data.frame that provides network-level summary stats 
# for each year of the network
summary(mid_long_network)
```

We might also want to look at a specific set of actors' statistics overtime. Our built-in function, `summary_actor` will calculate in-degree, out-degree, average degree, and eigenvector centrality for each actor in each time period.

```{r}
# every year & every actor (messy!)
summary_actor_mids <- summary_actor(mid_long_network)
```

```{r, fig.width=6, fig.height=6}

# top 5 GDP countries
top_5 <-c("2", "710", "740", "750", "200")

# zoom in on specific parts of the network for actor viz
mid_subset <- summary_actor_mids %>%
              filter(actor %in% top_5)

plot_actorStats(mid_subset, 
                net_stat = "degree",
                longitudinal = TRUE) +
                facet_wrap(~actor)
```

Instead of looking at summary statistics, we also might want to simply visualize the entire network. We can do this by plotting the netify object. Or we can first subset the object so that we plot only select time periods.

```{r, fig.width=6, fig.height=6}
plot.netify(mid_long_network)
```

## Step 3: Advance 🚀

Once we have created and explored our network object, we might want to continue analyzing the data using different modeling approaches. `netify` makes this simple!

```{r}
# install (if necessary) and load amen
if('amen' %in% rownames(installed.packages())){
  install.packages('amen', repos='https://cloud.r-project.org') }
library(amen)

# install (if necessary) and load ergm
if('ergm' %in% rownames(installed.packages())){
  install.packages('ergm', repos='https://cloud.r-project.org') }
library(ergm)

# lets take a cross-section of the network object

```

